---
title:
  "Beyond the Binary: Uncovering a Hidden Customer Segment in Wholesale Purchasing Data"
subtitle: "SMM068 (2025-26) Coursework 03 - Group 01"
author:
  - "Abdulrahman Alolyan"
  - "Benjamin Evans"
  - "Basmah Khan"
  - "Ardi Wira Sudarmo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
papersize: a4
fontsize: 12pt
output:
  bookdown::html_document2:
    self_contained: true
    css: style.css
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: paper
    highlight: pygments
    df_print: paged
    pandoc_args: ["--lua-filter=custom-boxes.lua"]
  # runtime: shiny
  bookdown::pdf_document2:
    latex_engine: lualatex
    includes:
      in_header: preamble.tex
    pandoc_args: ["--lua-filter=custom-boxes.lua"]
    number_sections: true
    toc: true
    # citation_package: natbib
    keep_tex: true
bibliography: references.bib
csl: ieee.csl
---

```{=latex}
\newpage
```

```{=html}
<h1>Information</h1>

This R Markdown document was created as part of a group assignment for SMM636 at Bayes Business School, City St George's, University of London in Term 2 2025-26.

```

```{r asthetic-header, include=FALSE, echo=FALSE}
# ============================================================================ #
# Key Information ====
# SMM636 Machine Learning
# Group Coursework 2025-26
# Group:        Group 01
# Authors (in alphabetical order):
#   - Abdulrahman Alolyan
#   - Benjamin Evans
#   - Basmah Khan
#   - Ardi Wira Sudarmo
# Professor:    Dr. Rui Zhu
# Institution:  Bayes Business School - City St George's, University of London
# Date:         03 Apr 2026
#
# Description:  Term 2 group project for SMM636 Machine Learning
#               (25% of coursework grade - 25% of module grade). The R code
#               below has been exported directly from an R Markdown (.rmd) file.
#               Hence the knitr settings.
#
# Dependencies:
#   - tidyverse: loads dplyr, tidyr, ggplot2, purrr (& others)
#   - TBI
#
# Acknowledgements
#   - TBI
# ============================================================================ #
```

```{r setup-knitr, include=FALSE, echo=FALSE}
# Setup & configuration ====
#----------------------- Initial setup (knitr settings) -----------------------#
dir.create("fig", showWarnings = FALSE)

# Defaults common to all outputs
knitr::opts_chunk$set(
  echo     = TRUE,
  message  = FALSE,
  warning  = FALSE,
  fig.align = "center",
  out.width = "100%",
  fig.path  = "fig/",
  dpi       = 300
)

# Output-specific settings
if (knitr::is_latex_output()) {
  knitr::opts_chunk$set(
    fig.width = 6,
    fig.height = 4,
    dev = "pdf",
    fig.pos = "!ht",
    out.extra = ""
  )
} else {
  knitr::opts_chunk$set(
    fig.width = 6,
    fig.height = 4,
    dev = "svglite"  # or "png"
  )
}
```

```{r setup-qol, include=FALSE, echo=FALSE}
#----------------------------- Clean environment ------------------------------#
rm(list = ls()) # Remove all objects
graphics.off() # Close all graphical devices
cat("\014") # Clean console
```

```{r load-dependencies, include=FALSE, echo=FALSE}
#------------------- Load dependencies / external libraries -------------------#
# --- Core Data Manipulation ---
library(tidyverse)  # loads dplyr, tidyr, ggplot2, purrr, etc.

# --- Visualization & Plot Arrangement ---
library(patchwork)  # combining multiple ggplot objects
library(gridExtra)  # arranging base/grid plots
library(ggpubr)     # publication-ready plot themes/utilities
library(corrplot)   # correlation heatmaps
library(GGally)     # pairs plots
library(factoextra) # PCA & clustering visualizations

# --- Analytics & Modeling ---
library(NbClust)    # optimal-k diagnostics

# NOTE: The packages below are deliberately NOT loaded into the global 
# environment to prevent function masking conflicts (e.g., MASS::select 
# breaking dplyr::select). Call their functions directly using namespaces:
# cluster::daisy(), MASS::isoMDS()
# library(MASS)     
# library(cluster)

library(mclust)

# --- Reporting & Tables ---
library(kableExtra) # for formatting output tables
```

```{r custom-functions, include=FALSE, echo=FALSE}
#---------------------------- Custom QOL functions ----------------------------#
#####################################
# function: banner comments (used to to section up code)
# Usage: banner_comment("Element 1: data cleaning") -> then ctrl + v (or cmd+v)
#####################################
banner_comment <- function(text, width = 80, border = "#", fill = "-") {
  txt <- paste0(" ", text, " ")
  inner_width <- width - 2 * nchar(border)
  banner_string <- ""

  if (inner_width <= nchar(txt)) {
    banner_string <- paste0(border, txt, border)
  } else {
    pad_total <- inner_width - nchar(txt)
    pad_left <- pad_total %/% 2
    pad_right <- pad_total - pad_left

    banner_string <- paste0(
      border,
      strrep(fill, pad_left),
      txt,
      strrep(fill, pad_right),
      border
    )
  }

  cat(banner_string, "\n")
  # copy banner to allow direct pasting (requires clipr)
  clipr::write_clip(banner_string)
  # avoid [1] when printing if want to manually copy
  invisible(banner_string)
}
#####################################
# function: format p-values for text
# Usage (in-line): `r format_p_vals(ad_test_result$p.value)`
# Usage (console): format_p_vals(ad_test_result$p.value)
#####################################
format_p_vals <- function(p) {
  if (length(p) != 1L || is.na(p)) {
    stop("Error! p must be a single non-missing value")
  }
  if (p > 1) {
    stop("Error! Value greater than 1")
  }
  if (p < 0) {
    stop("Error! Value less than 0")
  }

  if (p >= 0.01) {
    paste0("= ", formatC(p, format = "f", digits = 2))
  } else if (p >= 0.001) {
    paste0("= ", formatC(p, format = "f", digits = 3))
  } else {
    "< 0.001"
  }
}
#####################################
# function: format confidence intervals for tables & text
# Usage (in-line): `r format_interval(el2_ci_normal_95[1], el2_ci_normal_95[2])`
# Usage (console): format_interval(el2_ci_normal_95[1], el2_ci_normal_95[2])
#####################################
format_interval <- function(lower, upper, digits = 3, small_interval = 5L) {
  paste0(
    "[",
    formatC(
      lower,
      format = "f",
      digits = digits,
      small.interval = small_interval,
      small.mark = " "
    ),
    ", ",
    formatC(
      upper,
      format = "f",
      digits = digits,
      small.interval = small_interval,
      small.mark = " "
    ),
    "]"
  )
}
#####################################
# function: format truncated ellipses
# Usage (in-line): `r tbi`
# Usage (console): tbi
#####################################
fmt_trunc_ellip <- function(
  x,
  digits = 7,
  ellip = "...",
  tol = 1e-12,
  trim_zeros_if_exact = TRUE
) {
  out <- rep(NA_character_, length(x))
  ok <- is.finite(x)
  scale <- 10^digits
  xt <- trunc(x[ok] * scale) / scale
  s <- formatC(xt, format = "f", digits = digits)
  add <- abs(x[ok] - xt) > tol * pmax(1, abs(x[ok]))
  if (trim_zeros_if_exact) {
    s_trim <- sub("0+$", "", s) # drop trailing zeros
    s_trim <- sub("\\.$", "", s_trim) # drop trailing decimal point
  } else {
    s_trim <- s
  }
  out[ok] <- ifelse(add, paste0(s, ellip), s_trim)
  out[is.na(x)] <- NA_character_
  out
}
#####################################
# function: function to format 6dp numbers with a space after the 3rd decimal
# digit
# Usage (in-line): TBI
# Usage (console): TBI
#####################################
format_spaced_6dp <- function(x) {
  # Format to 6 decimal places
  s <- sprintf("%.6f", x)
  # Insert space between the first 3 and last 3 decimal digits
  sub("(\\.[0-9]{3})([0-9]{3})$", "\\1 \\2", s)
}
#####################################
# function: format variable names in green in latex (& normal code in html)
# Usage (in-line): TBI `r format_var_name("dvisit")`
#####################################
format_var_name <- function(x) {
  if (knitr::is_latex_output()) {
    # replace all "_" with "\_" in va name
    paste0("\\greentt{", gsub("_", "\\\\_", x), "}")
  } else {
    # change to paste0("<code style='color:green'>", x, "</code>") for green
    # paste0("`", x, "`")
    paste0("<code style='color:green'>", x, "</code>")
  }
}
```

```{r word-count, include=FALSE}
rmd_lines <- readLines("01SMM363-assessment-03.rmd", warn = FALSE)

# ---- Remove YAML front matter ----
yaml_bounds <- which(rmd_lines == "---")
if (length(yaml_bounds) >= 2) {
  rmd_lines <- rmd_lines[-(yaml_bounds[1]:yaml_bounds[2])]
}

# ---- Remove code chunks ----
in_chunk <- FALSE
keep <- rep(TRUE, length(rmd_lines))
for (i in seq_along(rmd_lines)) {
  if (grepl("^```\\{", rmd_lines[i])) {
    in_chunk <- TRUE
    keep[i] <- FALSE
    next
  }
  if (in_chunk && grepl("^```\\s*$", rmd_lines[i])) {
    in_chunk <- FALSE
    keep[i] <- FALSE
    next
  }
  if (in_chunk) keep[i] <- FALSE
}
rmd_lines <- rmd_lines[keep]

# ---- Identify which lines fall inside counted sections ----
# Sections to include (level-1 headers only)
counted_ids <- c("intro", "qone", "qtwo", "qthr")
in_counted_section <- FALSE
include <- rep(FALSE, length(rmd_lines))

for (i in seq_along(rmd_lines)) {
  line <- rmd_lines[i]
  # Detect any level-1 header (# but not ##)
  if (grepl("^#\\s+[^#]", line)) {
    # Check if this header belongs to a counted section
    in_counted_section <- any(sapply(counted_ids, function(id) {
      # [^}]* allows any characters (like spaces, hyphens, classes) before the 
      # closing }
      # this is required to count the words in the unnumbered introduction
      grepl(paste0("\\{\\s*#", id, "[^}]*\\}"), line)
    }))
    next  # skip the header line itself
  }
  if (in_counted_section) include[i] <- TRUE
}

counted_lines <- rmd_lines[include]

# ---- Remove sub-headers (##, ###, etc.) ----
counted_lines <- counted_lines[!grepl("^#{2,}", counted_lines)]

# ---- Remove table lines (markdown pipes) ----
counted_lines <- counted_lines[!grepl("^\\|", counted_lines)]

# ---- Remove figure captions: ![caption](...) ----
counted_lines <- counted_lines[!grepl("^!\\[", counted_lines)]

# ---- Remove fenced div markers (::: name) ----
counted_lines <- counted_lines[!grepl("^:::", counted_lines)]

# ---- Remove LaTeX commands ----
counted_lines <- counted_lines[!grepl("^\\\\", counted_lines)]

# ---- Remove HTML comments ----
counted_lines <- counted_lines[!grepl("^<!--", counted_lines)]

# ---- Remove inline R references like \@ref() ----
counted_lines <- gsub("\\\\@ref\\([^)]*\\)", "", counted_lines)

# ---- Count ----
word_count <- sum(stringi::stri_count_words(counted_lines), na.rm = TRUE)

message("Word count (sections qone, qtwo, qthr, body text only): ", word_count)

# Write for LaTeX preamble
writeLines(
  paste0("\\newcommand{\\wordcount}{",
         formatC(word_count, big.mark = ","), "}"),
  "wordcount.tex"
)
```

# Introduction {#intro -}

This report analyses annual spending data from 440 clients of a Portuguese
wholesale distributor, comprising 298 Hotel/Restaurant/Café (HoReCa) and 142
Retail customers across six product categories: `r format_var_name("Fresh")`,
`r format_var_name("Milk")`, `r format_var_name("Grocery")`,
`r format_var_name("Frozen")`, `r format_var_name("Detergents_Paper")`, and
`r format_var_name("Delicassen")` [@Cardoso2013; @Abreu2011]. Data-driven
customer segmentation could inform inventory planning, pricing strategy, and
sales team deployment, yet the existing classification is a simple binary label.
This analysis uses Principal Component Analysis (PCA) and clustering to
investigate whether the existing two-category classification adequately captures
the structure of customer spending, or whether the data supports a more nuanced
segmentation.

```{r load-data, echo=FALSE, include=FALSE}
set.seed(1234)

data <- read.csv("data/Wholesale_customers_data.csv")
str(data)
summary(data)

# just get the $ amounts spent
spending <- data[, c(
  "Fresh",
  "Milk",
  "Grocery",
  "Frozen",
  "Detergents_Paper",
  "Delicassen"
)]

spending_log <- log1p(spending)

rescale01 <- function(x) (x - min(x)) / (max(x) - min(x))

raw_df <- spending %>%
  pivot_longer(everything(), names_to = "Category", values_to = "Spend") %>%
  mutate(Transform = "Raw")

log_df <- spending_log %>%
  pivot_longer(everything(), names_to = "Category", values_to = "Spend") %>%
  mutate(Transform = "Log-transformed")

# add data to import into in-line text
# standardise log-transformed data
scaled_data <- scale(spending_log)

# PCA
pca_result <- prcomp(spending_log, center = TRUE, scale. = TRUE)

# Skewness (raw)
skew_vals <- apply(spending, 2, e1071::skewness)
min_skew  <- min(skew_vals)
```

```{r data-table, echo=knitr::is_html_output(), eval=knitr::is_html_output()}
DT::datatable(
  data,
  filter   = "top",
  options  = list(pageLength = 15, scrollX = TRUE),
  caption  = "Wholesale customers"
)
```

# Principle component analysis (PCA) {#qone}

<!-- _Apply PCA to the dataset and summarise and interpret key your findings_ -->

A log-transformation was applied to normalise the distributions (Appendix,
Figures \@ref(fig:histograms-overlay) & \@ref(fig:corrplots-side-by-side)) and
address the right-skewness observed across all six spending variables 
(skewness > `r formatC(min_skew, format = "f", digits = 1)`
for all). The variables were then standardised to ensure equal contribution to
the PCA, preventing categories with naturally larger absolute expenditures from
dominating the model.

As shown in Figure \@ref(fig:pca-grid)(a) the first two
principal components capture the majority of the underlying data structure,
together explaining 71.3% of the total variance (PC1: 44.1%; PC2: 27.2%).

The component loadings in Figure \@ref(fig:pca-grid)(b) reveal distinct commercial
purchasing patterns. PC1 acts as the primary "channel axis," driven by strong
positive loadings for `r format_var_name("Grocery")` (0.571),
`r format_var_name("Detergents_Paper")` (0.551), and `r format_var_name("Milk")`
(0.542), contrasted with near-zero or negative loadings for 
`r format_var_name("Fresh")` and `r format_var_name("Frozen")`. PC2 serves as a
"perishable volume axis," characterised by high positive loadings for 
`r format_var_name("Fresh")` (0.590), `r format_var_name("Frozen")` (0.590), and 
`r format_var_name("Delicassen")` (0.530), capturing the scale of fresh food
purchasing independent of business type. The biplot (Figure \@ref(fig:pca-grid)(d))
confirms this visually, showing clear separation between HoReCa and Retail
customers along PC1. PCA is restricted to continuous inputs, necessitating the
exclusion of `r format_var_name("Channel")` and `r format_var_name("Region")`. 
A Gower-distance MDS analysis incorporating these variables (Appendix,
\@ref(apd:suppl-mds)) confirms that the categorical labels dominate the distance
metric, validating the decision to analyse spending variables independently.


```{r pca-fit, echo=FALSE, include=FALSE}
# ==============================================================================
# Q1 PCA
# ==============================================================================

pca_result <- prcomp(spending_log, center = TRUE, scale. = TRUE)
summary(pca_result)

cumvar <- cumsum(pca_result$sdev^2 / sum(pca_result$sdev^2))
cat("Cumulative variance explained:\n")
print(round(cumvar, 3))

loadings <- pca_result$rotation[, 1:3]
print(round(loadings, 3))
```

```{r pca-scree, echo=FALSE, include=FALSE, fig.width=7, fig.height=5}
fviz_eig(
  pca_result,
  addlabels = TRUE,
  ylim = c(0, 60),
  main = "Scree Plot - Variance Explained by Each PC"
)
```

```{r pca-biplot, echo=FALSE, include=FALSE, fig.width=8, fig.height=6}
fviz_pca_biplot(
  pca_result,
  label   = "var",
  col.var = "firebrick",
  col.ind = alpha("steelblue", 0.3),
  repel   = TRUE,
  title   = "PCA Biplot (log-transformed, standardised data)"
)
```

```{r pca-contrib-pc, echo=FALSE, include=FALSE, fig.width=7, fig.height=4}
# Opted to hide these for now 
# (don't think we have space and they don't add too much)
# Even including in appendix is limited - easier to include table
fviz_contrib(
  pca_result,
  choice = "var",
  axes   = 1,
  title  = "Variable Contributions to PC1"
)

fviz_contrib(
  pca_result,
  choice = "var",
  axes   = 2,
  title  = "Variable Contributions to PC2"
)
```

```{r pca-loadings-table, echo=FALSE, include=FALSE}
loadings_df <- as.data.frame(round(pca_result$rotation[, 1:3], 3))
colnames(loadings_df) <- c("PC1", "PC2", "PC3")

kable(
  loadings_df,
  caption  = "PCA loadings (first three components). Values over 0.4 are highlighted in bold.",
  linesep  = "",
  booktabs = TRUE
) %>%
  kable_styling(
    full_width = FALSE,
    position   = "center"
  ) %>%
  column_spec(2, bold = abs(loadings_df$PC1) >= 0.4) %>%
  column_spec(3, bold = abs(loadings_df$PC2) >= 0.4) %>%
  column_spec(4, bold = abs(loadings_df$PC3) >= 0.4)
```

```{r pca-channel, echo=FALSE, include=FALSE, fig.width=8, fig.height=6}
fviz_pca_ind(
  pca_result,
  geom.ind    = "point",
  col.ind     = factor(data$Channel),
  palette     = c("#E64B35", "#4DBBD5"),
  addEllipses = TRUE,
  legend.title = "Channel",
  title       = "PCA - Individuals Coloured by Channel (1=HoReCa, 2=Retail)"
)
```

```{r pca-grid, echo=FALSE, include=TRUE, fig.width=12, fig.height=10, fig.cap="Principal Component Analysis of log-transformed, standardised spending data. (a) Scree plot showing the percentage of variance explained by each component. (b) Loadings of the original variables on the first four principal components. Values exceeding 0.4 in magnitude indicate strong contributions. (c) Biplot overlaying variable loadings (red arrows) on individual observations. (d) Individuals projected onto PC1-PC2, coloured by distribution channel (1 = HoReCa, 2 = Retail) with 95% concentration ellipses."}

# --- Top left: Scree plot ---
p1 <- fviz_eig(
  pca_result,
  addlabels = TRUE,
  ylim = c(0, 60),
  main = "(a) Scree Plot - Variance Explained by Each PC"
)

# --- Top right: Loadings table as a grob ---
loadings_df <- as.data.frame(round(pca_result$rotation[, 1:3], 3))
colnames(loadings_df) <- c("PC1", "PC2", "PC3")

# --- Top right: Loadings table ---
loadings_df <- as.data.frame(pca_result$rotation[, 1:4])
colnames(loadings_df) <- c("PC1", "PC2", "PC3", "PC4")
# loadings_df <- as.data.frame(pca_result$rotation[, 1:3])
# colnames(loadings_df) <- c("PC1", "PC2", "PC3")
loadings_df[] <- lapply(loadings_df, function(x) format(round(x, 3), nsmall = 3))

p2 <- ggtexttable(
  loadings_df,
  theme = ttheme(
    "blank",
    tbody.style = tbody_style(
      size  = 13,
      hjust = 1.00, #controls alignment {0.00, 0.5, 1.00} 
      x     = 0.95, #controls alignment {0.05, 0.5, 0.95} 
      fill  = "white"
    ),
    colnames.style = colnames_style(
      size  = 13,
      face  = "bold",
      hjust = 1.00,
      x     = 0.95,
      fill  = "white"
    ),
    rownames.style = rownames_style(
      size  = 13,
      face  = "plain",
      hjust = 0,
      x     = 0.05,
      fill  = "white"
    )
  )
) %>%
  tab_add_hline(at.row = 1:2, row.side = "top", linewidth = 2) %>%
  tab_add_hline(at.row = nrow(loadings_df) + 1, row.side = "bottom", linewidth = 2) +
  labs(title = "(b) PCA Loadings") +
  theme(
    plot.title      = element_text(hjust = 0.5, size = 13),
    plot.background = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "white", colour = NA)
  )

# --- Bottom left: Biplot ---
p3 <- fviz_pca_biplot(
  pca_result,
  label   = "var",
  col.var = "firebrick",
  col.ind = alpha("steelblue", 0.3),
  repel   = TRUE,
  title   = "(c) PCA Biplot (log-transformed, standardised data)"
)

# --- Bottom right: Channel plot ---
p4 <- fviz_pca_ind(
  pca_result,
  geom.ind     = "point",
  col.ind      = factor(data$Channel),
  palette      = c("#E64B35", "#4DBBD5"),
  addEllipses  = TRUE,
  legend.title = "Channel",
  title        = "(d) PCA - Individuals Coloured by Channel (1=HoReCa, 2=Retail)"
)

# BE note: qol to get bottom two plots to align
# ggplot_build(p3)$layout$panel_params[[1]]$y.range
# ggplot_build(p4)$layout$panel_params[[1]]$y.range

p3 <- p3 + coord_cartesian(ylim = c(-5.58, 5.24))
p4 <- p4 + coord_cartesian(ylim = c(-5.58, 5.24))

# --- Combine 2x2 ---
(p1 | p2) / (p3 | p4)
```

TBI

```{=latex}
\clearpage
\newpage
```

# Customer clustering {#qtwo}

_Perform clustering on the customers_

```{r Q2, echo=FALSE, fig.pos="!ht"}
# ==============================================================================
# Q2 Clustering
# ==============================================================================
## cluster log-transformed, standardised data (same as PCA input)
scaled_data <- scale(spending_log)
 
```

TBI

## KMeans

```{r}

KMeans.results <- list()

for (k in 2:5) {
  km.out = kmeans(scaled_data, centers= k, nstart = 20)  
  KMeans.results[[k]] <- km.out
}

```


```{r}

for (k in 2:5) {
  km.out <- KMeans.results[[k]]
  print(fviz_cluster(km.out, scaled_data,
                     geom = c('point'),
                     ellipse.type = "norm",
                     main = sprintf("KMeans (k=%s)", k) )
  )
}

```

```{r}

fviz_nbclust(scaled_data, kmeans, method='silhouette')

```

```{r}

for (k in 2:5) {
  km.out <- KMeans.results[[k]]
  print(
    fviz_silhouette(cluster::silhouette(km.out$cluster, dist(scaled_data)^2))
  )
}

```

### Split up by Channel

```{r}

spending.cols = c('Fresh', 'Milk', 'Grocery', 'Frozen', 'Detergents_Paper', 'Delicassen')

scaled_df <- cbind(data[c('Channel', 'Region')], scaled_data)

ch1_scaled_df <- scaled_df |> filter(Channel == 1)

ch2_scaled_df <- scaled_df |> filter(Channel == 2)

```



```{r}


for (k in 2:5) {
  km.out = kmeans(ch1_scaled_df[spending.cols], centers= k, nstart = 20)  
  
  print(fviz_cluster(km.out,
                     ch1_scaled_df[spending.cols],
                     geom = c('point'),
                     ellipse.type = "norm",
                     main = sprintf("Channel 1KMeans (k=%s)", k) )
  )
}

```


## MDS scaled

```{r}

# Gower distance on all 8 variables scaled (Channel & Region as factors)
scaled_df$Channel <- factor(data$Channel)
scaled_df$Region  <- factor(data$Region)

gow_scaled <- cluster::daisy(scaled_df, metric = "gower")
 
# Non-metric MDS
mds_scaled <- MASS::isoMDS(gow_scaled, k = 2)

```


```{r}

mds_scaled_out <- data.frame(
  MDS1 = mds_scaled$points[, 1],
  MDS2 = mds_scaled$points[, 2],
  Region = scaled_df$Region,
  Channel = scaled_df$Channel
)

ggplot(mds_scaled_out) +
  geom_point(aes(x=MDS1, y=MDS2, colour=Region, shape=Channel)) +
  labs(title = 'MDS using Gower distance of Region, Channel and scaled 6 spending variables')

```

```{r}

for (k in 4:8) {
  km.out = kmeans(mds_scaled$points, centers= k, nstart = 20)  
  
  print(fviz_cluster(km.out,
                     mds_scaled$points,
                     geom = c('point'),
                     ellipse.type = "norm",
                     main = sprintf("KMeans (k=%s) on MDS output of scaled data", k),
                     xlab='' , ylab='yy')
  )
}

```


## GMM

```{r q2-gmm}

gmm_mds_scaled <- Mclust(data=mds_scaled_out[c('MDS1', 'MDS2')])

summary(gmm_mds_scaled, parameters=TRUE)

gmm_mds_scaled$BIC[1]
gmm_mds_scaled$modelName
gmm_mds_scaled$G

# BIC table
bic_table <- gmm_mds_scaled$BIC
# top 3
summary(gmm_mds_scaled$BIC)


```
```{r q2-gmm-mds-2}
gmm_clusters <- gmm_mds_scaled$classification

crosstab <- table(
  GMM     = gmm_clusters,
  Channel = data$Channel,
  Region  = data$Region
)

# flatten
ftable(crosstab)
# BE note:
# GMM on MDS performs as predicted:
# 6 of the 8 clusters map perfectly to the 6 Channel x Region groups.
# The "extra" 2 clusters seem to be made from taking the two biggest groups 
# (HoReCa/Other and Retail/Other) and chopping them in half because BIC rewards 
# the density fit
# (think cluster 5 is just 16 awkward boundary points caught in the middle).

```

```{r}
plot(gmm_mds_scaled, what="BIC")
plot(gmm_mds_scaled, what="classification")
plot(gmm_mds_scaled, what="uncertainty")
plot(gmm_mds_scaled, what="density")
```

Q: did GMM find 8 components because BIC rewarded the fit improvement from splitting two of the denser islands, not because it discovered genuinely novel spending structure. The uncertainty plot supports this: low uncertainty everywhere except at the boundaries where islands were artificially bisected, which means those splits are the least confident assignments. If they were real sub-segments, you'd expect the uncertainty to be low within each sub-group and high only at the edges between genuinely different populations.



```{r}
# Compare GMM-on-MDS clusters to K-Means-on-spending clusters
table(
  GMM      = gmm_clusters,
  KMeans_3 = KMeans.results[[3]]$cluster
)

# The GMM-on-MDS vs K-Means cross-tab shows that GMM cluster 2 (which is 145
# HoReCa/Other customers) gets split across K-Means clusters 1, 2, and 3 
# as (25, 116, 4). 
# This is not both methods capturing the same structure differently.  
# It's GMM lumped together customers that K-Means separated, because GMM was
# grouping by demographic label while K-Means was grouping by spending. 
# 
# The lack of clean one-to-one mapping between the two sets of clusters shows
# they're driven by different features. If they captured the same structure,
# we'd see block-diagonal patterns with each GMM cluster mapping mostly to
# one K-Means cluster. 
# Instead we get diffuse rows, meaning a single GMM cluster contains customers 
# from multiple spending-based segments.
```

```{r}
# GMM directly on the same scaled spending data as K-Means
gmm_spending <- Mclust(data = scaled_data)
summary(gmm_spending)

# Compare like-for-like
table(
  GMM_spending = gmm_spending$classification,
  KMeans_3     = KMeans.results[[3]]$cluster
)
#             KMeans_3
# GMM_spending   1   2   3
#            1   4   0  47  ← almost entirely KM3
#            2  91   1   9  ← almost entirely KM1
#            3  33 122   0  ← mostly KM2, some KM1
#            4  11  62   8  ← mostly KM2, some KM1
#            5   8  27  17  ← genuinely mixed


# validate against Channel
table(
  GMM_spending = gmm_spending$classification,
  Channel      = data$Channel
)
#             Channel
# GMM_spending   1   2
#            1   7  44  ← Retail
#            2  18  83  ← Retail (but 18 HoReCa leaking in)
#            3 149   6  ← HoReCa
#            4  77   4  ← HoReCa
#            5  47   5  ← HoReCa

# GMM has split both channels into sub-groups:
# Retail: clusters 1 and 2 (small vs large retailers?)
# HoReCa: clusters 3, 4, and 5 (three tiers of hospitality spending?)
# 
# This is more nuanced than K-Means k=3, which finds one Retail group and splits
# HoReCa in two (or vice versa).
# GMM's flexibility lets it find finer-grained structure
# But question is whether those extra splits are commercially meaningful or just
# overfitting density variation.

table(
  GMM_spending = gmm_spending$classification,
  Channel      = data$Channel,
  Region       = data$Region
) |> ftable()
#                     Region   1   2   3
# GMM_spending Channel                   
# 1            1                1   0   6   ← 7 HoReCa (scattered)
#              2                7   8  29   ← 44 Retail
# 2            1                5   0  13   ← 18 HoReCa
#              2                9  10  64   ← 83 Retail
# 3            1               34  15 100   ← 149 HoReCa
#              2                1   1   4   ← 6 Retail (leakage)
# 4            1               12  10  55   ← 77 HoReCa
#              2                1   0   3   ← 4 Retail (leakage)
# 5            1                7   3  37   ← 47 HoReCa
#              2                0   0   5   ← 5 Retail (leakage)
#
# Region is scattered within each GMM cluster.
# Cluster 3 (HoReCa) has 34 Lisbon, 15 Oporto, 100 Other - roughly
# proportional to the overall Region distribution (77/47/316).
# The sub-segments are spending-driven, not geographic.
# 
# GMM spending profiles (medians):
#   Cluster 1 (n=51):  Retail - low across board, very low Fresh/Frozen
#   Cluster 2 (n=101): Retail - higher Milk/Grocery/Detergents (big retailers?)
#   Cluster 3 (n=155): HoReCa - balanced, moderate Fresh, low Detergents
#   Cluster 4 (n=81):  HoReCa - lower spend tier
#   Cluster 5 (n=52):  HoReCa - lowest tier, minimal across all categories

# Both GMM and K-means on spending data operate on the same input, blind to labels. 
# K-Means says 2–3 segments; GMM says 5.
# Overlap on the major segments validates both. 
# Disagreement on finer is interesting -> discussion 
# Q: does the data genuinely support 5 actionable segments, or is K-Means'
# simpler solution more practical for a distributor who needs to operationalise
# the segmentation?
```

```{r q2-median-spend-gmm}
# Compute median spending per category for each GMM cluster
data$gmm_cluster <- gmm_spending$classification
data %>%
  group_by(gmm_cluster) %>%
  summarise(
    n = n(),
    across(all_of(spending.cols), median),
    .groups = "drop"
  )

# GMM cluster profiles (median raw spending):
#   GMM1 (n=51):  Grocery 13,567 | Det 6,830 | Milk 7,845 | Fresh 1,454
#                 → Heavy Retail: grocery/household dominant, minimal Fresh
#   GMM2 (n=101): Grocery 11,238 | Fresh 7,842 | Milk 7,243 | Det 4,621
#                 → Diversified Retail: buys across all categories
#   GMM3 (n=155): Fresh 14,276 | Frozen 3,383 | Grocery 2,405 | Det 325
#                 → High-volume HoReCa: Fresh-dominated, minimal household
#   GMM4 (n=81):  Fresh 5,969 | Milk 3,218 | Grocery 3,133 | Frozen 907
#                 → Small HoReCa: same category mix as GMM3 but ~40% of spend
#   GMM5 (n=52):  Grocery 3,712 | Fresh 2,270 | Milk 1,572 | Frozen 893
#                 → Lowest tier: minimal spend across all categories
#
# Clusters 3 and 4 (both HoReCa) differ in BOTH scale AND mix:
#   - Scale: GMM3 median Fresh is 14,276 vs GMM4's 5,969 (2.4x)
#   - Mix:   GMM3 is 83% perishable (Fresh+Frozen+Deli = 18,592/21,175)
#            GMM4 is more balanced (Fresh+Frozen+Deli = 7,532/14,903 = 51%)
#   This is a genuine behavioural split, not just big-vs-small.
#   GMM3 are likely restaurants buying fresh produce daily;
#   GMM4 are likely hotels/cafés with more diversified procurement.
#
# Retail split (GMM1 vs GMM2) is similarly informative:
#   - GMM1: concentrated on Grocery+Det+Milk (supermarket profile)
#   - GMM2: spreads spend more evenly including Fresh (convenience store?)
```

```{r q2-k-means-table-1, echo=FALSE, include=FALSE}
# ---- K-Means cluster profiles (k=2 and k=3) ----
km2 <- KMeans.results[[2]]
km3 <- KMeans.results[[3]]

data$km2_cluster <- km2$cluster
data$km3_cluster <- km3$cluster

# Median spending per cluster
km3_profiles <- data %>%
  group_by(km3_cluster) %>%
  summarise(
    n = n(),
    across(all_of(spending.cols), median),
    .groups = "drop"
  )
print(km3_profiles)

# Channel cross-tab
table(KMeans_3 = km3$cluster, Channel = data$Channel)

#            Channel
# KMeans_3    1    2
#        1  ??   ??   ← expect one cluster ≈ Retail
#        2  ??   ??   ← expect one cluster ≈ HoReCa
#        3  ??   ??   ← the "third segment" — check which channel it splits

# ---- Percentage views for both methods ----

# K-Means k=3: % of each cluster's total spend per category
km3_pct <- data %>%
  group_by(km3_cluster) %>%
  summarise(across(all_of(spending.cols), median), .groups = "drop") %>%
  mutate(
    total = Fresh + Milk + Grocery + Frozen + Detergents_Paper + Delicassen,
    across(all_of(spending.cols), ~ round(. / total * 100, 1))
  ) %>%
  select(-total)
print(km3_pct)

# GMM k=5: same treatment
gmm_pct <- data %>%
  group_by(gmm_cluster) %>%
  summarise(across(all_of(spending.cols), median), .groups = "drop") %>%
  mutate(
    total = Fresh + Milk + Grocery + Frozen + Detergents_Paper + Delicassen,
    across(all_of(spending.cols), ~ round(. / total * 100, 1))
  ) %>%
  select(-total)
print(gmm_pct)

# K-Means (k=3) spending profiles (% of median basket)
#   KM1 (n=147): Fresh 33%, Grocery 27%, Milk 20%  ← mixed basket, 56H/91R
#   KM2 (n=212): Fresh 58%, Frozen 13%             ← HoReCa dominant (209H/3R)
#   KM3 (n=81):  Grocery 45%, Milk 27%, Det 18%    ← Retail dominant (33H/48R)
#
# KM2 is a clean HoReCa cluster. KM3 is a clean Retail cluster.
# KM1 is the interesting one — the "middle ground" that mixes both channels.
# This is arguably the k=3 third segment: customers whose spending doesn't
# fit neatly into either profile.
#
# GMM (k=5) spending profiles (% of median basket)
#   GMM1 (n=51):  Grocery 44%, Det 22%, Milk 26%   ← heavy Retail (supermarkets)
#   GMM2 (n=101): Grocery 33%, Fresh 23%, Milk 21% ← moderate Retail (diversified)
#   GMM3 (n=155): Fresh 62%, Frozen 15%            ← high-volume HoReCa
#   GMM4 (n=81):  Fresh 41%, Milk 22%, Grocery 22% ← balanced HoReCa
#   GMM5 (n=52):  Grocery 42%, Fresh 26%           ← small/mixed (lowest spend)
#
# GMM splits Retail into concentrated (GMM1) vs diversified (GMM2)
# GMM splits HoReCa into Fresh-dominated (GMM3) vs balanced (GMM4)
# GMM5 is the low-spend boundary group — similar to KM1's "middle ground"
```

```{r q2-tables-2, echo=FALSE}
# ---- Combined table for report ----
km3_for_table <- km3_pct %>%
  mutate(Method = "K-Means", Cluster = paste0("KM-", km3_cluster)) %>%
  select(-km3_cluster)

gmm_for_table <- gmm_pct %>%
  mutate(Method = "GMM", Cluster = paste0("GMM-", gmm_cluster)) %>%
  select(-gmm_cluster)

combined <- bind_rows(km3_for_table, gmm_for_table) %>%
  select(Cluster, everything(), -Method)

kable(
  combined,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  col.names = c(
    "Cluster",
    "Fresh",
    "Milk",
    "Grocery",
    "Frozen",
    "Detergents\\_Paper",
    "Delicassen"
  ),
  align = c("l", rep("r", 6)),
  caption = "Cluster spending profiles (\\% of median basket)",
  linesep = ""
) %>%
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 10
  ) %>%
  pack_rows("K-Means (k = 3)", 1, 3, bold = TRUE, hline_before = FALSE) %>%
  pack_rows(
    "Gaussian Mixture Model (k = 5)",
    4,
    8,
    bold = TRUE,
    hline_before = TRUE
  )
```
```{r q2-tables-3, echo=FALSE}
# ---- Raw median spending table ----
km3_raw_table <- data %>%
  group_by(km3_cluster) %>%
  summarise(
    n = n(),
    across(all_of(spending.cols), median),
    .groups = "drop"
  ) %>%
  mutate(
    Method = "K-Means",
    Cluster = paste0("KM-", km3_cluster)
  ) %>%
  select(-km3_cluster)

gmm_raw_table <- data %>%
  group_by(gmm_cluster) %>%
  summarise(
    n = n(),
    across(all_of(spending.cols), median),
    .groups = "drop"
  ) %>%
  mutate(
    Method = "GMM",
    Cluster = paste0("GMM-", gmm_cluster)
  ) %>%
  select(-gmm_cluster)

combined_raw <- bind_rows(km3_raw_table, gmm_raw_table) %>%
  select(Cluster, everything(), -Method) %>%
  mutate(across(all_of(spending.cols), ~ format(round(.), big.mark = ",")))

kable(
  combined_raw,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  col.names = c(
    "Cluster",
    "$n$",
    "Fresh",
    "Milk",
    "Grocery",
    "Frozen",
    "Detergents\\_Paper",
    "Delicassen"
  ),
  align = c("l", "r", rep("r", 6)),
  caption = "Cluster spending profiles (median annual spending, m.u.)",
  linesep = ""
) %>%
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 10
  ) %>%
  pack_rows("K-Means (k=3)", 1, 3, bold = TRUE, hline_before = FALSE) %>%
  pack_rows(
    "Gaussian Mixture Model (k = 5)",
    4,
    8,
    bold = TRUE,
    hline_before = TRUE
  )
```

```{r q2-tables-4, echo=FALSE}
# each cluster's share of total category spend
# i.e., what % goes to each segment we have now identified

km3_totals <- data %>%
  group_by(km3_cluster) %>%
  summarise(
    n = n(),
    across(all_of(spending.cols), sum),
    .groups = "drop"
  )

km3_col_pct <- km3_totals %>%
  mutate(
    across(
      all_of(spending.cols),
      ~ round(. / sum(.) * 100, 1)
    ),
    Cluster = paste0("KM-", km3_cluster)
  ) %>%
  select(Cluster, n, all_of(spending.cols))

gmm_totals <- data %>%
  group_by(gmm_cluster) %>%
  summarise(
    n = n(),
    across(all_of(spending.cols), sum),
    .groups = "drop"
  )

gmm_col_pct <- gmm_totals %>%
  mutate(
    across(
      all_of(spending.cols),
      ~ round(. / sum(.) * 100, 1)
    ),
    Cluster = paste0("GMM-", gmm_cluster)
  ) %>%
  select(Cluster, n, all_of(spending.cols))

combined_col_pct <- bind_rows(km3_col_pct, gmm_col_pct)

kable(
  combined_col_pct,
  format    = "latex",
  booktabs  = TRUE,
  escape    = FALSE,
  col.names = c("Cluster", "$n$", "Fresh", "Milk", "Grocery",
                "Frozen", "Detergents\\_Paper", "Delicassen"),
  align     = c("l", "r", rep("r", 6)),
  caption   = "Revenue concentration: each cluster's share (\\%) of total
               category spend across all 440 customers.",
  linesep   = ""
) %>%
  kable_styling(
    latex_options = c("hold_position"),
    font_size     = 10
  ) %>%
  pack_rows("K-Means (k = 3)", 1, 3, bold = TRUE, hline_before = FALSE) %>%
  pack_rows("Gaussian Mixture Model (k = 5)", 4, 8, bold = TRUE, hline_before = TRUE)

```

```{r q2-plots-draft, echo=FALSE, include=TRUE, fig.width=10, fig.height=10, fig.cap = "Cluster assignments from K-Means (k=3, top) and GMM (k=5, bottom) projected onto the first two principal components, which together capture 71.3 percent of total variance. All clustering was performed in the full six-dimensional standardised log-spending space; these plots compress that structure into two dimensions for visualisation. Apparent overlap between clusters  may therefore reflect the loss of separating information in the remaining 28.7 percent of variance captured by PC3--PC6, rather than  genuine ambiguity in cluster assignment."}
# Clusters are computed in the full six-dimensional standardised space and
# projected onto the first two principal components for visualisation, which
# capture 71.3% of total variance.

# PCA scores (already computed)
pca_scores <- as.data.frame(pca_result$x[, 1:2])
colnames(pca_scores) <- c("PC1", "PC2")

# Axis labels with variance explained
pc1_var <- summary(pca_result)$importance[2, 1] * 100
pc2_var <- summary(pca_result)$importance[2, 2] * 100
pc1_lab <- sprintf("PC1 (%.1f%%)", pc1_var)
pc2_lab <- sprintf("PC2 (%.1f%%)", pc2_var)

# ---- Top panel: K-Means k=3 ----
pca_scores$KM3 <- factor(km3$cluster)

p_km <- ggplot(pca_scores, aes(x = PC1, y = PC2, colour = KM3)) +
  geom_point(alpha = 0.5, size = 1.5) +
  # stat_ellipse(level = 0.68, linewidth = 0.8) +
  scale_colour_brewer(palette = "Set1") +
  labs(
    title  = "(a) K-Means (k = 3)",
    x      = pc1_lab,
    y      = pc2_lab,
    colour = "Cluster"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# ---- Bottom panel: GMM k=5 ----
pca_scores$GMM5 <- factor(gmm_spending$classification)

p_gmm <- ggplot(pca_scores, aes(x = PC1, y = PC2, colour = GMM5)) +
  geom_point(alpha = 0.5, size = 1.5) +
  # stat_ellipse(level = 0.68, linewidth = 0.8) +
  scale_colour_brewer(palette = "Set1") +
  labs(
    title  = "(b) GMM (k = 5)",
    x      = pc1_lab,
    y      = pc2_lab,
    colour = "Cluster"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# ---- Stack ----
p_km / p_gmm
```

```{=latex}
\newpage
```

The strength of the spending-only clustering approach is precisely that it
operates without access to the Channel and Region labels. The near-perfect
<!-- recovery of the Channel distinction (Table @ref(tab:km2-channel-crosstab)) -->
validates the segmentation as behaviourally grounded, while the emergence of a
third cluster reveals sub-structure that the existing classification does not
capture. By contrast, incorporating categorical variables via Gower's distance
produces clusters that merely reproduce known labels (Appendix
<!-- @ref(apd:suppl-mds)), confirming that the unsupervised, spending-only approach -->
is both more rigorous and more commercially informative.

```{=latex}
\newpage
```

# Repeating \@ref(qone) & \@ref(qtwo) using an LLM {#qthr}


- Using same prompts for 2 LLMs for consistency / standardiseation.
- 'to ensure compatibility identical prompts were used for both models' 
(described in full in the appendix - link to section)

Example table:

```{r llm-comparison, echo=FALSE}
# define symbols
sym <- list(
  Y = "$\\checkmark$",    # ● addressed unprompted
  P = "$\\circ$",         # ○ partial / after prompting
  N = "\\texttimes"       # ✗ missed entirely
)

sym_legend <- paste0(
  sym$Y, "~= addressed unprompted; ",
  sym$P, "~= partial; ",
  sym$N, "~= missed entirely."
)

# data structure 
# > can replace "?" with Y/P/N after running the LLM experiment

rows <- tribble(
  ~group,            ~label,                                          ~ours, ~chatgpt, ~claude,
  "Preprocessing",   "Excluded Channel \\& Region",                   "Y",   "?",      "?",
  "Preprocessing",   "Identified skewness in spending variables",      "Y",   "?",      "?",
  "Preprocessing",   "Applied log transform",                         "Y",   "?",      "?",
  "Preprocessing",   "TBC",  "Y",   "?",      "?",
  "PCA",             "Scree plot / variance explained",                "Y",   "?",      "?",
  "PCA",             "Interpreted PC1 as channel axis",                "Y",   "?",      "?",
  "PCA",             "TBC",         "Y",   "?",      "?",
  "Clustering",      "TBC",       "Y",   "?",      "?",
  "Clustering",      "TBC",    "Y",   "?",      "?",
  "Clustering",      "TBC",              "Y",   "?",      "?",
  "Clustering",      "TBC",       "Y",   "?",      "?",
  "Interpretation",  "Commercial framing of segments",                 "Y",   "?",      "?",
  "Interpretation",  "Acknowledged limitations",                       "Y",   "?",      "?",
  "Interpretation",  "Suggested alternative methods",                  "Y",   "?",      "?"
)

# map codes to symbols
map_sym <- function(x) {
  dplyr::case_match(x,
    "Y" ~ sym$Y,
    "P" ~ sym$P,
    "N" ~ sym$N,
    .default = x
  )
}

# compute totals pre symbol mapping
score_col <- function(col) {
  n_total <- sum(col %in% c("Y", "P", "N"))
  if (n_total == 0) return("---")
  n_y <- sum(col == "Y")
  n_p <- sum(col == "P")
  # format: "10/14 (2P)" or "14/14" if no partials
  if (n_p > 0) {
    paste0(n_y, "/", n_total, " (", n_p, "P)")
  } else {
    paste0(n_y, "/", n_total)
  }
}

totals <- tibble(
  group   = "Totals",
  label   = "\\textbf{Score}",
  ours    = score_col(rows$ours),
  chatgpt = score_col(rows$chatgpt),
  claude  = score_col(rows$claude)
)

# combine rows + totals, then map symbols
all_rows <- bind_rows(rows, totals) %>%
  mutate(across(c(ours, chatgpt, claude), map_sym))

#compute pack_rows indices from group column
group_indices <- all_rows %>%
  mutate(row = row_number()) %>%
  group_by(group) %>%
  summarise(start = min(row), end = max(row), .groups = "drop")

group_indices <- group_indices %>%
  mutate(group = factor(group, levels = unique(all_rows$group))) %>%
  arrange(group)

# build kable
kable_data <- all_rows %>%
  dplyr::select(label, ours, chatgpt, claude)

colnames(kable_data) <- c("Analysis Step", "Our Analysis", "ChatGPT (4o)", "Claude")

caption_text <- paste0(
  "Comparison of analytical steps: our pipeline vs.\\ two LLMs. ",
  sym_legend
)

tbl <- kable(
  kable_data,
  format    = "latex",
  booktabs  = TRUE,
  escape    = FALSE,
  align     = c("l", "c", "c", "c"),
  caption   = caption_text,
  linesep   = ""
) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size     = 10
  )

# -apply pack_rows 
for (i in seq_len(nrow(group_indices))) {
  grp <- as.character(group_indices$group[i])
  # skip the totals row - it stands alone with a rule above it
  if (grp == "Totals") next
  tbl <- tbl %>%
    pack_rows(
      grp,
      group_indices$start[i],
      group_indices$end[i],
      bold         = TRUE,
      italic       = FALSE,
      hline_before = (i > 1)
    )
}

# if (knitr::is_latex_output()) {
#   # cat("\\begingroup\n")
#   cat("\\renewcommand{\\arraystretch}{1.3}\n")
# }

tbl 

# if (knitr::is_latex_output()) {
#   # cat("\\endgroup\n")
# }
# # tbl
```

```{=latex}
\newpage
```

<!-- ```{=html} -->
<!-- # References -->
<!-- <div id="refs"></div> -->
<!-- ``` -->

<!-- ```{=latex} -->
<!-- \bibliography{references.bib} -->
<!-- ``` -->

```{=latex}
\newpage
\section*{Appendices}
\addcontentsline{toc}{section}{Appendices}

% --- restore default section numbering in the appendix ---
\renewcommand{\thesection}{\arabic{section}}
\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
\renewcommand{\thesubsubsection}{\thesubsection.\arabic{subsubsection}}
```

# (APPENDIX) Appendices {-}

# Prompts and LLM outputs {#apd-llm-information}

In this appendix, we opted to use a consistent visual format to distinguish
between prompts sent to the LLM, its responses, and our own commentary. The key
elements we decided to use are:

::: {.promptbox}

**Prompts to the LLM** - shown with a right-hand border. These are the instructions and
questions we provided to the model.

:::

:::: {.responsebox}

**LLM Response** - shown with a left-hand border, forming a conversational
flow with the prompt. Responses may include a mix of explanation and code:
```r
summary(mtcars$mpg)
```

::::

::: {.insightbox}

**Insight and critical interpretation** - highlighted observations or key findings drawn from the output.

:::

::: {.alertbox}

**Alert** - flags potential concerns, limitations, or caveats with the
generated output.

:::


## \@ref(qone) PCA LLM dialogue

_Example format below (literally asked base ChatGPT to come up with it's own
prompt and then answer it - change when we do for real)._

<!-- prettier-ignore -->
::: promptbox 
**LLM Prompt:** Act as an expert data scientist. I have the
Wholesale Customers dataset from Kaggle. Please write an R script to perform
Principal Component Analysis (PCA) on this dataset. Provide the code, generate
the necessary plots, and summarise the key findings. 
:::

<!-- prettier-ignore -->
:::: {.responsebox}

Here is the LLM's generated code:
```r
# =========================
# 0. Setup
# =========================
rm(list = ls())
packages <- c("tidyverse", "factoextra", "corrplot")
installed <- packages %in% rownames(installed.packages())
if(any(!installed)) install.packages(packages[!installed])
```

Which handles string formatting cleanly.

::::

<!-- prettier-ignore -->
::: insightbox 
**Our interpretation:** The PCA biplot shows that Fresh and Milk
spending are nearly orthogonal, suggesting they are independent purchasing
behaviors. 
:::

<!-- prettier-ignore -->
::: alertbox 
**Our concern:** PCA assumes linear relationships between
variables. If the underlying structure is non-linear, consider using t-SNE or
UMAP instead. Also note that the results are sensitive to outliers - always
check for extreme values before interpreting the components. 
:::


## \@ref(qtwo) Clustering LLM dialogue

```{=latex}
\clearpage
\newpage
```

# Supplementary methodology {#apd:suppl-methods}

## Raw data distribution, correlations, and skewness {#apd:suppl-skew}

```{r histograms-overlay, echo=FALSE, include=TRUE, fig.width=10, fig.height=7, fig.cap="Shape comparison of raw and log-transformed spending distributions by product category. To facilitate direct comparison of the distribution shapes, all spend values were min-max scaled to a 0–1 range within their respective categories. The raw data (blue) demonstrates a right skew across all categories, influenced by extreme outliers. The `log1p` transformation (red) effectively compresses this long tail, yielding more symmetric distributions that ensure variance-based methods (PCA) and distance-based methods (clustering) are not dominated by extreme outliers."}

bind_rows(raw_df, log_df) %>%
  mutate(
    Transform = factor(Transform, levels = c("Raw", "Log-transformed"))
  ) %>%
  group_by(Category, Transform) %>%
  mutate(Spend_scaled = rescale01(Spend)) %>%
  ungroup() %>%
  ggplot(aes(x = Spend_scaled, fill = Transform)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 40,
    colour = "white",
    alpha = 0.5,
    position = "identity"
  ) +
  scale_fill_manual(
    values = c("Raw" = "steelblue", "Log-transformed" = "firebrick")
  ) +
  facet_wrap(~Category, scales = "free_y") +
  labs(
    title = "Shape Comparison: Raw vs Log-transformed Spending",
    x = "Normalised Spend (0–1 scaled within each group)",
    y = "Density"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

The overlapping histograms in Figure \@ref(fig:histograms-overlay) highlight the
rationale for transforming the spending data. The raw spending figures exhibit a
strong rightward skew across all six product categories, indicating that while
most customers spend relatively small amounts, a long tail of high-spending
customers dominates the scale. Applying a log transformation (`log1p`)
successfully 'pulls in' these extreme values, centring the bulk of the data and
producing much more symmetric distributions.


```{r corrplots-side-by-side, echo=knitr::is_html_output(), include=TRUE, fig.width=12, fig.height=5.5, fig.cap="Pairwise correlation matrices of the six product categories, comparing linear relationships using raw spending data (left) versus log-transformed spending data (right)."}
par(mfrow = c(1, 2))

corrplot(
  cor(spending),
  method      = "color",
  type        = "upper",
  addCoef.col = TRUE,
  number.cex  = 0.8,
  tl.col      = "black",
  title       = "Raw spending",
  mar         = c(0, 0, 2, 0)
)

corrplot(
  cor(spending_log),
  method      = "color",
  type        = "upper",
  addCoef.col = TRUE,
  number.cex  = 0.8,
  tl.col      = "black",
  title       = "Log-transformed spending",
  mar         = c(0, 0, 2, 0)
)
```

The side-by-side correlation plots reveal the underlying purchasing
relationships between product categories before and after adjusting for scale.
Both matrices highlight a distinct cluster of highly correlated spending:
`r format_var_name("Grocery")`, `r format_var_name("Detergents_Paper")`, and
`r format_var_name("Milk")` show strong positive linear relationships,
suggesting they are frequently purchased together in bulk. While the log
transformation preserves this core structure, it slightly softens the most
extreme correlations (for instance, the correlation between
`r format_var_name("Grocery")` and `r format_var_name("Detergents_Paper")`
drops from `r formatC(cor(spending$Grocery, spending$Detergents_Paper), format = "f", digits = 2)`
to `r formatC(cor(spending_log$Grocery, spending_log$Detergents_Paper), format = "f", digits = 2)`).
This indicates that in the raw data, a few extreme outlier purchases were
artificially inflating the strength of that linear relationship.


## Multi-dimensional scaling {#apd:suppl-mds}

```{r mds-side-by-side, echo=FALSE, include=TRUE, fig.width=12, fig.height=5.5, fig.cap="Non-metric MDS using Gower’s distance across all eight variables (six log-transformed continuous spending categories plus Channel and Region). Points are coloured by Channel (left) and Region (right). Unlike PCA, Gower’s distance forces distinct spatial separation based on categorical mismatches, resulting in six artificial clusters corresponding to the cross-tabulation of the two Channels and three Regions."}

# Gower distance on all 8 variables
mds_data <- data.frame(
  Channel = factor(data$Channel, labels = c("HoReCa", "Retail")),
  Region = factor(data$Region, labels = c("Lisbon", "Oporto", "Other")),
  spending_log
)

gow <- cluster::daisy(mds_data, metric = "gower")
mds_fit <- MASS::isoMDS(gow, k = 2, trace = FALSE)
mds_pts <- as.data.frame(mds_fit$points)
colnames(mds_pts) <- c("MDS1", "MDS2")

mds_pts$Channel <- mds_data$Channel
mds_pts$Region <- mds_data$Region

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Left: coloured by Channel
plot(
  mds_pts$MDS1,
  mds_pts$MDS2,
  col = ifelse(mds_pts$Channel == "HoReCa", "#E64B35", "#000000"),
  # 19 = Circle, 17 = Triangle
  pch = ifelse(mds_pts$Channel == "HoReCa", 19, 17),
  cex = 0.8,
  xlab = "MDS1",
  ylab = "MDS2",
  main = "(a) Coloured by Channel"
)
legend(
  "topleft",
  legend = c("HoReCa", "Retail"),
  col = c("#E64B35", "#000000"),
  pch = c(19, 17),
  cex = 0.9,
  bty = "n"
)

# Right: coloured by Region
region_cols <- c(Lisbon = "#E69F00", Oporto = "#56B4E9", Other = "#CC79A7")
plot(
  mds_pts$MDS1,
  mds_pts$MDS2,
  col = region_cols[mds_pts$Region],
  pch = 19,
  cex = 0.8,
  xlab = "MDS1",
  ylab = "MDS2",
  main = "(b) Coloured by Region"
)
legend(
  "topleft",
  legend = names(region_cols),
  col = region_cols,
  pch = 19,
  cex = 0.9,
  bty = "n"
)
# mds_fit$stress
```

Figure \@ref(fig:mds-side-by-side) presents a non-metric MDS solution computed
from Gower's distance across all eight variables, incorporating the categorical
`r format_var_name("Channel")` and `r format_var_name("Region")` columns excluded
from the PCA. Because Gower's metric treats any categorical mismatch as a
maximal distance contribution on that variable, the resulting spatial map is
dominated by these labels rather than underlying spending behaviour.

Panel (a) shows separation by `r format_var_name("Channel")` along the primary
axis (MDS1), while Panel (b) reveals further stratification by 
`r format_var_name("Region")` along MDS2. The result is a forced stratification
into six spatial clusters reflecting the Channel $\times$ Region
cross-tabulation. While there is minor regional intermingling (e.g., two Oporto
customers mapping near the Lisbon cluster), the overall spatial map merely
reflects the cross-tabulation of the categorical labels rather than organic,
dataset-wide purchasing patterns. The solution's stress value of
`r formatC(mds_fit$stress, format = "fg", digits = 3)` further indicates a 
mediocre fit, falling above the conventional threshold of 10 for reliable
low-dimensional representation. This confirms that the PCA approach, relying
purely on continuous spending variables, is better suited to uncovering natural
commercial structure in this dataset.


# Reproducibility, accessibility & declarations

<!-- BE note: taken out as we are submitting R file separately -->
<!-- Use: `knitr::purl("cs1-group07.rmd", documentation = 0)` to generate r file -->
<!-- Use: `tools::showNonASCIIfile("cs1-group07.rmd")` to debug -->
<!-- ## R Code Documentation -->
<!-- ```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
``` -->

## Reproducibility & accessibility

An accessible HTML version of this report is available via a public GitHub page:

- https://ytterbiu.github.io/SMM636-ML-group-coursework-03-g01/

This report was created in R Markdown. The source code is open source and is
available via:

- https://github.com/ytterbiu/SMM636-ML-group-coursework-03-g01

Changes made to this document were tracked using Git and are also available via
the same repository
(https://github.com/ytterbiu/SMM636-ML-group-coursework-03-g01/commits/main/).
Please note that this commit history reflects updates made to the project
repository and is not a comprehensive timeline of all contributions by group
members.

## Word count

The total word count for the core sections of this report is **`r word_count`**
words (`r ifelse(word_count > 1000, "above", "below")` the 1000 word limit).



The calculation was performed programmatically directly from the source `.rmd`
file. An **R** script was used to read the raw document lines and strip out
non-text elements, including the YAML front matter, R code chunks, sub-headers,
markdown formatting (tables and figures), and LaTeX/HTML syntax. The script
strictly isolates lines falling within the core analytical sections (`#intro`,
`#qone`, `#qtwo`, and `#qthr`) before calculating the final tally using the
`stringi` package.

The code chunk that performs this calculation is executed silently at the
beginning of this document, but is reproduced below for transparency:

```{r display-word-count-code, ref.label="word-count", eval=FALSE, echo=TRUE}
```

## Generative AI Declaration

- Will be best to ask about this during the lecture - will almost want this for
  the first part - or is the first part completely AI free?

<!-- Generative AI (GenAI) tools were used throughout this project to assist with -->
<!-- project planning, methodological validation, and debugging. In practice, these -->
<!-- tools functioned as an additional technical team member, providing immediate -->
<!-- feedback, troubleshooting code, and challenging statistical assumptions. -->

<!-- ### Tools Used -->

<!-- Google Gemini 3 Pro (Deep Research), ChatGPT 5.2 -->

<!-- ### Prompts used -->

<!-- The GenAI tools were provided with the 'coursework_CM2_250120.pdf' file and -->
<!-- extracts from our `R` code. -->

<!-- A non-exhaustive list of prompts used is shown below. We opted to include at -->
<!-- least one example of each _type_ of prompt used, rather than a list of all -->
<!-- prompts. For example, if we entered the prompts 'How can we put two ggplots side -->
<!-- by side?' and 'How to remove grey background on a ggplot', we would include only -->
<!-- the first of these in the list below, as both relate to plotting issues (& are -->
<!-- similar to 'querying' with a traditional search engine). -->

<!-- - "...the top of both graphs in r are slightly different. (Pasting R code.) Can -->
<!--   this be fixed for the plot area?" -->
<!-- - "Help to identify the key points here" -->
<!-- - "How can this error be tracked down in an rmd file? ! Text line contains an -->
<!--   invalid character. l.1 ^^@^^@..." -->

<!-- ### Use of GenAI outputs & subsequent revisions -->

<!-- The GenAI tools served as a "consultant" member of the group. Notably, the AI -->
<!-- output: -->

<!-- - Helped debug plotting issues. -->
<!-- - Commented on the structure of R code snippets and the report structure. -->

<!-- ### Changes made: -->

<!-- All AI suggestions were critically reviewed and debated by the group before -->
<!-- implementation. For example: -->

<!-- - While the AI suggested code for complex layouts, the specific aesthetic -->
<!--   choices and annotations were manually adjusted to ensure clarity. -->
<!-- - Where the AI proposed specific interpretations, these were cross-referenced -->
<!--   with academic literature and statistics textbooks. -->
<!-- - The final R code for implementation was written and debugged by the group with -->
<!--   AI snippets being treated as drafts rather than final solutions. -->

<!-- While GenAI was treated as a collaborator, the final analysis, statistical -->
<!-- interpretation, and the written content of this report were generated and -->
<!-- verified by the (human) group members. -->

---

# References {-}