---
title:
  "Beyond the binary: Uncovering hidden customer segments in wholesale purchasing data"
subtitle: "SMM636 (2025-26) Coursework 03 - Group 01"
author:
  - "Abdulrahman Alolyan"
  - "Benjamin Evans"
  - "Basmah Khan"
  - "Ardi Wira Sudarmo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
papersize: a4
fontsize: 12pt
output:
  bookdown::html_document2:
    self_contained: true
    css: style.css
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: paper
    highlight: pygments
    df_print: paged
    pandoc_args: ["--lua-filter=custom-boxes.lua"]
  # runtime: shiny
  bookdown::pdf_document2:
    latex_engine: lualatex
    includes:
      in_header: preamble.tex
    pandoc_args: ["--lua-filter=custom-boxes.lua"]
    number_sections: true
    toc: true
    # citation_package: natbib
    keep_tex: true
bibliography: references.bib
csl: ieee.csl
---

```{=latex}
\newpage
```

```{=html}
<h1>Information</h1>

This R Markdown document was created as part of a group assignment for SMM636 at Bayes Business School, City St George's, University of London in Term 2 2025-26.

```

```{r asthetic-header, include=FALSE, echo=FALSE}
# ============================================================================ #
# Key Information ====
# SMM636 Machine Learning
# Group Coursework 2025-26
# Group:        Group 01
# Authors (in alphabetical order):
#   - Abdulrahman Alolyan
#   - Benjamin Evans
#   - Basmah Khan
#   - Ardi Wira Sudarmo
# Professor:    Dr. Rui Zhu
# Institution:  Bayes Business School - City St George's, University of London
# Date:         03 Apr 2026
#
# Description:  Term 2 group project for SMM636 Machine Learning
#               (25% of coursework grade - 25% of module grade). The R code
#               below has been exported directly from an R Markdown (.rmd) file.
#               Hence the knitr settings.
#
# Dependencies:
#   - tidyverse: loads dplyr, tidyr, ggplot2, purrr (& others)
#   - TBI
#
# Acknowledgements
#   - TBI
# ============================================================================ #
```

```{r setup-knitr, include=FALSE, echo=FALSE}
# Setup & configuration ====
#----------------------- Initial setup (knitr settings) -----------------------#
dir.create("fig", showWarnings = FALSE)

# Defaults common to all outputs
knitr::opts_chunk$set(
  echo     = TRUE,
  message  = FALSE,
  warning  = FALSE,
  fig.align = "center",
  out.width = "100%",
  fig.path  = "fig/",
  dpi       = 300
)

# Output-specific settings
if (knitr::is_latex_output()) {
  knitr::opts_chunk$set(
    fig.width = 6,
    fig.height = 4,
    dev = "pdf",
    fig.pos = "!ht",
    out.extra = ""
  )
} else {
  knitr::opts_chunk$set(
    fig.width = 6,
    fig.height = 4,
    dev = "svglite"  # or "png"
  )
}
```

```{r setup-qol, include=FALSE, echo=FALSE}
#----------------------------- Clean environment ------------------------------#
rm(list = ls()) # Remove all objects
graphics.off() # Close all graphical devices
cat("\014") # Clean console
```

```{r load-dependencies, include=FALSE, echo=FALSE}
#------------------- Load dependencies / external libraries -------------------#
# --- Core Data Manipulation ---
library(tidyverse)  # loads dplyr, tidyr, ggplot2, purrr, etc.

# --- Visualization & Plot Arrangement ---
library(patchwork)  # combining multiple ggplot objects
library(gridExtra)  # arranging base/grid plots
library(ggpubr)     # publication-ready plot themes/utilities
library(corrplot)   # correlation heatmaps
library(GGally)     # pairs plots
library(factoextra) # PCA & clustering visualizations

# --- Analytics & Modeling ---
library(NbClust)    # optimal-k diagnostics

# NOTE: The packages below are deliberately NOT loaded into the global 
# environment to prevent function masking conflicts (e.g., MASS::select 
# breaking dplyr::select). Call their functions directly using namespaces:
# cluster::daisy(), MASS::isoMDS()
# library(MASS)     
# library(cluster)

library(mclust)

library(ggsankeyfier)
library(Polychrome)

# --- Reporting & Tables ---
library(kableExtra) # for formatting output tables
```

```{r custom-functions, include=FALSE, echo=FALSE}
#---------------------------- Custom QOL functions ----------------------------#
#####################################
# function: banner comments (used to to section up code)
# Usage: banner_comment("Element 1: data cleaning") -> then ctrl + v (or cmd+v)
#####################################
banner_comment <- function(text, width = 80, border = "#", fill = "-") {
  txt <- paste0(" ", text, " ")
  inner_width <- width - 2 * nchar(border)
  banner_string <- ""

  if (inner_width <= nchar(txt)) {
    banner_string <- paste0(border, txt, border)
  } else {
    pad_total <- inner_width - nchar(txt)
    pad_left <- pad_total %/% 2
    pad_right <- pad_total - pad_left

    banner_string <- paste0(
      border,
      strrep(fill, pad_left),
      txt,
      strrep(fill, pad_right),
      border
    )
  }

  cat(banner_string, "\n")
  # copy banner to allow direct pasting (requires clipr)
  clipr::write_clip(banner_string)
  # avoid [1] when printing if want to manually copy
  invisible(banner_string)
}
#####################################
# function: format p-values for text
# Usage (in-line): `r format_p_vals(ad_test_result$p.value)`
# Usage (console): format_p_vals(ad_test_result$p.value)
#####################################
format_p_vals <- function(p) {
  if (length(p) != 1L || is.na(p)) {
    stop("Error! p must be a single non-missing value")
  }
  if (p > 1) {
    stop("Error! Value greater than 1")
  }
  if (p < 0) {
    stop("Error! Value less than 0")
  }

  if (p >= 0.01) {
    paste0("= ", formatC(p, format = "f", digits = 2))
  } else if (p >= 0.001) {
    paste0("= ", formatC(p, format = "f", digits = 3))
  } else {
    "< 0.001"
  }
}
#####################################
# function: format confidence intervals for tables & text
# Usage (in-line): `r format_interval(el2_ci_normal_95[1], el2_ci_normal_95[2])`
# Usage (console): format_interval(el2_ci_normal_95[1], el2_ci_normal_95[2])
#####################################
format_interval <- function(lower, upper, digits = 3, small_interval = 5L) {
  paste0(
    "[",
    formatC(
      lower,
      format = "f",
      digits = digits,
      small.interval = small_interval,
      small.mark = " "
    ),
    ", ",
    formatC(
      upper,
      format = "f",
      digits = digits,
      small.interval = small_interval,
      small.mark = " "
    ),
    "]"
  )
}
#####################################
# function: format truncated ellipses
# Usage (in-line): `r tbi`
# Usage (console): tbi
#####################################
fmt_trunc_ellip <- function(
  x,
  digits = 7,
  ellip = "...",
  tol = 1e-12,
  trim_zeros_if_exact = TRUE
) {
  out <- rep(NA_character_, length(x))
  ok <- is.finite(x)
  scale <- 10^digits
  xt <- trunc(x[ok] * scale) / scale
  s <- formatC(xt, format = "f", digits = digits)
  add <- abs(x[ok] - xt) > tol * pmax(1, abs(x[ok]))
  if (trim_zeros_if_exact) {
    s_trim <- sub("0+$", "", s) # drop trailing zeros
    s_trim <- sub("\\.$", "", s_trim) # drop trailing decimal point
  } else {
    s_trim <- s
  }
  out[ok] <- ifelse(add, paste0(s, ellip), s_trim)
  out[is.na(x)] <- NA_character_
  out
}
#####################################
# function: function to format 6dp numbers with a space after the 3rd decimal
# digit
# Usage (in-line): TBI
# Usage (console): TBI
#####################################
format_spaced_6dp <- function(x) {
  # Format to 6 decimal places
  s <- sprintf("%.6f", x)
  # Insert space between the first 3 and last 3 decimal digits
  sub("(\\.[0-9]{3})([0-9]{3})$", "\\1 \\2", s)
}
#####################################
# function: format variable names in green in latex (& normal code in html)
# Usage (in-line): TBI `r format_var_name("dvisit")`
#####################################
format_var_name <- function(x) {
  if (knitr::is_latex_output()) {
    # replace all "_" with "\_" in va name
    paste0("\\greentt{", gsub("_", "\\\\_", x), "}")
  } else {
    # change to paste0("<code style='color:green'>", x, "</code>") for green
    # paste0("`", x, "`")
    paste0("<code style='color:green'>", x, "</code>")
  }
}
```

```{r word-count, include=FALSE}
rmd_lines <- readLines("01SMM363-assessment-03.rmd", warn = FALSE)

# ---- Remove YAML front matter ----
yaml_bounds <- which(rmd_lines == "---")
if (length(yaml_bounds) >= 2) {
  rmd_lines <- rmd_lines[-(yaml_bounds[1]:yaml_bounds[2])]
}

# ---- Remove code chunks ----
in_chunk <- FALSE
keep <- rep(TRUE, length(rmd_lines))
for (i in seq_along(rmd_lines)) {
  if (grepl("^```\\{", rmd_lines[i])) {
    in_chunk <- TRUE
    keep[i] <- FALSE
    next
  }
  if (in_chunk && grepl("^```\\s*$", rmd_lines[i])) {
    in_chunk <- FALSE
    keep[i] <- FALSE
    next
  }
  if (in_chunk) keep[i] <- FALSE
}
rmd_lines <- rmd_lines[keep]

# ---- Identify which lines fall inside counted sections ----
# Sections to include (level-1 headers only)
counted_ids <- c("intro", "qone", "qtwo", "qthr")
in_counted_section <- FALSE
include <- rep(FALSE, length(rmd_lines))

for (i in seq_along(rmd_lines)) {
  line <- rmd_lines[i]
  # Detect any level-1 header (# but not ##)
  if (grepl("^#\\s+[^#]", line)) {
    # Check if this header belongs to a counted section
    in_counted_section <- any(sapply(counted_ids, function(id) {
      # [^}]* allows any characters (like spaces, hyphens, classes) before the 
      # closing }
      # this is required to count the words in the unnumbered introduction
      grepl(paste0("\\{\\s*#", id, "[^}]*\\}"), line)
    }))
    next  # skip the header line itself
  }
  if (in_counted_section) include[i] <- TRUE
}

counted_lines <- rmd_lines[include]

# ---- Remove sub-headers (##, ###, etc.) ----
counted_lines <- counted_lines[!grepl("^#{2,}", counted_lines)]

# ---- Remove table lines (markdown pipes) ----
counted_lines <- counted_lines[!grepl("^\\|", counted_lines)]

# ---- Remove figure captions: ![caption](...) ----
counted_lines <- counted_lines[!grepl("^!\\[", counted_lines)]

# ---- Remove fenced div markers (::: name) ----
counted_lines <- counted_lines[!grepl("^:::", counted_lines)]

# ---- Remove LaTeX commands ----
counted_lines <- counted_lines[!grepl("^\\\\", counted_lines)]

# ---- Remove HTML comments ----
counted_lines <- counted_lines[!grepl("^<!--", counted_lines)]

# ---- Remove inline R references like \@ref() ----
counted_lines <- gsub("\\\\@ref\\([^)]*\\)", "", counted_lines)

# ---- Count ----
word_count <- sum(stringi::stri_count_words(counted_lines), na.rm = TRUE)

message("Word count (sections qone, qtwo, qthr, body text only): ", word_count)

# Write for LaTeX preamble
writeLines(
  paste0("\\newcommand{\\wordcount}{",
         formatC(word_count, big.mark = ","), "}"),
  "wordcount.tex"
)
```

```{=latex}
\newpage
\pagenumbering{arabic}
```

# Introduction {#intro -}

This report analyses annual spending data from 440 clients of a Portuguese
wholesale distributor [@Cardoso2013; @Abreu2011], comprising 298
Hotel/Restaurant/Café (HoReCa) and 142 Retail customers across six product
categories: `r format_var_name("Fresh")`,
`r format_var_name("Milk")`, `r format_var_name("Grocery")`,
`r format_var_name("Frozen")`, `r format_var_name("Detergents_Paper")`, and
`r format_var_name("Delicassen")`. Data-driven
customer segmentation informs inventory planning, pricing strategy, and
sales team deployment, yet the existing classification is a simple binary label
(`r format_var_name("Channel")`).
This analysis uses Principal Component Analysis (PCA) and clustering to
investigate whether the existing two-category classification adequately captures
the structure of customer spending, or whether the data supports a more nuanced
segmentation.

```{r load-data, echo=FALSE, include=FALSE}
set.seed(1234)

data <- read.csv("data/Wholesale_customers_data.csv")
str(data)
summary(data)

# just get the $ amounts spent
spending <- data[, c(
  "Fresh",
  "Milk",
  "Grocery",
  "Frozen",
  "Detergents_Paper",
  "Delicassen"
)]

spending_log <- log1p(spending)

rescale01 <- function(x) (x - min(x)) / (max(x) - min(x))

raw_df <- spending %>%
  pivot_longer(everything(), names_to = "Category", values_to = "Spend") %>%
  mutate(Transform = "Raw")

log_df <- spending_log %>%
  pivot_longer(everything(), names_to = "Category", values_to = "Spend") %>%
  mutate(Transform = "Log-transformed")

# add data to import into in-line text
# standardise log-transformed data
scaled_data <- scale(spending_log)

# PCA
pca_result <- prcomp(spending_log, center = TRUE, scale. = TRUE)

# Skewness (raw)
skew_vals <- apply(spending, 2, e1071::skewness)
min_skew  <- min(skew_vals)
```

```{r data-table, echo=knitr::is_html_output(), eval=knitr::is_html_output()}
DT::datatable(
  data,
  filter   = "top",
  options  = list(pageLength = 15, scrollX = TRUE),
  caption  = "Wholesale customers"
)
```

# Principle component analysis (PCA) {#qone}

<!-- _Apply PCA to the dataset and summarise and interpret key your findings_ -->

A log-transformation was applied to normalise the distribution and
address the right-skewness observed across all six spending variables 
(skewness > `r formatC(min_skew, format = "f", digits = 1)`
for all - see Appendix, Figures \@ref(fig:histograms-overlay) &
\@ref(fig:corrplots-side-by-side)). The variables were then standardised to
ensure equal contribution to the PCA, preventing categories with naturally
larger absolute expenditures from dominating the model.

The first two principal components capture 71.3% of the total variance (PC1:
44.1%; PC2: 27.2%; Figure \@ref(fig:pca-grid)(a)). The component loadings in
Figure \@ref(fig:pca-grid)(b) reveal distinct commercial purchasing patterns.
PC1 acts as the primary "channel axis," driven by strong positive loadings for
`r format_var_name("Grocery")` (0.571),
`r format_var_name("Detergents_Paper")` (0.551), and 
`r format_var_name("Milk")` (0.542), 
contrasted with near-zero or negative loadings for 
`r format_var_name("Fresh")` and `r format_var_name("Frozen")`. 
PC2 serves as a "perishable volume axis," characterised by high positive
loadings for `r format_var_name("Fresh")` (0.590), `r format_var_name("Frozen")`
(0.590), and `r format_var_name("Delicassen")` (0.530), capturing the scale of
fresh food purchasing independent of business type.

Figure \@ref(fig:pca-grid)(d) confirms this visually, showing clear
separation between HoReCa and Retail customers along PC1. PCA is restricted to
continuous inputs, necessitating the exclusion of `r format_var_name("Channel")`
and `r format_var_name("Region")`. A Gower-distance MDS analysis incorporating
these variables (Appendix, \@ref(apd:suppl-mds)) confirms that the categorical
labels dominate the distance metric, validating the decision to analyse spending
variables independently.


```{r pca-fit, echo=FALSE, include=FALSE}
# ==============================================================================
# Q1 PCA
# ==============================================================================

pca_result <- prcomp(spending_log, center = TRUE, scale. = TRUE)
summary(pca_result)

cumvar <- cumsum(pca_result$sdev^2 / sum(pca_result$sdev^2))
cat("Cumulative variance explained:\n")
print(round(cumvar, 3))

loadings <- pca_result$rotation[, 1:3]
print(round(loadings, 3))
```

```{r pca-scree, echo=FALSE, include=FALSE, fig.width=7, fig.height=5}
fviz_eig(
  pca_result,
  addlabels = TRUE,
  ylim = c(0, 60),
  main = "Scree Plot - Variance Explained by Each Principle Component (PC)"
)
```

```{r pca-biplot, echo=FALSE, include=FALSE, fig.width=8, fig.height=6}
fviz_pca_biplot(
  pca_result,
  label   = "var",
  col.var = "firebrick",
  col.ind = alpha("steelblue", 0.3),
  repel   = TRUE,
  title   = "PCA Biplot (log-transformed, standardised data)"
)
```

```{r pca-contrib-pc, echo=FALSE, include=FALSE, fig.width=7, fig.height=4}
# Opted to hide these for now 
# (don't think we have space and they don't add too much)
# Even including in appendix is limited - easier to include table
fviz_contrib(
  pca_result,
  choice = "var",
  axes   = 1,
  title  = "Variable Contributions to PC1"
)

fviz_contrib(
  pca_result,
  choice = "var",
  axes   = 2,
  title  = "Variable Contributions to PC2"
)
```

```{r pca-loadings-table, echo=FALSE, include=FALSE}
loadings_df <- as.data.frame(round(pca_result$rotation[, 1:3], 3))
colnames(loadings_df) <- c("PC1", "PC2", "PC3")

kable(
  loadings_df,
  caption  = "PCA loadings (first three components). Values over 0.4 are highlighted in bold.",
  linesep  = "",
  booktabs = TRUE
) %>%
  kable_styling(
    full_width = FALSE,
    position   = "center"
  ) %>%
  column_spec(2, bold = abs(loadings_df$PC1) >= 0.4) %>%
  column_spec(3, bold = abs(loadings_df$PC2) >= 0.4) %>%
  column_spec(4, bold = abs(loadings_df$PC3) >= 0.4)
```

```{r pca-channel, echo=FALSE, include=FALSE, fig.width=8, fig.height=6}
fviz_pca_ind(
  pca_result,
  geom.ind    = "point",
  col.ind     = factor(data$Channel),
  palette     = c("#E64B35", "#4DBBD5"),
  addEllipses = TRUE,
  legend.title = "Channel",
  title       = "PCA - Individuals Coloured by Channel (1=HoReCa, 2=Retail)"
)
```

```{r pca-grid, echo=FALSE, include=TRUE, fig.width=12, fig.height=10, fig.cap="Principal Component Analysis of log-transformed, standardised spending data. (a) Scree plot showing the percentage of variance explained by each component. (b) Loadings of the original variables on the first four principal components. Values exceeding 0.4 in magnitude indicate strong contributions. (c) Biplot overlaying variable loadings (red arrows) on individual observations. (d) Individuals projected onto PC1-PC2, coloured by distribution channel (1 = HoReCa, 2 = Retail) with 95% concentration ellipses."}

# --- Top left: Scree plot ---
p1 <- fviz_eig(
  pca_result,
  addlabels = TRUE,
  ylim = c(0, 60),
  main = "(a) Scree Plot - Variance Explained by Principle Component (PC)"
)

# --- Top right: Loadings table as a grob ---
loadings_df <- as.data.frame(round(pca_result$rotation[, 1:3], 3))
colnames(loadings_df) <- c("PC1", "PC2", "PC3")

# --- Top right: Loadings table ---
loadings_df <- as.data.frame(pca_result$rotation[, 1:4])
colnames(loadings_df) <- c("PC1", "PC2", "PC3", "PC4")
# loadings_df <- as.data.frame(pca_result$rotation[, 1:3])
# colnames(loadings_df) <- c("PC1", "PC2", "PC3")
loadings_df[] <- lapply(loadings_df, function(x) format(round(x, 3), nsmall = 3))

p2 <- ggtexttable(
  loadings_df,
  theme = ttheme(
    "blank",
    tbody.style = tbody_style(
      size  = 13,
      hjust = 1.00, #controls alignment {0.00, 0.5, 1.00} 
      x     = 0.95, #controls alignment {0.05, 0.5, 0.95} 
      fill  = "white"
    ),
    colnames.style = colnames_style(
      size  = 13,
      face  = "bold",
      hjust = 1.00,
      x     = 0.95,
      fill  = "white"
    ),
    rownames.style = rownames_style(
      size  = 13,
      face  = "plain",
      hjust = 0,
      x     = 0.05,
      fill  = "white"
    )
  )
) %>%
  tab_add_hline(at.row = 1:2, row.side = "top", linewidth = 2) %>%
  tab_add_hline(at.row = nrow(loadings_df) + 1, row.side = "bottom", linewidth = 2) +
  labs(title = "(b) PCA Loadings") +
  theme(
    plot.title      = element_text(hjust = 0.5, size = 13),
    plot.background = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "white", colour = NA)
  )

# --- Bottom left: Biplot ---
p3 <- fviz_pca_biplot(
  pca_result,
  label   = "none",
  col.var = "firebrick",
  col.ind = alpha("steelblue", 0.3),
  title   = "(c) PCA Biplot (log-transformed, standardised data)"
)

# Pull arrow endpoints from layer 4 of the built plot
arrows <- ggplot_build(p3)$data[[4]]

var_df <- data.frame(
  x     = arrows$xend,
  y     = arrows$yend,
  label = rownames(pca_result$rotation)
)

p3 <- p3 +
  ggrepel::geom_text_repel(
    data        = var_df,
    aes(x = x, y = y, label = label),
    colour      = "firebrick",
    size        = 3.5,
    bg.color    = "white",
    bg.r        = 0.15,
    inherit.aes = FALSE
  )

# --- Bottom right: Channel plot ---
p4 <- fviz_pca_ind(
  pca_result,
  geom.ind     = "point",
  col.ind      = factor(data$Channel),
  palette      = c("#E64B35", "#4DBBD5"),
  addEllipses  = TRUE,
  legend.title = "Channel",
  title        = "(d) PCA - Individuals Coloured by Channel (1=HoReCa, 2=Retail)"
)

# BE note: qol to get bottom two plots to align vertically
# ggplot_build(p3)$layout$panel_params[[1]]$y.range
# ggplot_build(p4)$layout$panel_params[[1]]$y.range

p3 <- p3 + coord_cartesian(ylim = c(-5.58, 5.24))
p4 <- p4 + coord_cartesian(ylim = c(-5.58, 5.24))

# --- Combine 2x2 ---
(p1 | p2) / (p3 | p4)
```


<!-- ```{=latex} -->
<!-- \clearpage -->
<!-- \newpage -->
<!-- ``` -->

# Customer clustering {#qtwo}

```{r q2-prep, echo=FALSE, include=FALSE}
# ==============================================================================
# Q2 Clustering
# ==============================================================================
spending.cols <- c("Fresh", "Milk", "Grocery", "Frozen", "Detergents_Paper", "Delicassen")

# K-Means for k=2 through 5
KMeans.results <- list()
for (k in 2:5) {
  KMeans.results[[k]] <- kmeans(scaled_data, centers = k, nstart = 25)
}

# Assign K-Means labels programmatically
km3 <- KMeans.results[[3]]
km3_ch_tab <- table(km3$cluster, data$Channel)
horeca_pct <- km3_ch_tab[, 1] / rowSums(km3_ch_tab)

horeca_cluster <- as.character(which.max(horeca_pct))
retail_cluster <- as.character(which.min(horeca_pct))
mixed_cluster  <- as.character(setdiff(1:3, c(horeca_cluster, retail_cluster)))

km3_names <- setNames(
  c("HoReCa Dominant", "Retail Dominant", "Mixed Basket"),
  c(horeca_cluster, retail_cluster, mixed_cluster)
)
km3_colours <- setNames(
  c("#B2182B", "#2166AC", "#FDB863"),
  c(horeca_cluster, retail_cluster, mixed_cluster)
)
km3_order <- c(retail_cluster, mixed_cluster, horeca_cluster)

# silhouette scores (referenced inline)
sil_scores <- sapply(2:10, function(k) {
  km <- kmeans(scaled_data, centers = k, nstart = 25)
  sil <- cluster::silhouette(km$cluster, dist(scaled_data))
  mean(sil[, 3])
})
names(sil_scores) <- 2:10

# GMM on spending data (referenced inline for $G, $modelName, $bic)
gmm_spending <- Mclust(data = scaled_data)

# BIC comparison table (referenced inline)
bic_vals    <- gmm_spending$BIC
best_per_k  <- apply(bic_vals, 1, max, na.rm = TRUE)

# Ward's hierarchical clustering (referenced in main text)
dist_mat    <- dist(scaled_data, method = "euclidean")
hc          <- hclust(dist_mat, method = "ward.D2")
hc_clusters <- cutree(hc, k = 3)

# Shorthand references used in main text ====
km3 <- KMeans.results[[3]]
km2 <- KMeans.results[[2]]
data$km3_cluster <- km3$cluster
data$km2_cluster <- km2$cluster
data$gmm_cluster <- as.integer(gmm_spending$classification)

# Median basket % per cluster (K-Means)
km3_medians <- aggregate(
  cbind(Fresh, Milk, Grocery, Frozen, Detergents_Paper, Delicassen) ~ km3_cluster,
  data = data, FUN = median
)
km3_medians$total <- rowSums(km3_medians[, spending.cols])
for (col in spending.cols) {
  km3_medians[[paste0(col, "_pct")]] <- round(km3_medians[[col]] / km3_medians$total * 100, 1)
}

# Revenue concentration % per cluster (K-Means)
km3_sums_all <- aggregate(
  cbind(Fresh, Milk, Grocery, Frozen, Detergents_Paper, Delicassen) ~ km3_cluster,
  data = data, FUN = sum
)
for (col in spending.cols) {
  km3_sums_all[[paste0(col, "_rev_pct")]] <- round(km3_sums_all[[col]] / sum(km3_sums_all[[col]]) * 100, 1)
}

# Same for GMM
gmm_medians <- aggregate(
  cbind(Fresh, Milk, Grocery, Frozen, Detergents_Paper, Delicassen) ~ gmm_cluster,
  data = data, FUN = median
)
gmm_medians$total <- rowSums(gmm_medians[, spending.cols])
for (col in spending.cols) {
  gmm_medians[[paste0(col, "_pct")]] <- round(gmm_medians[[col]] / gmm_medians$total * 100, 1)
}

gmm_sums_all <- aggregate(
  cbind(Fresh, Milk, Grocery, Frozen, Detergents_Paper, Delicassen) ~ gmm_cluster,
  data = data, FUN = sum
)
for (col in spending.cols) {
  gmm_sums_all[[paste0(col, "_rev_pct")]] <- round(gmm_sums_all[[col]] / sum(gmm_sums_all[[col]]) * 100, 1)
}

# Helper: look up a value by cluster ID
km3_stat <- function(cluster_id, col) {
  row <- km3_medians$km3_cluster == as.integer(cluster_id)
  km3_medians[row, col]
}
km3_rev <- function(cluster_id, col) {
  row <- km3_sums_all$km3_cluster == as.integer(cluster_id)
  km3_sums_all[row, col]
}
gmm_stat <- function(cluster_id, col) {
  row <- gmm_medians$gmm_cluster == cluster_id
  gmm_medians[row, col]
}
gmm_rev <- function(cluster_id, col) {
  row <- gmm_sums_all$gmm_cluster == cluster_id
  gmm_sums_all[row, col]
}
gmm_n <- function(cluster_id) {
  sum(data$gmm_cluster == cluster_id)
}
gmm_pct_customers <- function(cluster_id) {
  round(sum(data$gmm_cluster == cluster_id) / nrow(data) * 100)
}
```


```{r q2-revenue-triptych, echo=FALSE, fig.width=12, fig.height=10, fig.cap="Revenue concentration by product category under three segmentation approaches. Top: the distributor's existing binary Channel classification. Middle: K-Means (k=3) spending-based clusters. Bottom: Gaussian Mixture Modelling (GMM) (k=5) spending-based clusters. Crucially, the K-Means and GMM segmentations were derived solely from log-transformed, standardised spending data without access to Channel labels. Each bar shows the percentage of total category spend attributable to each segment and the Total column shows each segment's share of overall revenue."}

n_total <- nrow(data)
spend_cols <- c(
  "Fresh",
  "Milk",
  "Grocery",
  "Frozen",
  "Detergents_Paper",
  "Delicassen"
)
cat_levels <- c(spend_cols, "Total")

# ==============================================================================
# Helper: from a grouping vector, produce a revenue-concentration data frame
# ==============================================================================
build_shares <- function(group_vec, group_name = "cluster") {
  
  # Sums per group
  sums <- aggregate(
    cbind(Fresh, Milk, Grocery, Frozen, Detergents_Paper, Delicassen) ~ grp,
    data = cbind(data[, spend_cols], grp = group_vec),
    FUN  = sum
  )
  
  # Column percentages
  pcts <- sums
  for (col in spend_cols) {
    pcts[[col]] <- sums[[col]] / sum(sums[[col]]) * 100
  }
  
  # Pivot long
  pcts_long <- reshape(
    pcts, direction = "long",
    varying = spend_cols, v.names = "Pct",
    timevar = "Category", times = spend_cols
  )
  pcts_long$Cluster <- factor(pcts_long$grp)
  row.names(pcts_long) <- NULL
  
  # Total column
  sums$total <- sums$Fresh + sums$Milk + sums$Grocery +
                sums$Frozen + sums$Detergents_Paper + sums$Delicassen
  totals <- data.frame(
    grp      = sums$grp,
    Category = "Total",
    Pct      = sums$total / sum(sums$total) * 100,
    Cluster  = factor(sums$grp)
  )
  
  # Combine
  out <- rbind(pcts_long[, c("grp", "Category", "Pct", "Cluster")], totals)
  out$Category <- factor(out$Category, levels = cat_levels)
  out
}

# ==============================================================================
# Helper: from a grouping vector, produce legend labels "Label-X (n=Y, Z%)"
# ==============================================================================
build_labels <- function(group_vec, names_dict = NULL, prefix = "Cluster") {
  counts <- as.data.frame(table(group_vec))
  colnames(counts) <- c("grp", "n")
  
  # Calculate percentages
  counts$pct <- (counts$n / n_total) * 100
  
  # Construct the label string
  if (!is.null(names_dict)) {
    # If a dictionary is provided, map the group number to your custom name
    counts$desc <- names_dict[as.character(counts$grp)]
    counts$label <- sprintf("%s (n=%d, %.0f%%)", counts$desc, counts$n, counts$pct)
  } else {
    # Fallback to the original Prefix-X format if no dictionary is supplied
    counts$label <- sprintf("%s-%s (n=%d, %.0f%%)", prefix, counts$grp, counts$n, counts$pct)
  }
  
  # Return a named vector required by scale_fill_manual / scale_fill_brewer
  setNames(counts$label, as.character(counts$grp))
}

# ==============================================================================
# Helper: standard revenue concentration plot
# ==============================================================================
plot_shares <- function(share_data, fill_labels, title, palette = "Set1",
                        manual_colours = NULL) {
  p <- ggplot(share_data, aes(x = Category, y = Pct, fill = Cluster)) +
    geom_col(position = "stack", width = 0.7) +
    geom_text(
      aes(label = sprintf("%.0f%%", Pct)),
      position = position_stack(vjust = 0.5),
      size = 3, colour = "white"
    ) +
    geom_vline(xintercept = 6.5, linetype = "dashed", colour = "grey40") +
    labs(title = title, y = "Share of total category spend (%)",
         x = "", fill = "Segment") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
  
  if (!is.null(manual_colours)) {
    p <- p + scale_fill_manual(values = manual_colours, labels = fill_labels)
  } else {
    p <- p + scale_fill_brewer(palette = palette, labels = fill_labels)
  }
  p
}

# ==============================================================================
# Build all three plots
# ==============================================================================

# Define the descriptive segment names based on the spending profiles
channel_names <- c(
  "1" = "HoReCa",
  "2" = "Retail"
)

# km3_names <- c(
#   "1" = "Mixed Basket",
#   "2" = "HoReCa Dominant",
#   "3" = "Retail Dominant"
# )

gmm_names <- c(
  "1" = "Heavy Retail",
  "2" = "Diversified Retail",
  "3" = "High-volume HoReCa",
  "4" = "Balanced HoReCa",
  "5" = "Low-spend / Mixed"
)

# factor ordering: HoReCa-type first (bottom), Retail-type last (top)
channel_order <- c("2", "1") 
# HoReCa, Retail

# km3_order <- c("3", "1", "2") 
# # HoReCa Dominant, Mixed, Retail Dominant

gmm_order <- c("1", "2", "5", "4", "3") 
# High-vol HoReCa, Balanced HoReCa, Low-spend, Diversified Retail, Heavy Retail

gmm_colours <- c(
  "3" = "#B2182B",   # deep red - high-vol HoReCa
  "4" = "#E66101",   # orange - balanced HoReCa
  "5" = "#FDB863",   # amber - low-spend mixed
  "2" = "#92C5DE",   # light blue - diversified Retail
  "1" = "#2166AC"    # deep blue - heavy Retail
)
# km3_colours <- c(
#   "2" = "#B2182B",   # HoReCa Dominant
#   "1" = "#FDB863",   # Mixed Basket
#   "3" = "#2166AC"    # Retail Dominant
# )
ch_colours <- c("1" = "#E64B35", "2" = "#4DBBD5")

palettes <- list(
  rdylbu = list(
    gmm = c("3"="#B2182B", "4"="#E66101", "5"="#FDB863", "2"="#92C5DE", "1"="#2166AC"),
    km3 = c("2"="#B2182B", "1"="#FDB863", "3"="#2166AC"),
    ch  = c("1"="#E64B35", "2"="#4DBBD5")
  ),
  okabe_ito = list(
    gmm = c("3"="#D55E00", "4"="#E69F00", "5"="#F0E442", "2"="#56B4E9", "1"="#0072B2"),
    km3 = c("2"="#D55E00", "1"="#F0E442", "3"="#0072B2"),
    ch  = c("1"="#D55E00", "2"="#0072B2")
  ),
  viridis = list(
    gmm = c("3"="#440154", "4"="#3B528B", "5"="#21918C", "2"="#5EC962", "1"="#FDE725"),
    km3 = c("2"="#440154", "1"="#21918C", "3"="#FDE725"),
    ch  = c("1"="#440154", "2"="#FDE725")
  )
)

active <- palettes$rdylbu

gmm_colours <- active$gmm
km3_colours <- active$km3
ch_colours  <- active$ch

# (a) Channel baseline
ch_vec    <- as.integer(data$Channel)
ch_shares <- build_shares(ch_vec)
ch_shares$Cluster  <- factor(ch_shares$Cluster,  levels = channel_order)
ch_labels <- build_labels(ch_vec, names_dict = channel_names)

p_channel <- plot_shares(
  ch_shares,
  ch_labels,
  title = "(a) Baseline: Channel",
  manual_colours = ch_colours
) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank()
  )

# (b) K-Means k=3
km3_vec    <- as.integer(KMeans.results[[3]]$cluster)
km3_shares <- build_shares(km3_vec)
km3_shares$Cluster <- factor(km3_shares$Cluster, levels = km3_order)
km3_labels <- build_labels(km3_vec, names_dict = km3_names)

p_km3 <- plot_shares(
  km3_shares,
  km3_labels,
  title = "(b) K-Means (k = 3)",
  manual_colours = km3_colours
) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank()
  )

# (c) Gaussian Mixture Modelling k=5
gmm_vec    <- as.integer(gmm_spending$classification)
gmm_shares <- build_shares(gmm_vec)
gmm_shares$Cluster <- factor(gmm_shares$Cluster, levels = gmm_order)
gmm_labels <- build_labels(gmm_vec, names_dict = gmm_names)

p_gmm <- plot_shares(
  gmm_shares, gmm_labels,
  title = "(c) Gaussian Mixture Modelling (k = 5)", 
  manual_colours = gmm_colours
)

# ==============================================================================
# Stack
# ==============================================================================
p_channel / p_km3 / p_gmm + plot_layout(heights = c(1, 1, 1.15))
```


K-Means clustering was applied to the same log-transformed, standardised
spending data used for PCA. Silhouette analysis identifies k=2 as optimal 
(`r formatC(sil_scores["2"], format = "f", digits = 3)`), which recovers the known
Channel distinction and validates the clustering approach. However, the
silhouette score for k=3 remains strong 
(`r formatC(sil_scores["3"], format = "f", digits = 3)`), and the marginal 
reduction in cluster quality yields a substantially richer segmentation. Beyond
k=3, scores plateau around 0.19 with no meaningful recovery (Appendix, Figure
\@ref(fig:silhouette-plot)), suggesting additional clusters fragment genuine
structure rather than reveal new segments.

K-Means was chosen for its interpretability and computational efficiency on this
dataset. As robustness checks, both Ward's hierarchical clustering and Gaussian
Mixture Modelling (GMM) were applied to the same input. Ward's method (k=3)
agrees with K-Means on the dominant cluster structure, with disagreement
confined to boundary observations between the smaller segments (Appendix
\@ref(apd:wards-clust)). GMM relaxes K-Means' spherical cluster assumption: the
BIC selected a `r gmm_spending$G`-component `r gmm_spending$modelName` model
(Appendix, Figure \@ref(fig:bic-plot)), splitting both channels into finer
sub-segments. Cross-tabulation against Channel and Region (Appendix, Tables
\@ref(tab:km3-channel-crosstab) & \@ref(tab:gmm-region-crosstab)) confirms that
these sub-segments are spending-driven rather than geographic, with Region
distributed roughly proportionally within each cluster.

Figure \@ref(fig:q2-revenue-triptych) presents the commercial impact of each
segmentation. Under the existing binary Channel label, HoReCa customers account
for 76% of Fresh revenue but only 68% of customers. 
K-Means (k=3) identifies a "`r km3_names[mixed_cluster]`" segment 
(`r sum(km3$cluster == as.integer(mixed_cluster))` customers, 
`r round(sum(km3$cluster == as.integer(mixed_cluster)) / nrow(data) * 100)`%) 
with spending patterns that cut across both channels.
GMM further distinguishes Heavy Retail from Diversified Retail, and separates
High-volume HoReCa from Balanced HoReCa, each with distinct category
concentration profiles.

The strength of this approach is that it operates entirely without access to
Channel or Region labels. The near-perfect recovery of the Channel distinction
validates the segmentation as behaviourally grounded, while the emergence of
additional clusters reveals sub-structure the existing classification does not
capture. However, the commercial value of a 5-segment model over a 3-segment
model cannot be determined from a single cross-sectional snapshot. Validation
would require longitudinal purchasing data, margin information, and ideally a
controlled trial of differentiated service strategies across segments.


# Repeating \@ref(qone) & \@ref(qtwo) using LLMs {#qthr}

Two LLMs were evaluated: (LLM1) and (LLM2). To ensure comparability, identical
prompts were used for both models and neither was given the raw `csv` data file.
Instead, the dataset was described in the first prompt and the LLMs were asked
to produce R code for independent execution. This design choice was deliberate.
By withholding the data, the procedure replicated a commercial setting in which
an organisation may not wish to upload private, potentially valuable data to an
LLM provider. It also ensured that neither model could adapt its recommendations
to the observed distributions, testing whether each model's default
methodological choices were appropriate for this dataset. Approximately one hour
was spent per model across four prompts, with the final prompt kept consistent
in intent but adapted to each model's prior output. Full prompts, responses, and
figures produced by each LLM's generated code are provided in Appendix
\@ref(apd-llm-information).

- Want to mention that the last comment was slightly dynamic and fed the input back into the LLM so slightly different, but kept the command under that consistent.

- Using same prompts for 2 LLMs for consistency / standardiseation.
- 'to ensure compatibility identical prompts were used for both models' 
(described in full in the appendix \@ref(apd-llm-information))

Example table:

```{r llm-comparison, echo=FALSE}
# define symbols
sym <- list(
  Y = "$\\checkmark$",    # ● addressed unprompted
  P = "$\\circ$",         # ○ partial / after prompting
  N = "\\texttimes"       # ✗ missed entirely
)

sym_legend <- paste0(
  sym$Y, "~= addressed unprompted; ",
  sym$P, "~= partial; ",
  sym$N, "~= missed entirely."
)

# data structure 
# > can replace "?" with Y/P/N after running the LLM experiment

rows <- tribble(
  ~group,            ~label,                                          ~ours, ~chatgpt, ~claude,
  "Preprocessing",   "Excluded Channel \\& Region from PCA/clustering", "Y",  "?",      "?",
  "Preprocessing",   "Identified skewness / recommended log transform",  "Y",  "?",      "?",
  "Preprocessing",   "Standardised variables before PCA",               "Y",  "?",      "?",
  "PCA",             "Produced interpretable scree plot / variance",     "Y",  "?",      "?",
  "PCA",             "Interpreted PC1 as channel axis",                  "Y",  "?",      "?",
  "PCA",             "Validated PCA against Channel label",              "Y",  "?",      "?",
  "Clustering",      "Justified choice of clustering method",            "Y",  "?",      "?",
  "Clustering",      "Used formal $k$ selection (silhouette / BIC)",     "Y",  "?",      "?",
  "Clustering",      "Identified structure beyond $k{=}2$",              "Y",  "?",      "?",
  "Clustering",      "Suggested robustness check (alternative method)",  "Y",  "?",      "?",
  "Interpretation",  "Commercial framing of segments",                   "Y",  "?",      "?",
  "Interpretation",  "Acknowledged limitations of the analysis",         "Y",  "?",      "?",
  "Interpretation",  "Suggested further methods or extensions",          "Y",  "?",      "?"
)

# map codes to symbols
map_sym <- function(x) {
  dplyr::case_match(x,
    "Y" ~ sym$Y,
    "P" ~ sym$P,
    "N" ~ sym$N,
    .default = x
  )
}

# compute totals pre symbol mapping
score_col <- function(col) {
  n_total <- sum(col %in% c("Y", "P", "N"))
  if (n_total == 0) return("---")
  n_y <- sum(col == "Y")
  n_p <- sum(col == "P")
  # format: "10/14 (2P)" or "14/14" if no partials
  if (n_p > 0) {
    paste0(n_y, "/", n_total, " (", n_p, "P)")
  } else {
    paste0(n_y, "/", n_total)
  }
}

totals <- tibble(
  group   = "Totals",
  label   = "\\textbf{Score}",
  ours    = score_col(rows$ours),
  chatgpt = score_col(rows$chatgpt),
  claude  = score_col(rows$claude)
)

# combine rows + totals, then map symbols
all_rows <- bind_rows(rows, totals) %>%
  mutate(across(c(ours, chatgpt, claude), map_sym))

#compute pack_rows indices from group column
group_indices <- all_rows %>%
  mutate(row = row_number()) %>%
  group_by(group) %>%
  summarise(start = min(row), end = max(row), .groups = "drop")

group_indices <- group_indices %>%
  mutate(group = factor(group, levels = unique(all_rows$group))) %>%
  arrange(group)

# build kable
kable_data <- all_rows %>%
  dplyr::select(label, ours, chatgpt, claude)

colnames(kable_data) <- c("Analysis Step", "Our Analysis", "ChatGPT (4o)", "Claude")

caption_text <- paste0(
  "Comparison of analytical steps: our pipeline vs \\ two LLMs. ",
  sym_legend
)

tbl <- kable(
  kable_data,
  booktabs  = TRUE,
  escape    = FALSE,
  align     = c("l", "c", "c", "c"),
  caption   = caption_text,
  linesep   = ""
) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size     = 10
  )

# -apply pack_rows 
for (i in seq_len(nrow(group_indices))) {
  grp <- as.character(group_indices$group[i])
  # skip the totals row - it stands alone with a rule above it
  if (grp == "Totals") next
  tbl <- tbl %>%
    pack_rows(
      grp,
      group_indices$start[i],
      group_indices$end[i],
      bold         = TRUE,
      italic       = FALSE,
      hline_before = (i > 1)
    )
}

# if (knitr::is_latex_output()) {
#   # cat("\\begingroup\n")
#   cat("\\renewcommand{\\arraystretch}{1.3}\n")
# }

tbl 

# if (knitr::is_latex_output()) {
#   # cat("\\endgroup\n")
# }
# # tbl
```


```{=latex}
\clearpage
\newpage
```

<!-- ```{=html} -->
<!-- # References -->
<!-- <div id="refs"></div> -->
<!-- ``` -->

<!-- ```{=latex} -->
<!-- \bibliography{references.bib} -->
<!-- ``` -->

```{=latex}
\newpage
\section*{Appendices}
\addcontentsline{toc}{section}{Appendices}

% --- restore default section numbering in the appendix ---
\renewcommand{\thesection}{\arabic{section}}
\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
\renewcommand{\thesubsubsection}{\thesubsection.\arabic{subsubsection}}
```

# (APPENDIX) Appendices {-}

# Supplementary methodology {#apd:suppl-methods}

## Raw data distribution, correlations, and skewness {#apd:suppl-skew}

```{r boxplots-raw-data, echo=knitr::is_html_output(), include=TRUE, fig.width=12, fig.height=5.5, fig.cap="Customer spending profiles across six wholesale categories. Top: boxplots showing the highly skewed distribution of individual annual spend. Bottom: the total aggregate volume, highlighting Fresh and Grocery as the largest product segments. Both plots are sorted by the  median spend."}
box_data <- reshape(
  data[, spending.cols],
  direction = "long",
  varying = spending.cols,
  v.names = "Spend",
  timevar = "Category",
  times = spending.cols
)

cat_totals <- data.frame(
  Category = spending.cols,
  Total    = colSums(data[, spending.cols]),
  Median   = apply(data[, spending.cols], 2, median)
)

# option to sort by median vs total; 
# BE note: have chosen median here to highlight the difference in median spend
# despite large total gap between bottom two categories
cat_order <- levels(reorder(cat_totals$Category, cat_totals$Median))
# cat_order <- levels(reorder(cat_totals$Category, cat_totals$Total))
box_data$Category <- factor(box_data$Category, levels = cat_order)
cat_totals$Category <- factor(cat_totals$Category, levels = cat_order)
row.names(box_data) <- NULL


p_box <- ggplot(
  box_data,
  aes(x = Category, y = Spend)
) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::comma,
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "(a) Spending Distribution per Customer",
    y = "Annual Spend (m.u.)",
    x = ""
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

p_bar <- ggplot(cat_totals, aes(x = Category, y = Total)) +
  geom_col(fill = "steelblue", alpha = 0.7, width = 0.7) +
  geom_text(
    aes(label = format(round(Total), big.mark = ",")),
    hjust = -0.1,
    size = 3.5
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "(b) Total Category Spend",
    y = "Total Spend (m.u.)",
    x = ""
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

p_box / p_bar
```

Figure \@ref(fig:boxplots-raw-data) provides a baseline overview of purchasing
behaviour across the six categories. Median spend is relatively modest in all
categories, but every distribution exhibits a long right tail of high-value
customers. At the aggregate level, `r format_var_name("Fresh")` and 
`r format_var_name("Grocery")` dominate total revenue, while 
`r format_var_name("Delicassen")` remains a niche segment. 

```{r histograms-overlay, echo=FALSE, include=TRUE, fig.width=10, fig.height=6, fig.cap="Shape comparison of raw and log-transformed spending distributions by product category. To facilitate direct comparison of the distribution shapes, all spend values were min-max scaled to a 0–1 range within their respective categories. The raw data (blue) demonstrates a right skew across all categories, influenced by extreme outliers. The `log1p` transformation (red) effectively compresses this long tail, yielding more symmetric distributions that ensure variance-based methods (PCA) and distance-based methods (clustering) are not dominated by extreme outliers."}

bind_rows(raw_df, log_df) %>%
  mutate(
    Transform = factor(Transform, levels = c("Raw", "Log-transformed"))
  ) %>%
  group_by(Category, Transform) %>%
  mutate(Spend_scaled = rescale01(Spend)) %>%
  ungroup() %>%
  ggplot(aes(x = Spend_scaled, fill = Transform)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 40,
    colour = "white",
    alpha = 0.5,
    position = "identity"
  ) +
  scale_fill_manual(
    values = c("Raw" = "steelblue", "Log-transformed" = "firebrick")
  ) +
  facet_wrap(~Category, scales = "free_y") +
  labs(
    title = "Shape Comparison: Raw vs Log-transformed Spending",
    x = "Normalised Spend (0-1 scaled within each group)",
    y = "Density"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

This skewness is examined more closely in Figure \@ref(fig:histograms-overlay),
which overlays raw and log-transformed distributions for each category. The raw
spending figures exhibit a strong rightward skew across all six product
categories, indicating that while most customers spend relatively small amounts,
a long tail of high-spending customers dominates the scale. Applying a log
transformation (`log1p`) successfully 'pulls in' these extreme values, centring
the bulk of the data and producing much more symmetric distributions.


```{r corrplots-side-by-side, echo=knitr::is_html_output(), include=TRUE, fig.width=12, fig.height=5.5, fig.cap="Pairwise correlation matrices of the six product categories, comparing linear relationships using raw spending data (left) versus log-transformed spending data (right)."}
par(mfrow = c(1, 2))

soft_palette <- colorRampPalette(c("#D73027", "#F46D43", "#FFFFFF", "#74ADD1", "#4575B4"))(200)

corrplot(
  cor(spending),
  method      = "color",
  type        = "upper",
  col         = soft_palette,     
  addCoef.col = "black",          
  number.cex  = 0.8,
  tl.col      = "black",
  title       = "Raw spending",
  mar         = c(0, 0, 2, 0)
)

corrplot(
  cor(spending_log),
  method      = "color",
  type        = "upper",
  col         = soft_palette, 
  addCoef.col = "black",        
  number.cex  = 0.8,
  tl.col      = "black",
  title       = "Log-transformed spending",
  mar         = c(0, 0, 2, 0)
)
```

The side-by-side correlation plots reveal the underlying purchasing
relationships between product categories before and after adjusting for scale.
Both matrices highlight a distinct cluster of highly correlated spending:
`r format_var_name("Grocery")`, `r format_var_name("Detergents_Paper")`, and
`r format_var_name("Milk")` show strong positive linear relationships,
suggesting they are frequently purchased together in bulk. While the log
transformation preserves this core structure, it slightly softens the most
extreme correlations (for instance, the correlation between
`r format_var_name("Grocery")` and `r format_var_name("Detergents_Paper")`
drops from `r formatC(cor(spending$Grocery, spending$Detergents_Paper), format = "f", digits = 2)`
to `r formatC(cor(spending_log$Grocery, spending_log$Detergents_Paper), format = "f", digits = 2)`).
This indicates that in the raw data, a few extreme outlier purchases were
artificially inflating the strength of that linear relationship.

## Multi-dimensional scaling {#apd:suppl-mds}

```{r mds-side-by-side, echo=FALSE, include=TRUE, fig.width=12, fig.height=5.5, fig.cap="Non-metric MDS using Gower’s distance across all eight variables (six log-transformed continuous spending categories plus Channel and Region). Points are coloured by Channel (left) and Region (right). Unlike PCA, Gower’s distance forces distinct spatial separation based on categorical mismatches, resulting in six artificial clusters corresponding to the cross-tabulation of the two Channels and three Regions."}

# Gower distance on all 8 variables
mds_data <- data.frame(
  Channel = factor(data$Channel, labels = c("HoReCa", "Retail")),
  Region = factor(data$Region, labels = c("Lisbon", "Oporto", "Other")),
  spending_log
)

gow <- cluster::daisy(mds_data, metric = "gower")
mds_fit <- MASS::isoMDS(gow, k = 2, trace = FALSE)
mds_pts <- as.data.frame(mds_fit$points)
colnames(mds_pts) <- c("MDS1", "MDS2")

mds_pts$Channel <- mds_data$Channel
mds_pts$Region <- mds_data$Region

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Left: coloured by Channel
plot(
  mds_pts$MDS1,
  mds_pts$MDS2,
  col = ifelse(mds_pts$Channel == "HoReCa", "#E64B35", "#000000"),
  # 19 = Circle, 17 = Triangle
  pch = ifelse(mds_pts$Channel == "HoReCa", 19, 17),
  cex = 0.8,
  xlab = "MDS1",
  ylab = "MDS2",
  main = "(a) Coloured by Channel"
)
legend(
  "topleft",
  legend = c("HoReCa", "Retail"),
  col = c("#E64B35", "#000000"),
  pch = c(19, 17),
  cex = 0.9,
  bty = "n"
)

# Right: coloured by Region
region_cols <- c(Lisbon = "#E69F00", Oporto = "#56B4E9", Other = "#CC79A7")
plot(
  mds_pts$MDS1,
  mds_pts$MDS2,
  col = region_cols[mds_pts$Region],
  pch = 19,
  cex = 0.8,
  xlab = "MDS1",
  ylab = "MDS2",
  main = "(b) Coloured by Region"
)
legend(
  "topleft",
  legend = names(region_cols),
  col = region_cols,
  pch = 19,
  cex = 0.9,
  bty = "n"
)
# mds_fit$stress
```

Figure \@ref(fig:mds-side-by-side) presents a non-metric MDS solution computed
from Gower's distance across all eight variables, incorporating the categorical
`r format_var_name("Channel")` and `r format_var_name("Region")` columns excluded
from the PCA. Because Gower's metric treats any categorical mismatch as a
maximal distance contribution on that variable, the resulting spatial map is
dominated by these labels rather than underlying spending behaviour.

Panel (a) shows separation by `r format_var_name("Channel")` along the primary
axis (MDS1), while Panel (b) reveals further stratification by 
`r format_var_name("Region")` along MDS2. The result is a forced stratification
into six spatial clusters reflecting the Channel $\times$ Region
cross-tabulation. While there is minor regional intermingling (e.g., two Oporto
customers mapping near the Lisbon cluster), the overall spatial map merely
reflects the cross-tabulation of the categorical labels rather than organic,
dataset-wide purchasing patterns. The solution's stress value of
`r formatC(mds_fit$stress, format = "fg", digits = 3)` further indicates a 
mediocre fit, falling above the conventional threshold of 10 for reliable
low-dimensional representation. This confirms that the PCA approach, relying
purely on continuous spending variables, is better suited to uncovering natural
commercial structure in this dataset.

```{=latex}
\newpage
```

## Clustering diagnostics {#apd:clustering-diagnostics}

### Optimal cluster selection {#apd:cluster-selection}
```{r silhouette-plot, echo=FALSE, fig.width=10, fig.height=4, fig.cap="Silhouette analysis for K-Means clustering (k=2 to 10). The optimal number of clusters is k=2 followed by k=3 as a strong second. Beyond k=3, scores plateau around 0.19."}
fviz_nbclust(scaled_data, kmeans, method = 'silhouette', nstart = 25)
```

```{r bic-plot, echo=FALSE, fig.width=10, fig.height=5, fig.cap="BIC values across component counts for all model types evaluated by `mclust`. Higher (less negative) BIC indicates better fit. The BIC peaks at 5 components under the VVE (ellipsoidal, variable volume and shape, equal orientation) model."}
plot(
  gmm_spending,
  what = "BIC",
  legendArgs = list(x = "bottomright", ncol = 4, cex = 0.8)
)
```

Silhouette analysis (Figure \@ref(fig:silhouette-plot)) identifies k=2 as
optimal (`r formatC(sil_scores["2"], format = "f", digits = 3)`), with k=3 as a
close second (`r formatC(sil_scores["3"], format = "f", digits = 3)`). Beyond
k=3, scores plateau around 0.19.

The GMM's BIC (Figure \@ref(fig:bic-plot)) peaks at
`r gmm_spending$G` components
(`r formatC(gmm_spending$bic, format = "f", digits = 1, big.mark = ",")`), 
compared to
`r formatC(best_per_k["4"], format = "f", digits = 1, big.mark = ",")` 
for four and
`r formatC(best_per_k["6"], format = "f", digits = 1, big.mark = ",")` for six. 
Interestingly k=4 is worse than k=3, suggesting the data supports either 3 or 5
components but not an intermediate value. Both methods agree that k=3 is a
meaningful threshold and the disagreement on whether to proceed to k=5 reflects 
a genuine trade-off between parsimony and granularity.


### Channel and Region cross-tabulations {#apd:cluster-crosstabs}

```{r km3-channel-crosstab, echo=FALSE}
km3_3way <- table(
  KM = km3$cluster,
  Channel = data$Channel,
  Region = data$Region
)

km_rows <- list()
for (k in 1:3) {
  for (ch in 1:2) {
    km_rows[[length(km_rows) + 1]] <- data.frame(
      KM = k,
      Channel = ch,
      Lisbon = km3_3way[k, ch, 1],
      Oporto = km3_3way[k, ch, 2],
      Other = km3_3way[k, ch, 3]
    )
  }
}
km_region_df <- do.call(rbind, km_rows)

# Build pack_rows labels programmatically
km3_pack_labels <- sapply(1:3, function(k) {
  paste0("KM-", k, " (", km3_names[as.character(k)], ")")
})

kable(
  km_region_df[, c("Channel", "Lisbon", "Oporto", "Other")],
  booktabs = TRUE,
  row.names = FALSE,
  caption = "K-Means (k=3) cluster assignments vs \\ Channel (1=HoReCa, 2=Retail) and Region.",
  align = c("l", "r", "r", "r")
) %>%
  kable_styling(latex_options = "hold_position", font_size = 10) %>%
  pack_rows(km3_pack_labels[1], 1, 2, bold = TRUE, hline_before = FALSE) %>%
  pack_rows(km3_pack_labels[2], 3, 4, bold = TRUE, hline_before = TRUE) %>%
  pack_rows(km3_pack_labels[3], 5, 6, bold = TRUE, hline_before = TRUE)
```

```{r gmm-region-crosstab, echo=FALSE}
gmm_3way <- table(
  GMM = data$gmm_cluster,
  Channel = data$Channel,
  Region = data$Region
)

rows_list <- list()
for (g in 1:5) {
  for (ch in 1:2) {
    rows_list[[length(rows_list) + 1]] <- data.frame(
      GMM = g,
      Channel = ch,
      Lisbon = gmm_3way[g, ch, 1],
      Oporto = gmm_3way[g, ch, 2],
      Other = gmm_3way[g, ch, 3]
    )
  }
}
region_df <- do.call(rbind, rows_list)

gmm_pack_labels <- sapply(1:5, function(g) {
  paste0("GMM-", g, " (", gmm_names[as.character(g)], ")")
})

kable(
  region_df[, c("Channel", "Lisbon", "Oporto", "Other")],
  booktabs = TRUE,
  row.names = FALSE,
  caption = "GMM (k=5) cluster assignments vs \\ Channel (1=HoReCa, 2=Retail) and Region.",
  align = c("l", "r", "r", "r")
) %>%
  kable_styling(latex_options = "hold_position", font_size = 10) %>%
  pack_rows(gmm_pack_labels[1], 1, 2, bold = TRUE, hline_before = FALSE) %>%
  pack_rows(gmm_pack_labels[2], 3, 4, bold = TRUE, hline_before = TRUE) %>%
  pack_rows(gmm_pack_labels[3], 5, 6, bold = TRUE, hline_before = TRUE) %>%
  pack_rows(gmm_pack_labels[4], 7, 8, bold = TRUE, hline_before = TRUE) %>%
  pack_rows(gmm_pack_labels[5], 9, 10, bold = TRUE, hline_before = TRUE)
```


Table \@ref(tab:km3-channel-crosstab) shows that K-Means (k=3) produces one
cluster that is almost exclusively HoReCa 
(KM-`r horeca_cluster`, "`r km3_names[horeca_cluster]`": 
`r km3_ch_tab[as.integer(horeca_cluster), 1]` of 
`r sum(km3$cluster == as.integer(horeca_cluster))` customers), one that leans 
Retail (KM-`r retail_cluster`, "`r km3_names[retail_cluster]`": 
`r km3_ch_tab[as.integer(retail_cluster), 2]` of 
`r sum(km3$cluster == as.integer(retail_cluster))`), and a third 
(KM-`r mixed_cluster`, "`r km3_names[mixed_cluster]`") that draws substantially
from both channels 
(`r km3_ch_tab[as.integer(mixed_cluster), 1]` HoReCa, 
`r km3_ch_tab[as.integer(mixed_cluster), 2]` Retail). The "Retail Dominant" 
label reflects the spending profile rather than exclusive Channel membership: 
these customers concentrate on `r format_var_name("Grocery")` and 
`r format_var_name("Detergents_Paper")` regardless of their administrative 
Channel label. 

```{=latex}
\newpage
```

### Cluster spending profiles {#apd:cluster-profiles}

```{r q2-tables-3-apd, echo=FALSE}
# Build from precomputed km3_medians and gmm_medians
km3_raw_table <- km3_medians[, c("km3_cluster", spending.cols)]
km3_raw_table$n <- sapply(km3_raw_table$km3_cluster, function(k) {
  sum(data$km3_cluster == k)
})
km3_raw_table$Cluster <- paste0("KM-", km3_raw_table$km3_cluster)
km3_raw_table$Label <- km3_names[as.character(km3_raw_table$km3_cluster)]

gmm_raw_table <- gmm_medians[, c("gmm_cluster", spending.cols)]
gmm_raw_table$n <- sapply(gmm_raw_table$gmm_cluster, function(g) {
  sum(data$gmm_cluster == g)
})
gmm_raw_table$Cluster <- paste0("GMM-", gmm_raw_table$gmm_cluster)
gmm_raw_table$Label <- gmm_names[as.character(gmm_raw_table$gmm_cluster)]

combined_raw <- rbind(
  km3_raw_table[, c("Cluster", "Label", "n", spending.cols)],
  gmm_raw_table[, c("Cluster", "Label", "n", spending.cols)]
)
for (col in spending.cols) {
  combined_raw[[col]] <- format(round(combined_raw[[col]]), big.mark = ",")
}

kable(
  combined_raw,
  booktabs = TRUE,
  escape = FALSE,
  col.names = c(
    "Cluster",
    "Label",
    "$n$",
    "Fresh",
    "Milk",
    "Grocery",
    "Frozen",
    "Detergents\\_Paper",
    "Delicassen"
  ),
  align = c("l", "l", "r", rep("r", 6)),
  caption = "Cluster spending profiles (median annual spending, m.u.)",
  linesep = ""
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  pack_rows("K-Means (k = 3)", 1, 3, bold = TRUE, hline_before = FALSE) %>%
  pack_rows(
    "Gaussian Mixture Model (k = 5)",
    4,
    8,
    bold = TRUE,
    hline_before = TRUE
  )
```
```{r q2-tables-2-apd, echo=FALSE}
km3_pct_table <- km3_medians[, c("km3_cluster", paste0(spending.cols, "_pct"))]
km3_pct_table$Cluster <- paste0("KM-", km3_pct_table$km3_cluster)
km3_pct_table$Label <- km3_names[as.character(km3_pct_table$km3_cluster)]
colnames(km3_pct_table)[2:7] <- spending.cols

gmm_pct_table <- gmm_medians[, c("gmm_cluster", paste0(spending.cols, "_pct"))]
gmm_pct_table$Cluster <- paste0("GMM-", gmm_pct_table$gmm_cluster)
gmm_pct_table$Label <- gmm_names[as.character(gmm_pct_table$gmm_cluster)]
colnames(gmm_pct_table)[2:7] <- spending.cols

combined_pct <- rbind(
  km3_pct_table[, c("Cluster", "Label", spending.cols)],
  gmm_pct_table[, c("Cluster", "Label", spending.cols)]
)

kable(
  combined_pct,
  booktabs = TRUE,
  escape = FALSE,
  col.names = c(
    "Cluster",
    "Label",
    "Fresh",
    "Milk",
    "Grocery",
    "Frozen",
    "Detergents\\_Paper",
    "Delicassen"
  ),
  align = c("l", "l", rep("r", 6)),
  caption = "Cluster spending profiles (\\% of median basket)",
  linesep = ""
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  pack_rows("K-Means (k = 3)", 1, 3, bold = TRUE, hline_before = FALSE) %>%
  pack_rows(
    "Gaussian Mixture Model (k = 5)",
    4,
    8,
    bold = TRUE,
    hline_before = TRUE
  )
```

```{r q2-tables-4-apd, echo=FALSE}
km3_rev_table <- km3_sums_all[, c(
  "km3_cluster",
  paste0(spending.cols, "_rev_pct")
)]
km3_rev_table$n <- sapply(km3_rev_table$km3_cluster, function(k) {
  sum(data$km3_cluster == k)
})
km3_rev_table$Cluster <- paste0("KM-", km3_rev_table$km3_cluster)
km3_rev_table$Label <- km3_names[as.character(km3_rev_table$km3_cluster)]
colnames(km3_rev_table)[2:7] <- spending.cols

gmm_rev_table <- gmm_sums_all[, c(
  "gmm_cluster",
  paste0(spending.cols, "_rev_pct")
)]
gmm_rev_table$n <- sapply(gmm_rev_table$gmm_cluster, function(g) {
  sum(data$gmm_cluster == g)
})
gmm_rev_table$Cluster <- paste0("GMM-", gmm_rev_table$gmm_cluster)
gmm_rev_table$Label <- gmm_names[as.character(gmm_rev_table$gmm_cluster)]
colnames(gmm_rev_table)[2:7] <- spending.cols

combined_rev <- rbind(
  km3_rev_table[, c("Cluster", "Label", spending.cols)],
  gmm_rev_table[, c("Cluster", "Label", spending.cols)]
)

kable(
  combined_rev,
  booktabs = TRUE,
  escape = FALSE,
  col.names = c(
    "Cluster",
    "Label",
    "Fresh",
    "Milk",
    "Grocery",
    "Frozen",
    "Detergents\\_Paper",
    "Delicassen"
  ),
  align = c("l", "l", rep("r", 6)),
  caption = "Revenue concentration: each cluster's share (\\%) of total
             category spend across all 440 customers.",
  linesep = ""
) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  pack_rows("K-Means (k = 3)", 1, 3, bold = TRUE, hline_before = FALSE) %>%
  pack_rows(
    "Gaussian Mixture Model (k = 5)",
    4,
    8,
    bold = TRUE,
    hline_before = TRUE
  )
```


Table \@ref(tab:q2-tables-3-apd), \@ref(tab:q2-tables-2-apd), and
\@ref(tab:q2-tables-4-apd) present the cluster spending profiles from three
perspectives.

1. Table \@ref(tab:q2-tables-2-apd) shows the percentage composition of each
cluster's median basket examining how each segment allocates its spending across
categories.
1. Table \@ref(tab:q2-tables-3-apd) reports the raw median annual spending
(monetary units [m.u.]) per category, revealing the absolute scale differences
between segments. 
1. Table \@ref(tab:q2-tables-4-apd) reports revenue concentration and
corresponds directly to the stacked bars in Figure
\@ref(fig:q2-revenue-triptych). Note that because Table
\@ref(tab:q2-tables-3-apd) reports the **median** annual spending, it is not
directly linked to Table \@ref(tab:q2-tables-4-apd).

The K-Means "`r km3_names[horeca_cluster]`" cluster 
(KM-`r horeca_cluster`) allocates `r km3_stat(horeca_cluster, "Fresh_pct")`% of 
its median basket 
(Table \@ref(tab:q2-tables-2-apd)) to `r format_var_name("Fresh")` and accounts 
for `r km3_rev(horeca_cluster, "Fresh_rev_pct")`% of total Fresh revenue 
(Table \@ref(tab:q2-tables-4-apd)), 
while "`r km3_names[retail_cluster]`" (KM-`r retail_cluster`) concentrates on 
`r format_var_name("Grocery")` 
(`r km3_stat(retail_cluster, "Grocery_pct")`%) and 
`r format_var_name("Detergents_Paper")` 
(`r km3_stat(retail_cluster, "Detergents_Paper_pct")`%) but comprises only 
`r sum(km3$cluster == as.integer(retail_cluster))` customers 
(`r round(sum(km3$cluster == as.integer(retail_cluster)) / nrow(data) * 100)`% 
of the base). At the GMM level, the distinction sharpens further: the median 
GMM-3 customer ("`r gmm_names["3"]`") dedicates `r gmm_stat(3, "Fresh_pct")`% 
of its basket 
to `r format_var_name("Fresh")` and, despite representing 
`r gmm_pct_customers(3)`% of customers, captures 
`r gmm_rev(3, "Fresh_rev_pct")`% of total Fresh revenue. The Retail split is 
equally informative: the typical GMM-1 customer ("`r gmm_names["1"]`") 
concentrates on `r format_var_name("Grocery")` (`r gmm_stat(1, "Grocery_pct")`%) 
and `r format_var_name("Detergents_Paper")` 
(`r gmm_stat(1, "Detergents_Paper_pct")`%), while the typical GMM-2 customer 
("`r gmm_names["2"]`") spreads spend more evenly across categories including 
`r format_var_name("Fresh")` (`r gmm_stat(2, "Fresh_pct")`%), suggesting a 
convenience-store or diversified retailer profile.


### Sankey diagram of spending for Gaussian Mixed Mixture and KMeans clusters

```{r sankey-plot, echo=FALSE, fig.width=12, fig.height=10, fig.cap="Sankey diagram of spending across clusters. Each stacked bar shows the percentages of total spendings broken down into product categories (left), Gaussian Mixed Mixture clusters (middle), and KMeans (k=3) clusters (right). The flows on the left describes how each GMM cluster is comprised of spendings in different product categories. The flows on the right represent how the spendings are segmented differently by KMeans. Only flows with percentage value greater than 3% are labelled to minimise clutter."}

categories <- c("Fresh", "Milk", "Grocery", "Frozen", "Detergents_Paper", "Delicassen")

sk_wide <- data %>%
  pivot_longer(cols = all_of(categories),
               names_to = "category",
               values_to = "spend") %>%
  group_by(category, km3_cluster, gmm_cluster) %>%
  summarise(total_spend = sum(spend), .groups = "drop") %>%
  mutate(
    km3_cluster = paste0("KM-", km3_cluster),
    gmm_cluster = paste0("GMM-", gmm_cluster)
  ) %>%
  mutate(perc = total_spend / sum(total_spend))

# In case we want to set color based on a column
sk_color_lookup <- sk_wide %>%
  select(category, km3_cluster, gmm_cluster) %>%
  #mutate(edge_color = km3_cluster)
  mutate(edge_color = gmm_cluster)

sankeyfier_df <- sk_wide %>%
  left_join(sk_color_lookup, by = c("category", "km3_cluster", "gmm_cluster")) %>%
  pivot_stages_longer(
    #stages      = c("category", "km3_cluster", "gmm_cluster"),
    #stages      = c("km3_cluster", "category", "gmm_cluster"), # category in the middle
    stages      = c("category", "gmm_cluster", "km3_cluster"), # gmm_cluster in the middle
    values_from = "perc",
    additional_aes_from = c('edge_color')
  )

sk_palette <- createPalette(N = length(unique(sankeyfier_df$node)),
                            seedcolors = c("#B22222", "#228B22", "#4682B4")
                            )

names(sk_palette) <- unique(sankeyfier_df$node)

sk_pos  <- position_sankey(width = 0.3, nudge_x = - 0.15)
sk_pos_text_node <- position_sankey(text = TRUE, nudge_x = - 0.28)
sk_pos_text_edge <- position_sankey(text = TRUE, nudge_x = 0.1)

sk_label_threshold <- 0.03

ggplot(
  sankeyfier_df,
  aes(
    x        = stage,
    y        = perc,
    group    = node,
    connector = connector,
    edge_id  = edge_id
  )
) +
  # nodes
  geom_sankeynode(
    aes(fill = node),
    position = sk_pos,
    color    = "black"
  ) +
  # node labels
  geom_text(
    aes(label = node),
    stat     = "sankeynode",
    position = sk_pos_text_node,
    hjust    = 0,
    size     = 3
  ) +
  # edges (flows)
  geom_sankeyedge(
    aes(fill = edge_color),
    #aes(fill = node),
    position = sk_pos,
    alpha    = 0.8
  ) +
  # edge-level percentage labels
  geom_text(
    stat     = "sankeyedge",
    aes(
      label = ifelse(after_stat(y) > sk_label_threshold, scales::percent(after_stat(y), accuracy = 0.1), ' ')
    ),
    position = sk_pos_text_edge,
    size     = 3,
    color    = "black"
  ) +
  scale_fill_manual(values = sk_palette) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_discrete(labels = c(
    category     = "Category",
    km3_cluster  = "KMeans Cluster",
    gmm_cluster  = "GMM Cluster"
  )) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    axis.title.x    = element_blank()
  ) +
  labs(
    title = "",
    y     = "Percentage of Total Spending",
  )

```

Figure \@ref(fig:sankey-plot) is a Sankey diagram visualising the relationship between Gaussian Mixed Mixture (GMM) and KMeans clusters in terms of percentages of total spending. Most notably, `GMM-3` captures a segment from both `KM-1` and `KM-3` customers that have high spending in `r format_var_name("Fresh")` and `r format_var_name("Frozen")` products. The remaining subset of `KM-1` mainly becomes `GMM-2`, which predominantly are in retail channel.


### Cluster assignments in PCA space {#apd:cluster-pca}

```{r q2-plots-apd, echo=FALSE, fig.width=10, fig.height=12, fig.cap="Cluster assignments from K-Means (k=3, top) and GMM (k=5, bottom) projected onto the first two principal components, which together capture 71.3 percent of total variance. All clustering was performed in the full six-dimensional standardised log-spending space. Apparent overlap between clusters may reflect the loss of separating information in the remaining 28.7 percent of variance captured by PC3--PC6, rather than genuine ambiguity in cluster assignment."}

pca_scores <- as.data.frame(pca_result$x[, 1:2])
colnames(pca_scores) <- c("PC1", "PC2")

pc1_var <- summary(pca_result)$importance[2, 1] * 100
pc2_var <- summary(pca_result)$importance[2, 2] * 100
pc1_lab <- sprintf("PC1 (%.1f%%)", pc1_var)
pc2_lab <- sprintf("PC2 (%.1f%%)", pc2_var)

pca_scores$KM3 <- factor(km3$cluster)
pca_scores$GMM5 <- factor(data$gmm_cluster)

p_km <- ggplot(pca_scores, aes(x = PC1, y = PC2, colour = KM3, shape = KM3)) +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.3) +
  geom_vline(xintercept = 0, colour = "grey80", linewidth = 0.3) +
  geom_point(alpha = 0.7, size = 1.5) +
  scale_colour_brewer(palette = "Set1") +
  labs(
    title = "(a) K-Means (k = 3)",
    x = pc1_lab,
    y = pc2_lab,
    colour = "Cluster",
    shape = "Cluster"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

p_gmm_pca <- ggplot(
  pca_scores,
  aes(x = PC1, y = PC2, colour = GMM5, shape = GMM5)
) +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.3) +
  geom_vline(xintercept = 0, colour = "grey80", linewidth = 0.3) +
  geom_point(alpha = 0.99, size = 1.5) +
  scale_colour_brewer(palette = "Set2") +
  scale_shape_manual(values = c(16, 17, 15, 18, 3)) +
  labs(
    title = "(b) GMM (k = 5)",
    x = pc1_lab,
    y = pc2_lab,
    colour = "Cluster",
    shape = "Cluster"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

p_km / p_gmm_pca
```

Figure \@ref(fig:q2-plots-apd) projects the K-Means and GMM cluster assignments
onto PCA space. Both panels show that the dominant cluster boundaries divide the
data along PC1 (the "channel axis" identified in \@ref(qone)), confirming that
the clustering and PCA analyses converge on the same dominant structure. The GMM
panel shows finer subdivision, particularly along PC2, consistent with GMM's
ability to model ellipsoidal clusters of varying density.

## Ward's hierarchical clustering {#apd:wards-clust}

```{r apd-wards-cluster, echo=FALSE, message=FALSE, warning=FALSE}
# Ward's hierarchical clustering
# this code was moved up to Q2 (commented out here for transparency)
# dist_mat <- dist(scaled_data, method = "euclidean")
# hc <- hclust(dist_mat, method = "ward.D2")
# hc_clusters <- cutree(hc, k = 3)

# table(KMeans = KMeans.results[[3]]$cluster, Ward = hc_clusters)
# cross-tabulation
cross_tab <- table(KMeans.results[[3]]$cluster, Ward = hc_clusters)

# Ename the rows and columns 
rownames(cross_tab) <- paste("K-Means", 1:3)
colnames(cross_tab) <- paste("Ward", 1:3)

# format
kable(
  cross_tab, 
  booktabs = TRUE, 
  caption = "K-Means vs Ward's hierarchical clustering assignments (k=3)."
) %>%
  kable_styling(latex_options = "hold_position") 
```

Table \@ref(tab:apd-wards-cluster) cross-tabulates K-Means (k=3) and Ward's
hierarchical clustering (k=3) assignments. The largest cluster shows
near-perfect agreement (201 of 212 observations). Disagreement is concentrated
among boundary customers between the two smaller segments, reflecting the
different optimisation strategies. K-Means minimises within-cluster sum of
squares globally, while Ward's method does so greedily through successive
agglomeration. 


```{=latex}
\clearpage
\newpage
```

# Prompts and LLM outputs {#apd-llm-information}

```{r llm-runner-setup, include=FALSE}
library(callr)

llm_plot_dir <- "LLM-output/figures"
dir.create(llm_plot_dir, showWarnings = FALSE, recursive = TRUE)

run_llm_script <- function(script_path, plot_prefix = "llm") {
  plot_dir <- file.path(llm_plot_dir, plot_prefix)
  dir.create(plot_dir, showWarnings = FALSE, recursive = TRUE)
  
  old_plots <- list.files(plot_dir, pattern = "\\.png$", full.names = TRUE)
  if (length(old_plots) > 0) file.remove(old_plots)
  
  result <- callr::r(function(sp, pd) {
    setwd(dirname(sp))
    png(file.path(pd, "plot_%03d.png"), width = 900, height = 600, res = 150)
    
    output <- capture.output({
      tryCatch(
        source(sp, print.eval = TRUE),
        error = function(e) cat("\n== SOURCE ERROR ==\n", e$message, "\n")
      )
    }, type = "output")
    
    dev.off()
    list(output = output)
  }, args = list(
    sp = normalizePath(script_path),
    pd = normalizePath(plot_dir)
  ), show = FALSE)
  
  # Print text output inside a fenced code block so it formats properly
  if (length(result$output) > 0) {
    cat("\n```\n")
    cat(result$output, sep = "\n")
    cat("\n```\n\n")
  }
  
  # Embed plots using raw markdown syntax (works with results='asis')
  plots <- sort(list.files(plot_dir, pattern = "\\.png$", full.names = TRUE))
  for (p in plots) {
    cat(paste0("![](", p, ")\n\n"))
  }
}
```

A consistent visual format is used to distinguish between prompts sent to the
LLM, its responses, and our commentary.

::: {.promptbox}

**Prompts to the LLM** - shown with a right-hand border. These are the instructions and
questions we provided to the model.

:::

:::: {.responsebox}

**LLM Response** - shown with a left-hand border, forming a conversational
flow with the prompt. Responses may include a mix of explanation and code:
```r
summary(mtcars$mpg)
```

::::

::: {.insightbox}

**Our group insight and critical evaluation** - observations or key findings drawn from the output (i.e., what we found).

:::

::: {.alertbox}

**Caution: points requiring verification** - potential concerns, limitations, or caveats with the generated output (i.e., what we question).

:::

The R code generated by each LLM was executed programmatically within this
document using a sandboxed runtime environment. Each script was run in an
isolated R subprocess via the `callr` package, ensuring that no libraries,
variables, or side effects from one script could contaminate the main document
environment or subsequent scripts. Console output and diagnostic messages were
captured and reproduced verbatim, while graphical output was intercepted via a
PNG device and embedded automatically. Any runtime errors were caught and
reported inline rather than halting the document compilation. 

This approach was used to ensure that the displayed output faithfully reflected
what each LLM-generated script produced when executed from a clean R session.

```{=latex}
\clearpage
\newpage
```


## Claude (Sonnet 4.6) dialogue

```{r clean-env-for-llm-claude, include=FALSE} 
keep_vars <- c("word_count", "keep_vars")
# rm(list = setdiff(ls(), keep_vars))
```

The full chat history is available via: https://claude.ai/share/516ed676-d419-43ec-98a1-63a5eee67549 


<!-- prettier-ignore -->
::: promptbox 

### Prompt 1

Role : You are a data scientist working in a company that does wholesale
distribution to clients located in Portugal.

\vspace{12pt}

task : Your company is looking to implement a Data-driven customer segmentation
to inform inventory planning, pricing strategy, and sales team deployment.

\vspace{12pt}

I have a dataset called Wholesale_customers_data.csv located at
`.\Wholesale_customers_data.csv`. It contains 440 rows and 8 columns: Channel
(1=Hotel/Restaurant/Café, 2=Retail), Region (1=Lisbon, 2=Oporto, 3=Other), and
six continuous spending variables (Fresh, Milk, Grocery, Frozen,
Detergents_Paper, Delicassen) representing annual spending in monetary units.
Please write R code to perform PCA on this dataset. Include any preprocessing
you think is necessary.

:::

<!-- prettier-ignore -->
:::: {.responsebox}

### Response 1

`> Check for relevant skills`

No file uploaded, but I have a clear description of the dataset. Let me write
comprehensive R code for PCA on this wholesale customers dataset.

`> Created a file, read a file`

Here's a walkthrough of every decision made in the script: 

**Pre-processing Rationale**

**Why separate Channel & Region from the PCA?** They are categorical identifiers,
not spending measurements. Mixing measurement scales would corrupt the
covariance structure. They are kept as labels to colour the biplots afterwards,
which is far more analytically useful.

**Why log1p-transform before PCA?** Wholesale spend data is almost always severely
right-skewed (a few very large clients dominate). PCA is sensitive to scale and
variance, so without this step a single high-spend outlier can hijack the first
principal component. log1p (log(x+1)) compresses the long tail and handles any
zero values gracefully.

**Why scale. = TRUE inside prcomp?** Even after log-transformation the six variables
have different magnitudes (e.g., Fresh spending is typically much larger than
Delicassen). Standardising to unit variance ensures each variable contributes
equally to the decomposition.

**Why run KMO & Bartlett's tests?** These confirm that the correlation structure is
strong enough to make PCA meaningful. Bartlett's test (H₀: correlation matrix =
identity) should be rejected (p < 0.05), and KMO > 0.6 indicates adequate
sampling adequacy.

**What the Script Produces**

- Section 2: Missing value check, outlier boxplots, skewness table...
- Section 3: `prcomp` PCA object, full rotation matrix & summary
- Section 4: Eigenvalue table, Scree plot with Kaiser line...
- Section 5: Loading heatmap for retained PCs, cos² contribution plot
- Section 6: Biplots coloured by Channel and Region separately...
- Section 7: PC scores merged back → exported as `wholesale_pca_scores.csv`

**What to Expect Analytically**

Based on known patterns in this dataset:

- PC1 (~44% variance) - a general spending scale axis; high scores = large/high-value clients
- PC2 (~22% variance) - a channel contrast axis; Fresh loads negatively (HoReCa), Grocery/Detergents_Paper positively (Retail)
- PC3 (~12% variance) - a niche category axis differentiating Frozen/Delicassen buyers

Together, ~2–3 PCs typically explain 75–85% of variance, making the PC scores an
excellent lower-dimensional input for your subsequent K-Means or Hierarchical
clustering step.

::::

The **R** code returned, and the outputs generated by running this code, are
shown below:

```{r llm-claude-prompt-01-display, code=suppressWarnings(readLines("LLM-output/LLM-code-claude-prompt-01.R")), eval=FALSE, echo=TRUE, attr.source='.numberLines'}
```
```{r llm-claude-prompt-01-run, echo=FALSE, results='asis', error=TRUE}
run_llm_script("LLM-output/LLM-code-claude-prompt-01.R", "llm_claude_prompt01")
```


<!-- prettier-ignore -->
::: insightbox 

Claude's preprocessing pipeline closely mirrors our own: `log1p` transformation,
standardisation via `prcomp(scale. = TRUE)`, and exclusion of Channel and Region.
The inclusion of KMO and Bartlett's tests for sampling adequacy goes beyond our
analysis and represents a useful diagnostic we did not perform. The code
produces a comprehensive set of outputs including loadings heatmaps, cos²
contribution plots, and biplots coloured by both Channel and Region.

:::

<!-- prettier-ignore -->
::: alertbox 
Claude's prediction that PC1 represents a "general spending scale axis"
contradicts the actual loadings, where PC1 clearly separates
Grocery/Milk/Detergents_Paper from Fresh/Frozen - a channel axis, not a scale
axis. This is corrected in later discussion but illustrates how LLM-generated
interpretations can be confidently wrong when not grounded in computed results.
The claim that PC2 captures ~22% of variance also differs from the actual ~27%,
suggesting these numbers are recalled from training data rather than derived.
:::

---

::: {.promptbox}

### Prompt 2

Now write R code to perform clustering on this dataset. Explain your choice of
method and number of clusters.

:::

:::: {.responsebox}

### Response 2

`> Check the PCA R script for context on variable names and preprocessing steps used`

Good - I have full context from the PCA script. Now I'll write a comprehensive,
well-structured clustering script that picks up exactly where the PCA left off.

`> Created a file, read a file`

Here is a full explanation of every methodological decision made in the script.

**Method Choice: Why a Two-Stage Approach?**

**Stage 1 - Hierarchical Agglomerative Clustering (HAC) with Ward.D2.**
HAC is used first and exploratorily because it requires no pre-specified *k*.
The dendrogram shows the full merge history of the data, and natural groupings
can be identified by looking for the largest vertical gaps (the longest branches
before clusters merge).

**Why Ward.D2 specifically?** Ward's linkage merges the two clusters that produce
the smallest increase in total within-cluster sum of squares - it is inherently
trying to form compact, internally coherent groups. The other common options have
practical problems here: single linkage suffers from chaining (long thin
clusters), complete linkage is sensitive to outliers (which wholesale spend data
has), and average linkage is a compromise but rarely as clean. The agglomerative
coefficient printed by the script confirms this - Ward.D2 will produce a value
closest to 1.0, signalling the strongest cluster structure.

**Stage 2 - K-Means on PCA Scores.**
Once *k* is confirmed, K-Means is used for the final assignment because it is
interpretable, fast, and produces hard cluster memberships that can be exported
straight into a CRM or ERP system. Running it on the retained PC scores (rather
than all six raw variables) removes the noise carried by the low-variance PCs
and avoids double-counting the high correlations between Grocery, Milk and
Detergents_Paper.

**Number of Clusters: Why k = 3?**
Four independent criteria are computed and they all point to the same answer:

- Dendrogram cut height - the two largest height jumps in the Ward.D2 tree occur at the merges that would collapse 3 → 2 and 4 → 3, indicating 3 is the most stable cut
- Elbow (WSS) - the rate of WSS reduction flattens sharply after *k* = 3
- Average Silhouette Width - peaks at *k* = 3, meaning observations are most clearly assigned to their own cluster at that point
- Gap Statistic - *k* = 3 satisfies the `firstSEmax` rule (first *k* within one SE of the maximum gap)
- NbClust 30-index vote - the plurality of the 30 different cluster-quality indices vote for *k* = 3

Beyond the statistics, *k* = 3 also has a compelling business logic: it maps onto
three operationally distinct customer archetypes (see the summary at the bottom
of the script), each requiring a different inventory, pricing, and sales
approach. *k* = 2 is too coarse to distinguish within HoReCa; *k* = 4+ starts
splitting segments that are not meaningfully different for planning purposes.

**What the Script Produces (in order)**

- Section 4: Agglomerative coefficients table, full dendrogram, *k* = 3 coloured dendrogram
- Section 5: Elbow plot, Silhouette width plot, Gap Statistic plot, NbClust majority-vote table
- Section 6: Final K-Means solution, cluster sizes, between/total SS ratio
- Section 7: Silhouette plot for final solution, Calinski-Harabasz index
- Section 8: PCA biplot with 90% ellipses, cluster × Channel overlay, cluster × Region overlay
- Section 9: Cluster profile table, radar chart, grouped bar chart, per-category boxplots
- Section 10: HAC vs K-Means agreement check (expect >85%)
- Section 11: `wholesale_clustered.csv` - full dataset with `Cluster` column appended

::::

The **R** code returned, and the outputs generated by running this code, are
shown below:

```{r llm-claude-prompt-02-display, code=suppressWarnings(readLines("LLM-output/LLM-code-claude-prompt-02.R")), eval=FALSE, echo=TRUE, attr.source='.numberLines'}
```
```{r llm-claude-prompt-02-run, echo=FALSE, results='asis', error=TRUE}
run_llm_script("LLM-output/LLM-code-claude-prompt-02.R", "llm_claude_prompt02")
```

<!-- prettier-ignore -->
::: insightbox 
The two-stage approach (HAC for exploration, K-Means for final assignment) is
methodologically sound and well-justified. The rationale for Ward.D2 over other
linkages is detailed and accurate. The decision to cluster on PCA scores rather
than raw variables is a defensible alternative to our approach of clustering on
all six standardised spending variables - and the similar outcomes strengthen
confidence in the segmentation. The use of four independent k-selection criteria
(dendrogram, elbow, silhouette, gap statistic) exceeds our two-method approach.
:::

<!-- prettier-ignore -->
::: alertbox 
The script contains two runtime errors: (1) the HAC linkage method string is
misspecified, causing the agglomerative clustering to fail; (2) in section 9, a
column rename collision occurs when joining cluster labels to the profile table,
producing incorrect output. More fundamentally, k=3 is presented as the expected
answer before any diagnostic is computed - the code is structured to confirm a
predetermined conclusion rather than discover one. Claude's detailed cluster
profile descriptions (e.g., "High-Volume Fresh Buyers") are stated with
certainty despite no data having been analysed, raising the same memorisation
concern identified in the main text.
:::

---

::: {.promptbox}

### Prompt 3

Are there any alternative methods or robustness checks you would recommend?

:::

:::: {.responsebox}

### Response 3

**Alternative Clustering Methods**

**1. Gaussian Mixture Models (GMM / Model-Based Clustering).**
K-Means assumes clusters are spherical and equally sized - a hard geometric
constraint. GMM relaxes this by fitting a probabilistic mixture of multivariate
Gaussians, allowing clusters to be elliptical, differently sized, and differently
oriented. Each customer gets a soft probability of belonging to each cluster
rather than a hard label. In R: `mclust::Mclust()` - it uses BIC to
automatically select both the number of clusters and the covariance structure
(e.g., diagonal vs. full, tied vs. variable). On wholesale data, where the
HoReCa segment often has a very different spread from the Retail segment, GMM
frequently produces more honest assignments. Prefer it over K-Means when the
biplot shows clusters that are elongated or unequal in density - which is likely
here given the strong Channel separation on PC2.

**2. DBSCAN (Density-Based Spatial Clustering).**
DBSCAN identifies clusters as dense regions of points separated by sparse
regions, and critically, it classifies low-density points as noise/outliers
rather than forcing them into a cluster. This is valuable for wholesale data
because a handful of very large spenders (e.g., a major supermarket chain) can
distort K-Means centroids significantly. In R: `dbscan::dbscan()` - requires
tuning two parameters: `eps` (neighbourhood radius) and `minPts` (minimum points
to form a core). The `kNNdistplot()` function helps select `eps` via a knee in
the k-NN distance curve. Limitation: DBSCAN struggles with clusters of varying
density (common in spending data), and it does not generalise naturally to new
observations - a limitation if you want to score new customers joining the
database.

**3. K-Medoids / PAM (Partitioning Around Medoids).**
Like K-Means but uses actual data points as cluster centres (medoids) rather than
arithmetic means. This makes it substantially more robust to outliers, since a
single extreme spender cannot pull a centroid arbitrarily far. It also works with
any distance metric, not just Euclidean. In R: `cluster::pam()` - also produces
a silhouette plot natively. A useful direct comparison: if PAM and K-Means
produce very similar cluster assignments, it confirms the K-Means solution is not
being distorted by outliers.

**4. Agglomerative HAC with Alternative Linkages.**
The script already uses Ward.D2, but for a robustness check you should formally
compare it against Complete and Average linkage and compute the Cophenetic
Correlation Coefficient for each - this measures how faithfully the dendrogram
preserves the original pairwise distances. The linkage with the highest
cophenetic correlation is the most geometrically honest representation of the
data's structure.

**Robustness Checks**

**5. Stability Analysis via Bootstrap (`clusterboot`).**
The most important robustness check you can run. `fpc::clusterboot()` repeatedly
resamples the data (with replacement), re-clusters each bootstrap sample, and
measures how consistently the same clusters re-emerge using the Jaccard
similarity coefficient. A cluster with mean Jaccard > 0.75 is considered stable;
below 0.5 it is likely an artefact of the particular sample. If the three
clusters survive this test, there is strong evidence that the segmentation is a
genuine feature of the population of wholesale customers in Portugal, not just of
these 440 observations.

**6. Sensitivity to Outliers.**
Wholesale spend data almost always contains extreme values (e.g., a single very
large hotel group). Two targeted checks: first, winsorising the top 1–5% of each
spending variable before clustering, then comparing cluster assignments to the
unwinsorised solution using the Adjusted Rand Index (ARI) - an ARI close to 1.0
means the solution is insensitive to those extreme points. Second, leave-one-out
influence analysis: iteratively remove the top 10 highest-spend customers and
re-run K-Means. If cluster memberships for the remaining 430 customers shift
substantially, the centroids are being pulled by individual outliers.

**7. Sensitivity to the Number of PCs.**
The script retains PCs by the Kaiser criterion (eigenvalue ≥ 1), which typically
gives 3 PCs explaining ~80% of variance. A robustness check is to re-run
clustering using (a) 2 PCs only and (b) all 6 standardised variables directly
(no dimensionality reduction), and compare the three solutions with ARI.
Convergence across all three input representations confirms that the structure is
not an artefact of the PCA truncation.

**8. Multiple Random Initialisations & Consensus Clustering.**
K-Means with `nstart = 50` already guards against poor local minima, but a
stronger check is consensus clustering via the `ConsensusClusterPlus` package. It
runs the algorithm hundreds of times on subsampled data and builds a consensus
matrix - an N×N matrix where entry (i,j) records the proportion of runs in which
customers i and j ended up in the same cluster. A well-separated block structure
in the consensus heatmap is strong visual confirmation that the clusters are
stable and not sensitive to initialisation.

**9. External Validation Against Channel and Region.**
Since Channel and Region are known labels, they can be used as external
validators without having been used in the clustering itself. Compute the
Adjusted Rand Index between cluster assignments and Channel labels, and the
Normalised Mutual Information between cluster assignments and Region labels. A
high ARI against Channel (expect ~0.5–0.7 on this dataset) confirms the
unsupervised algorithm is recovering a structure that aligns with a known
business reality - a strong sign of face validity.

**Summary Recommendation**
In priority order for a report:

- Run `clusterboot()` - the single most persuasive robustness check for any clustering paper
- Fit a GMM with `Mclust()` - the most principled alternative; compare ARI against K-Means
- Run PAM and compare ARI - confirms outlier robustness
- Compute cophenetic correlations across linkage methods - validates the HAC choice
- Test sensitivity to PC retention - shows the structure is not an artefact of dimensionality reduction

If the consensus across all five checks points to three clusters with consistent
membership, you have a very defensible segmentation - both statistically and for
a business audience.

::::

<!-- prettier-ignore -->
::: insightbox 

The response provides a comprehensive and well-structured overview of
alternative methods and robustness checks, including GMM, DBSCAN, PAM, bootstrap
stability analysis (clusterboot), outlier sensitivity, and consensus clustering.
The priority ordering at the end is practical and well-reasoned. The suggestion
to test sensitivity to the number of retained PCs is particularly valuable -
this is a methodological check we did not perform.


:::

<!-- prettier-ignore -->
::: alertbox 

No executable code was provided for this prompt, making it impossible to verify
whether the suggested approaches would work on this dataset. Several
recommendations (e.g., ConsensusClusterPlus) require Bioconductor packages not
available in standard R installations, which could cause issues for
reproducibility. The response reads more as a textbook survey than a
data-specific recommendation - none of the suggested parameter values (e.g., eps
for DBSCAN) are grounded in the actual data.

:::

---

::: {.promptbox}

### Prompt 4

*Pasting back comments the LLM produced.*

```{r, echo=FALSE, comment=""}
cat("\n")
cat("================================================================\n")
cat("          CUSTOMER SEGMENTATION – BUSINESS SUMMARY             \n")
cat("================================================================\n")
cat("\n  METHOD:  HAC (Ward.D2) to identify structure +\n")
cat("           K-Means (k=3) on PCA scores for final assignment\n")
cat("  INPUT:   log1p-transformed, standardised spend variables\n")
cat("           projected onto retained PCs (Kaiser criterion)\n")
cat("\n  CLUSTER 1 – 'High-Volume Fresh Buyers' (HoReCa dominant)\n")
cat("    • Highest Fresh and Frozen spend\n")
cat("    • Low Grocery, Detergents_Paper\n")
cat("    → Hotels, restaurants, cafés; perishable-led ordering\n")
cat("    IMPLICATION: Prioritise cold-chain logistics; daily/weekly\n")
cat("    replenishment cycles; dedicated HoReCa sales reps\n")
cat("\n  CLUSTER 2 – 'Packaged Goods Retailers' (Retail dominant)\n")
cat("    • Highest Grocery, Milk, Detergents_Paper\n")
cat("    • Low Fresh spend\n")
cat("    → Supermarkets, convenience stores; ambient goods focus\n")
cat("    IMPLICATION: Bulk-order pricing tiers; shelf-stable\n")
cat("    inventory buffers; quarterly promotions on detergents\n")
cat("\n  CLUSTER 3 – 'Small / Mixed Buyers'\n")
cat("    • Below-average spend across all categories\n")
cat("    • Mixed Channel and Region membership\n")
cat("    → Small independent shops, delis, mixed traders\n")
cat("    IMPLICATION: Route-optimised delivery (lower margin);\n")
cat("    upsell programmes; self-service ordering portal\n")
cat("\n  NOTE: Exact cluster labels (1/2/3) may differ from above\n")
cat("  depending on K-Means initialisation; inspect profile_raw\n")
cat("  and the radar/bar charts to assign the business labels.\n")
cat("================================================================\n")
```
these were your comments  and they are correct and apply exactly on the data
that i have, so i'm wondering did you somehow analysized the data yourself

:::

:::: {.responsebox}

### Response 4

**Did the LLM analyse the actual data?**
No - the LLM did not have access to the data file. The descriptions it provided
were an informed prediction based on two things:

**1. This is a well-known public dataset.** The UCI Wholesale Customers dataset is
one of the most widely used benchmark datasets in clustering tutorials and
academic papers. The LLM had seen it discussed extensively in its training data,
and therefore knew its typical structure - the HoReCa/Retail split on PC2, the
three-cluster solution, and the rough spending profiles each cluster tends to
produce.

**2. Domain logic for Portuguese wholesale distribution.** Even without knowing the
specific dataset, the business logic is fairly predictable: restaurants and
hotels will dominate Fresh and Frozen spend - that is the nature of their
business. Supermarkets and retailers will dominate Grocery, Milk, and
Detergents_Paper - ambient shelf goods are their core stock. Any wholesale
customer base will contain a long tail of smaller, lower-spend mixed buyers. The
cluster descriptions were therefore a combination of prior knowledge about this
specific dataset and general domain reasoning about wholesale food distribution.

**Implication for interpretation.** Since this dataset is publicly known, the
"expected" results are well established. What differentiates a strong analysis is
not just reproducing the three clusters, but the quality of the methodological
justification - why `log1p`, why Ward.D2, why PCA before clustering, and why the
robustness checks matter. That reasoning is what the code and discussion above
has been building toward.

::::

<!-- prettier-ignore -->
::: insightbox 

Claude's candid acknowledgment that it recognised the dataset from training data
and was predicting known patterns rather than deriving results is valuable
transparency. The distinction between "prior knowledge about this specific
dataset" and "general domain reasoning" is honest and well-articulated.

:::

<!-- prettier-ignore -->
::: alertbox 
This response confirms the central limitation identified in the main text: the
LLM's analytical outputs for this dataset cannot be distinguished from memorised
pattern-matching. The practical implication is that for novel, proprietary
datasets - precisely the commercial setting motivating data privacy - the LLM's
recommendations would lack this hidden grounding and may be less reliable than
they appeared here.
:::

```{=latex}
\clearpage
\newpage
```

## ChatGPT (GPT-5.3) dialogue

```{r clean-env-for-llm-chatgpt, include=FALSE} 
# rm(list = setdiff(ls(), keep_vars))
```

The full chat history is available via four separate links (one for each prompt): 

1. https://chatgpt.com/s/t_69ceeadadfa48191ae79291539197208
2. https://chatgpt.com/s/t_69cef39137948191b9a8d8c3f29f7843
3. https://chatgpt.com/s/t_69cef34ba0ac8191aa01c2c15c4c8c2d
4. https://chatgpt.com/s/t_69cef077f1e481918fdab882f931936e

<!-- prettier-ignore -->
::: promptbox 

### Prompt 1

Role : You are a data scientist working in a company that does wholesale
distribution to clients located in Portugal.

\vspace{12pt}

task : Your company is looking to implement a Data-driven customer segmentation
to inform inventory planning, pricing strategy, and sales team deployment.

\vspace{12pt}

I have a dataset called Wholesale_customers_data.csv located at
`.\Wholesale_customers_data.csv`. It contains 440 rows and 8 columns: Channel
(1=Hotel/Restaurant/Café, 2=Retail), Region (1=Lisbon, 2=Oporto, 3=Other), and
six continuous spending variables (Fresh, Milk, Grocery, Frozen,
Detergents_Paper, Delicassen) representing annual spending in monetary units.
Please write R code to perform PCA on this dataset. Include any preprocessing
you think is necessary.

:::

<!-- prettier-ignore -->
:::: {.responsebox}

### Response 1

Here's a clean, reproducible R workflow to perform PCA (Principal Component
Analysis) on your wholesale dataset, including appropriate preprocessing steps.

**Key considerations before PCA**

- Categorical variables (`Channel`, `Region`) should NOT be used directly in PCA (since PCA is for continuous variables).
- Scaling is essential because spending variables are on very different scales.
- Optional: log transformation can help reduce skewness (common in spending data).

**R Code for PCA**
*Moved to below this response box*

**What this gives you**

- Principal components capturing key spending patterns
- Explained variance to decide how many components to retain
- Customer visualization (potential clusters/segments)
- Variable importance (which spending categories drive differences)

**Next Steps (for your segmentation task)**
After PCA, you can:

1. Use the first few principal components (e.g., PC1–PC3)
2. Apply clustering (e.g., `kmeans`, `hclust`)
3. Profile segments by:
   - Channel (HORECA vs Retail)
   - Region
   - Spending behavior

If you want, I can extend this into a full segmentation pipeline (PCA +
clustering + interpretation) tailored to your business objectives.

::::

The **R** code returned, and the outputs generated by running this code, are
shown below:

```{r llm-chatgpt-prompt-01-display, code=suppressWarnings(readLines("LLM-output/LLM-code-chatgpt-prompt-01.R")), eval=FALSE, echo=TRUE, attr.source='.numberLines'}
```
```{r llm-chatgpt-prompt-01-run, echo=FALSE, results='asis', error=TRUE}
run_llm_script("LLM-output/LLM-code-chatgpt-prompt-01.R", "chatgpt_prompt01")
```

<!-- prettier-ignore -->
::: insightbox 
ChatGPT's code executes without errors. Preprocessing
matches our pipeline: log transformation, standardisation, Channel/Region
excluded. The code is well-commented and concise, making it immediately usable.
The "Next Steps" section appropriately suggests clustering as a follow-on
analysis.
:::

<!-- prettier-ignore -->
::: alertbox 
The PCA interpretation is generic - PC1 and PC2 are not described in terms of
what they represent commercially. The code produces plots but the accompanying
text does not interpret the loadings or connect them to the Channel distinction.
A user relying solely on ChatGPT's commentary would not understand that PC1 is
the channel axis without examining the output themselves.
:::

---

::: {.promptbox}

### Prompt 2

Now write R code to perform clustering on this dataset. Explain your choice of
method and number of clusters.

:::

:::: {.responsebox}

### Response 2

Below is a complete R workflow for clustering, building directly on the PCA-ready
preprocessing. I'll also explain why this method and how to choose the number of
clusters.

**Why this approach?**

**Method: K-means clustering**

- Works well for continuous, scaled data
- Efficient for your dataset size (440 rows)
- Produces clear, actionable segments for business use (inventory, pricing, sales)

**Why use PCA first?**

- Reduces noise and multicollinearity (e.g., Grocery ↔ Detergents_Paper)
- Improves clustering stability
- Makes clusters easier to visualize

**Choosing number of clusters (k)**
We'll use:

- Elbow method (within-cluster variance)
- Silhouette score (cluster separation quality)

Typically, this dataset yields ~3–5 clusters, often:

- Retail-heavy bulk buyers
- HORECA fresh-focused buyers
- Low-volume or niche customers

**R Code: Clustering Pipeline**
*Moved to below this response box*

**Interpretation Guide**
After running this, you should:

1. Pick k
   - Look for:
     - "Elbow" bend in WSS plot
     - Peak silhouette score
   - Usually k = 3 or 4 works well here
2. Interpret clusters (example patterns)
   You may find:
   - Cluster 1: High Grocery + Detergents → Retail chains
   - Cluster 2: High Fresh + Frozen → HORECA clients
   - Cluster 3: Low across all → Small / infrequent buyers
3. Translate into business actions
   - Inventory planning: stock Fresh vs Grocery differently per segment
   - Pricing strategy: volume discounts for retail clusters
   - Sales deployment: assign reps by segment type

**Optional (better alternative)**
If you want more robust segmentation:

- Try Hierarchical clustering (for validation)
- Or Gaussian Mixture Models (GMM) for probabilistic clusters

If you want, I can help you turn this into a fully polished report with cluster
personas and business recommendations-that's usually what gets top marks.

::::

The **R** code returned, and the outputs generated by running this code, are
shown below:

```{r llm-chatgpt-prompt-02-display, code=suppressWarnings(readLines("LLM-output/LLM-code-chatgpt-prompt-02.R")), eval=FALSE, echo=TRUE, attr.source='.numberLines'}
```
```{r llm-chatgpt-prompt-02-run, echo=FALSE, results='asis', error=TRUE}
run_llm_script("LLM-output/LLM-code-chatgpt-prompt-02.R", "llm_chatgpt_prompt02")
```

<!-- prettier-ignore -->
::: insightbox 
The code runs cleanly and produces appropriate diagnostics (elbow plot,
silhouette scores). The decision to cluster on PCA scores is justified with a
clear rationale (noise reduction, multicollinearity). The "Interpretation Guide"
section provides a practical framework for translating cluster assignments into
business actions.
:::

<!-- prettier-ignore -->
::: alertbox 
K-Means is presented as the default without comparison to alternatives in the
initial response - hierarchical clustering and GMM are mentioned only as
"optional" afterthoughts. The cluster descriptions ("Retail chains", "HORECA
clients", "Small/infrequent buyers") are presented as expected outcomes before
any code is run, echoing the same memorisation concern as Claude. The silhouette
and elbow plots are produced but not programmatically interpreted - the user
must visually inspect them and decide k themselves.
:::

---

::: {.promptbox}

### Prompt 3

Are there any alternative methods or robustness checks you would recommend?

:::

:::: {.responsebox}

### Response 3

Yes - and this is exactly the kind of thing that strengthens your analysis.
K-means + PCA is a solid baseline, but for a data-driven segmentation, you
should always validate that your clusters are stable, meaningful, and not
method-dependent. Here are the most important alternatives and robustness checks,
with R code.

**1. Alternative Clustering Methods**

**(A) Hierarchical Clustering (Validation + interpretability)**
Why:

- No need to pre-specify k upfront
- Produces a dendrogram (useful for justification in reports)
- Helps confirm whether k-means structure is real

*R code moved to below this response box*

If clusters align well with k-means → strong evidence of robustness

**(B) Gaussian Mixture Models (GMM)**
Why:

- Allows non-spherical clusters (k-means assumes spherical)
- Provides probabilistic membership (useful for borderline customers)

*R code moved to below this response box*

If GMM selects similar k → supports your choice. If different → suggests more
complex structure

**(C) DBSCAN (Density-based)**
Why:

- Detects outliers/noise
- Does not require specifying k

*R code moved to below this response box*

Useful to identify unusual customers (important for risk or niche segments)

**2. Robustness Checks**

**(A) Stability Analysis (Resampling)**
Why:

- Ensures clusters aren't driven by random variation

*R code moved to below this response box*

Values:

- \> 0.85 → very stable
- 0.6–0.85 → moderately stable
- < 0.6 → weak clusters

**(B) Silhouette Analysis (Quality check)**

*R code moved to below this response box*

Interpretation:

- \> 0.5 → strong structure
- 0.25–0.5 → reasonable
- < 0.25 → weak clustering

**(C) Compare Clustering on Raw vs PCA Data**

*R code moved to below this response box*

If results are similar → PCA did not distort structure. If very different →
investigate further

**(D) Sensitivity to k**

*R code moved to below this response box*

Ensures your choice of k is not arbitrary

**How to Present This in Your Assignment**
You can frame it like this:

- Primary method: K-means on PCA-reduced data
- Validation methods:
  - Hierarchical clustering → confirms structure
  - GMM → checks distributional assumptions
  - Silhouette + Elbow → supports choice of k
  - Stability analysis → confirms robustness

**Strong Conclusion (what markers look for)**
A high-quality conclusion would say:

- Clusters are consistent across methods
- k = 3 (or 4) is supported by multiple criteria
- Segments are stable and interpretable
- Results are actionable for business decisions

If you want, I can help you turn this into a top-grade "Methodology + Validation"
write-up with proper academic language and justification.

::::

```{r llm-chatgpt-prompt-03-display, code=suppressWarnings(readLines("LLM-output/LLM-code-chatgpt-prompt-03.R")), eval=FALSE, echo=TRUE, attr.source='.numberLines'}
```
```{r llm-chatgpt-prompt-03-run, echo=FALSE, results='asis', error=TRUE}
run_llm_script("LLM-output/LLM-code-chatgpt-prompt-03.R", "llm_chatgpt_prompt03")
```

<!-- prettier-ignore -->
::: insightbox 

Provides executable R code for each suggested method (hierarchical, GMM, DBSCAN,
bootstrap stability, silhouette analysis), unlike Claude's text-only response.
The code is designed to run sequentially after the clustering script,
maintaining variable continuity. The interpretation thresholds provided (e.g.,
Jaccard > 0.85 = very stable) give the user actionable benchmarks.
:::

<!-- prettier-ignore -->
::: alertbox 
The code for Prompt 3 depends on objects created in Prompt 2 - if run in
isolation it will fail. This dependency is not documented. The DBSCAN parameter
suggestion lacks dataset-specific tuning guidance. The "How to Present This in
Your Assignment" section, while practical, reveals that the LLM is optimising
for marks rather than analytical rigour - the framing advice ("what markers look
for") substitutes presentation strategy for genuine critical evaluation.
:::

---

::: {.promptbox}

### Prompt 4

*Pasting back comments the LLM produced.*

```{r, echo=FALSE, comment=""}

```
these were your comments  and they are correct and apply exactly on the data
that i have, so i'm wondering did you somehow analysized the data yourself

:::

:::: {.responsebox}

### Response 4
Good question-and fair to ask. No, I didn't actually access or run your specific
dataset. I can't see files on your computer or execute code on them unless you
explicitly upload data here. What I did instead is rely on well-known, highly
consistent patterns in this exact dataset (Wholesale Customers Data Set, commonly
used in clustering/PCA coursework). This dataset has been analyzed many times,
and it almost always produces very similar structures because of how the
variables behave.

**Why my interpretation matched your results**
There are strong, repeatable patterns in this data:

**1. Correlation structure drives clustering.**
Grocery, Milk, Detergents_Paper → highly correlated → typically forms a
"Retail-heavy" cluster. Fresh, Frozen → less correlated with the above →
typically forms a "HORECA / fresh-food" cluster.

**2. Channel variable reinforces it.**
Even though clustering is unsupervised: Channel = 2 (Retail) → high Grocery +
Detergents. Channel = 1 (HORECA) → higher Fresh/Frozen. So clustering naturally
reconstructs this split.

**3. Skewed spending distribution.**
A large group of customers has low spending across categories. This almost always
becomes a "small buyers" cluster.

**Why this happens (intuitively)**
Think of your customers as three business types:

- Retailers - buy packaged goods in bulk → Grocery + Detergents spike
- Hotels/Restaurants (HORECA) - buy perishable goods → Fresh + Frozen dominate
- Small clients - low, inconsistent purchasing → low across all

These behavioral differences are strong enough that almost any reasonable
clustering method will rediscover them.

**Important (for your assignment)**
Even though your results match expectations, you should not present them as
assumed. You should:

- Show cluster summaries (means per cluster)
- Compare with Channel distribution
- Use plots (PCA, boxplots)
- Support with validation metrics

**A stronger way to phrase it in your report**
Instead of: "Clusters represent retail vs HORECA"

Say something like: "Cluster analysis reveals three distinct customer segments
characterized by (i) high Grocery and Detergents spending, (ii) high Fresh and
Frozen spending, and (iii) uniformly low spending. Cross-tabulation with the
Channel variable shows strong alignment with Retail and HORECA customer types,
validating the segmentation."

**Bottom line**

- I didn't see your data
- Your results match because the dataset has very stable, well-known structure
- This is actually a good sign → your pipeline is working correctly

If you want, I can help you go one level deeper-like naming your clusters
professionally (e.g., "Bulk Retailers", "Fresh-Focused HORECA", "Low-Value
Accounts") and tying them tightly to strategy.

::::

<!-- prettier-ignore -->
::: insightbox 

Like Claude, ChatGPT openly acknowledges that the dataset is publicly known and that its predictions were based on well-established patterns in its training data. The explanation of why the patterns are predictable (correlation structure, Channel variable, skewed spending) is clear and would be useful pedagogically.

:::

<!-- prettier-ignore -->
::: alertbox 
The suggestion to phrase results as though they were discovered ("Cluster
analysis reveals...") rather than predicted is problematic advice - it
encourages presenting memorised outcomes as empirical findings. The "Bottom
line" framing ("your pipeline is working correctly") assumes the user's results
match the expected pattern, which may not always be the case and could
discourage investigation of unexpected outcomes.
:::


```{=latex}
\clearpage
\newpage
```

# Reproducibility, accessibility & declarations

## Reproducibility & accessibility

An accessible HTML version of this report is available via a public GitHub page:

- https://ytterbiu.github.io/SMM636-ML-group-coursework-03-g01/

This report was created in R Markdown. The source code is open source and is
available via:

- https://github.com/ytterbiu/SMM636-ML-group-coursework-03-g01

Changes made to this document were tracked using Git and are also available via
the same repository
(https://github.com/ytterbiu/SMM636-ML-group-coursework-03-g01/commits/main/).
Please note that this commit history reflects updates made to the project
repository and is not a comprehensive timeline of all contributions by group
members.

## Word count

The total word count for the core sections of this report is **`r word_count`**
words (`r ifelse(word_count > 1000, "above", "below")` the 1000 word limit).

The calculation was performed programmatically directly from the source `.rmd`
file. An **R** script was used to read the raw document lines and strip out
non-text elements, including the YAML front matter, R code chunks, sub-headers,
markdown formatting (tables and figures), and LaTeX/HTML syntax. The script
isolates lines falling within `#intro`, `#qone`, `#qtwo`, and `#qthr`. It then 
calculates the final tally using the `stringi` package.

The code chunk that performs this calculation is executed silently at the
beginning of this document, but is reproduced below for transparency:

```{r display-word-count-code, ref.label="word-count", eval=FALSE, echo=TRUE}
```

## Generative AI Declaration

- Will be best to ask about this during the lecture - will almost want this for
  the first part - or is the first part completely AI free?

<!-- Generative AI (GenAI) tools were used throughout this project to assist with -->
<!-- project planning, methodological validation, and debugging. In practice, these -->
<!-- tools functioned as an additional technical team member, providing immediate -->
<!-- feedback, troubleshooting code, and challenging statistical assumptions. -->

<!-- ### Tools Used -->

<!-- Google Gemini 3 Pro (Deep Research), ChatGPT 5.2 -->

<!-- ### Prompts used -->

<!-- The GenAI tools were provided with the 'coursework_CM2_250120.pdf' file and -->
<!-- extracts from our `R` code. -->

<!-- A non-exhaustive list of prompts used is shown below. We opted to include at -->
<!-- least one example of each _type_ of prompt used, rather than a list of all -->
<!-- prompts. For example, if we entered the prompts 'How can we put two ggplots side -->
<!-- by side?' and 'How to remove grey background on a ggplot', we would include only -->
<!-- the first of these in the list below, as both relate to plotting issues (& are -->
<!-- similar to 'querying' with a traditional search engine). -->

<!-- - "...the top of both graphs in r are slightly different. (Pasting R code.) Can -->
<!--   this be fixed for the plot area?" -->
<!-- - "Help to identify the key points here" -->
<!-- - "How can this error be tracked down in an rmd file? ! Text line contains an -->
<!--   invalid character. l.1 ^^@^^@..." -->

<!-- ### Use of GenAI outputs & subsequent revisions -->

<!-- The GenAI tools served as a "consultant" member of the group. Notably, the AI -->
<!-- output: -->

<!-- - Helped debug plotting issues. -->
<!-- - Commented on the structure of R code snippets and the report structure. -->

<!-- ### Changes made: -->

<!-- All AI suggestions were critically reviewed and debated by the group before -->
<!-- implementation. For example: -->

<!-- - While the AI suggested code for complex layouts, the specific aesthetic -->
<!--   choices and annotations were manually adjusted to ensure clarity. -->
<!-- - Where the AI proposed specific interpretations, these were cross-referenced -->
<!--   with academic literature and statistics textbooks. -->
<!-- - The final R code for implementation was written and debugged by the group with -->
<!--   AI snippets being treated as drafts rather than final solutions. -->

<!-- While GenAI was treated as a collaborator, the final analysis, statistical -->
<!-- interpretation, and the written content of this report were generated and -->
<!-- verified by the (human) group members. -->


<!-- BE note: taken out as we are submitting R file separately -->
<!-- Use: `knitr::purl("01SMM363-assessment-03.rmd", documentation = 0)` to generate r file -->
<!-- Use: `tools::showNonASCIIfile("cs1-group07.rmd")` to debug -->
<!-- ## R Code Documentation -->
<!-- ```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
``` -->

---

# References {-}