---
title:
  "Beyond the binary: Uncovering hidden customer segments in wholesale purchasing data"
subtitle: "SMM068 (2025-26) Coursework 03 - Group 01"
author:
  - "Abdulrahman Alolyan"
  - "Benjamin Evans"
  - "Basmah Khan"
  - "Ardi Wira Sudarmo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
papersize: a4
fontsize: 12pt
output:
  bookdown::html_document2:
    self_contained: true
    css: style.css
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: paper
    highlight: pygments
    df_print: paged
    pandoc_args: ["--lua-filter=custom-boxes.lua"]
  # runtime: shiny
  bookdown::pdf_document2:
    latex_engine: lualatex
    includes:
      in_header: preamble.tex
    pandoc_args: ["--lua-filter=custom-boxes.lua"]
    number_sections: true
    toc: true
    # citation_package: natbib
    keep_tex: true
bibliography: references.bib
csl: ieee.csl
---

```{=latex}
\newpage
```

```{=html}
<h1>Information</h1>

This R Markdown document was created as part of a group assignment for SMM636 at Bayes Business School, City St George's, University of London in Term 2 2025-26.

```

```{r asthetic-header, include=FALSE, echo=FALSE}
# ============================================================================ #
# Key Information ====
# SMM636 Machine Learning
# Group Coursework 2025-26
# Group:        Group 01
# Authors (in alphabetical order):
#   - Abdulrahman Alolyan
#   - Benjamin Evans
#   - Basmah Khan
#   - Ardi Wira Sudarmo
# Professor:    Dr. Rui Zhu
# Institution:  Bayes Business School - City St George's, University of London
# Date:         03 Apr 2026
#
# Description:  Term 2 group project for SMM636 Machine Learning
#               (25% of coursework grade - 25% of module grade). The R code
#               below has been exported directly from an R Markdown (.rmd) file.
#               Hence the knitr settings.
#
# Dependencies:
#   - tidyverse: loads dplyr, tidyr, ggplot2, purrr (& others)
#   - TBI
#
# Acknowledgements
#   - TBI
# ============================================================================ #
```

```{r setup-knitr, include=FALSE, echo=FALSE}
# Setup & configuration ====
#----------------------- Initial setup (knitr settings) -----------------------#
dir.create("fig", showWarnings = FALSE)

# Defaults common to all outputs
knitr::opts_chunk$set(
  echo     = TRUE,
  message  = FALSE,
  warning  = FALSE,
  fig.align = "center",
  out.width = "100%",
  fig.path  = "fig/",
  dpi       = 300
)

# Output-specific settings
if (knitr::is_latex_output()) {
  knitr::opts_chunk$set(
    fig.width = 6,
    fig.height = 4,
    dev = "pdf",
    fig.pos = "!ht",
    out.extra = ""
  )
} else {
  knitr::opts_chunk$set(
    fig.width = 6,
    fig.height = 4,
    dev = "svglite"  # or "png"
  )
}
```

```{r setup-qol, include=FALSE, echo=FALSE}
#----------------------------- Clean environment ------------------------------#
rm(list = ls()) # Remove all objects
graphics.off() # Close all graphical devices
cat("\014") # Clean console
```

```{r load-dependencies, include=FALSE, echo=FALSE}
#------------------- Load dependencies / external libraries -------------------#
# --- Core Data Manipulation ---
library(tidyverse)  # loads dplyr, tidyr, ggplot2, purrr, etc.

# --- Visualization & Plot Arrangement ---
library(patchwork)  # combining multiple ggplot objects
library(gridExtra)  # arranging base/grid plots
library(ggpubr)     # publication-ready plot themes/utilities
library(corrplot)   # correlation heatmaps
library(GGally)     # pairs plots
library(factoextra) # PCA & clustering visualizations

# --- Analytics & Modeling ---
library(NbClust)    # optimal-k diagnostics

# NOTE: The packages below are deliberately NOT loaded into the global 
# environment to prevent function masking conflicts (e.g., MASS::select 
# breaking dplyr::select). Call their functions directly using namespaces:
# cluster::daisy(), MASS::isoMDS()
# library(MASS)     
# library(cluster)

library(mclust)

# --- Reporting & Tables ---
library(kableExtra) # for formatting output tables
```

```{r custom-functions, include=FALSE, echo=FALSE}
#---------------------------- Custom QOL functions ----------------------------#
#####################################
# function: banner comments (used to to section up code)
# Usage: banner_comment("Element 1: data cleaning") -> then ctrl + v (or cmd+v)
#####################################
banner_comment <- function(text, width = 80, border = "#", fill = "-") {
  txt <- paste0(" ", text, " ")
  inner_width <- width - 2 * nchar(border)
  banner_string <- ""

  if (inner_width <= nchar(txt)) {
    banner_string <- paste0(border, txt, border)
  } else {
    pad_total <- inner_width - nchar(txt)
    pad_left <- pad_total %/% 2
    pad_right <- pad_total - pad_left

    banner_string <- paste0(
      border,
      strrep(fill, pad_left),
      txt,
      strrep(fill, pad_right),
      border
    )
  }

  cat(banner_string, "\n")
  # copy banner to allow direct pasting (requires clipr)
  clipr::write_clip(banner_string)
  # avoid [1] when printing if want to manually copy
  invisible(banner_string)
}
#####################################
# function: format p-values for text
# Usage (in-line): `r format_p_vals(ad_test_result$p.value)`
# Usage (console): format_p_vals(ad_test_result$p.value)
#####################################
format_p_vals <- function(p) {
  if (length(p) != 1L || is.na(p)) {
    stop("Error! p must be a single non-missing value")
  }
  if (p > 1) {
    stop("Error! Value greater than 1")
  }
  if (p < 0) {
    stop("Error! Value less than 0")
  }

  if (p >= 0.01) {
    paste0("= ", formatC(p, format = "f", digits = 2))
  } else if (p >= 0.001) {
    paste0("= ", formatC(p, format = "f", digits = 3))
  } else {
    "< 0.001"
  }
}
#####################################
# function: format confidence intervals for tables & text
# Usage (in-line): `r format_interval(el2_ci_normal_95[1], el2_ci_normal_95[2])`
# Usage (console): format_interval(el2_ci_normal_95[1], el2_ci_normal_95[2])
#####################################
format_interval <- function(lower, upper, digits = 3, small_interval = 5L) {
  paste0(
    "[",
    formatC(
      lower,
      format = "f",
      digits = digits,
      small.interval = small_interval,
      small.mark = " "
    ),
    ", ",
    formatC(
      upper,
      format = "f",
      digits = digits,
      small.interval = small_interval,
      small.mark = " "
    ),
    "]"
  )
}
#####################################
# function: format truncated ellipses
# Usage (in-line): `r tbi`
# Usage (console): tbi
#####################################
fmt_trunc_ellip <- function(
  x,
  digits = 7,
  ellip = "...",
  tol = 1e-12,
  trim_zeros_if_exact = TRUE
) {
  out <- rep(NA_character_, length(x))
  ok <- is.finite(x)
  scale <- 10^digits
  xt <- trunc(x[ok] * scale) / scale
  s <- formatC(xt, format = "f", digits = digits)
  add <- abs(x[ok] - xt) > tol * pmax(1, abs(x[ok]))
  if (trim_zeros_if_exact) {
    s_trim <- sub("0+$", "", s) # drop trailing zeros
    s_trim <- sub("\\.$", "", s_trim) # drop trailing decimal point
  } else {
    s_trim <- s
  }
  out[ok] <- ifelse(add, paste0(s, ellip), s_trim)
  out[is.na(x)] <- NA_character_
  out
}
#####################################
# function: function to format 6dp numbers with a space after the 3rd decimal
# digit
# Usage (in-line): TBI
# Usage (console): TBI
#####################################
format_spaced_6dp <- function(x) {
  # Format to 6 decimal places
  s <- sprintf("%.6f", x)
  # Insert space between the first 3 and last 3 decimal digits
  sub("(\\.[0-9]{3})([0-9]{3})$", "\\1 \\2", s)
}
#####################################
# function: format variable names in green in latex (& normal code in html)
# Usage (in-line): TBI `r format_var_name("dvisit")`
#####################################
format_var_name <- function(x) {
  if (knitr::is_latex_output()) {
    # replace all "_" with "\_" in va name
    paste0("\\greentt{", gsub("_", "\\\\_", x), "}")
  } else {
    # change to paste0("<code style='color:green'>", x, "</code>") for green
    # paste0("`", x, "`")
    paste0("<code style='color:green'>", x, "</code>")
  }
}
```

```{r word-count, include=FALSE}
rmd_lines <- readLines("01SMM363-assessment-03.rmd", warn = FALSE)

# ---- Remove YAML front matter ----
yaml_bounds <- which(rmd_lines == "---")
if (length(yaml_bounds) >= 2) {
  rmd_lines <- rmd_lines[-(yaml_bounds[1]:yaml_bounds[2])]
}

# ---- Remove code chunks ----
in_chunk <- FALSE
keep <- rep(TRUE, length(rmd_lines))
for (i in seq_along(rmd_lines)) {
  if (grepl("^```\\{", rmd_lines[i])) {
    in_chunk <- TRUE
    keep[i] <- FALSE
    next
  }
  if (in_chunk && grepl("^```\\s*$", rmd_lines[i])) {
    in_chunk <- FALSE
    keep[i] <- FALSE
    next
  }
  if (in_chunk) keep[i] <- FALSE
}
rmd_lines <- rmd_lines[keep]

# ---- Identify which lines fall inside counted sections ----
# Sections to include (level-1 headers only)
counted_ids <- c("intro", "qone", "qtwo", "qthr")
in_counted_section <- FALSE
include <- rep(FALSE, length(rmd_lines))

for (i in seq_along(rmd_lines)) {
  line <- rmd_lines[i]
  # Detect any level-1 header (# but not ##)
  if (grepl("^#\\s+[^#]", line)) {
    # Check if this header belongs to a counted section
    in_counted_section <- any(sapply(counted_ids, function(id) {
      # [^}]* allows any characters (like spaces, hyphens, classes) before the 
      # closing }
      # this is required to count the words in the unnumbered introduction
      grepl(paste0("\\{\\s*#", id, "[^}]*\\}"), line)
    }))
    next  # skip the header line itself
  }
  if (in_counted_section) include[i] <- TRUE
}

counted_lines <- rmd_lines[include]

# ---- Remove sub-headers (##, ###, etc.) ----
counted_lines <- counted_lines[!grepl("^#{2,}", counted_lines)]

# ---- Remove table lines (markdown pipes) ----
counted_lines <- counted_lines[!grepl("^\\|", counted_lines)]

# ---- Remove figure captions: ![caption](...) ----
counted_lines <- counted_lines[!grepl("^!\\[", counted_lines)]

# ---- Remove fenced div markers (::: name) ----
counted_lines <- counted_lines[!grepl("^:::", counted_lines)]

# ---- Remove LaTeX commands ----
counted_lines <- counted_lines[!grepl("^\\\\", counted_lines)]

# ---- Remove HTML comments ----
counted_lines <- counted_lines[!grepl("^<!--", counted_lines)]

# ---- Remove inline R references like \@ref() ----
counted_lines <- gsub("\\\\@ref\\([^)]*\\)", "", counted_lines)

# ---- Count ----
word_count <- sum(stringi::stri_count_words(counted_lines), na.rm = TRUE)

message("Word count (sections qone, qtwo, qthr, body text only): ", word_count)

# Write for LaTeX preamble
writeLines(
  paste0("\\newcommand{\\wordcount}{",
         formatC(word_count, big.mark = ","), "}"),
  "wordcount.tex"
)
```

```{=latex}
\newpage
\pagenumbering{arabic}
```

# Introduction {#intro -}

This report analyses annual spending data from 440 clients of a Portuguese
wholesale distributor [@Cardoso2013; @Abreu2011], comprising 298
Hotel/Restaurant/Café (HoReCa) and 142 Retail customers across six product
categories: `r format_var_name("Fresh")`,
`r format_var_name("Milk")`, `r format_var_name("Grocery")`,
`r format_var_name("Frozen")`, `r format_var_name("Detergents_Paper")`, and
`r format_var_name("Delicassen")`. Data-driven
customer segmentation informs inventory planning, pricing strategy, and
sales team deployment, yet the existing classification is a simple binary label
(`r format_var_name("Channel")`).
This analysis uses Principal Component Analysis (PCA) and clustering to
investigate whether the existing two-category classification adequately captures
the structure of customer spending, or whether the data supports a more nuanced
segmentation.

```{r load-data, echo=FALSE, include=FALSE}
set.seed(1234)

data <- read.csv("data/Wholesale_customers_data.csv")
str(data)
summary(data)

# just get the $ amounts spent
spending <- data[, c(
  "Fresh",
  "Milk",
  "Grocery",
  "Frozen",
  "Detergents_Paper",
  "Delicassen"
)]

spending_log <- log1p(spending)

rescale01 <- function(x) (x - min(x)) / (max(x) - min(x))

raw_df <- spending %>%
  pivot_longer(everything(), names_to = "Category", values_to = "Spend") %>%
  mutate(Transform = "Raw")

log_df <- spending_log %>%
  pivot_longer(everything(), names_to = "Category", values_to = "Spend") %>%
  mutate(Transform = "Log-transformed")

# add data to import into in-line text
# standardise log-transformed data
scaled_data <- scale(spending_log)

# PCA
pca_result <- prcomp(spending_log, center = TRUE, scale. = TRUE)

# Skewness (raw)
skew_vals <- apply(spending, 2, e1071::skewness)
min_skew  <- min(skew_vals)
```

```{r data-table, echo=knitr::is_html_output(), eval=knitr::is_html_output()}
DT::datatable(
  data,
  filter   = "top",
  options  = list(pageLength = 15, scrollX = TRUE),
  caption  = "Wholesale customers"
)
```

# Principle component analysis (PCA) {#qone}

<!-- _Apply PCA to the dataset and summarise and interpret key your findings_ -->

A log-transformation was applied to normalise the distribution and
address the right-skewness observed across all six spending variables 
(skewness > `r formatC(min_skew, format = "f", digits = 1)`
for all - see Appendix, Figures \@ref(fig:histograms-overlay) &
\@ref(fig:corrplots-side-by-side)). The variables were then standardised to
ensure equal contribution to the PCA, preventing categories with naturally
larger absolute expenditures from dominating the model.

The first two principal components capture 71.3% of the total variance (PC1:
44.1%; PC2: 27.2%; Figure \@ref(fig:pca-grid)(a)). The component loadings in
Figure \@ref(fig:pca-grid)(b) reveal distinct commercial purchasing patterns.
PC1 acts as the primary "channel axis," driven by strong positive loadings for
`r format_var_name("Grocery")` (0.571),
`r format_var_name("Detergents_Paper")` (0.551), and 
`r format_var_name("Milk")` (0.542), 
contrasted with near-zero or negative loadings for 
`r format_var_name("Fresh")` and `r format_var_name("Frozen")`. 
PC2 serves as a "perishable volume axis," characterised by high positive
loadings for `r format_var_name("Fresh")` (0.590), `r format_var_name("Frozen")`
(0.590), and `r format_var_name("Delicassen")` (0.530), capturing the scale of
fresh food purchasing independent of business type.

Figure \@ref(fig:pca-grid)(d) confirms this visually, showing clear
separation between HoReCa and Retail customers along PC1. PCA is restricted to
continuous inputs, necessitating the exclusion of `r format_var_name("Channel")`
and `r format_var_name("Region")`. A Gower-distance MDS analysis incorporating
these variables (Appendix, \@ref(apd:suppl-mds)) confirms that the categorical
labels dominate the distance metric, validating the decision to analyse spending
variables independently.


```{r pca-fit, echo=FALSE, include=FALSE}
# ==============================================================================
# Q1 PCA
# ==============================================================================

pca_result <- prcomp(spending_log, center = TRUE, scale. = TRUE)
summary(pca_result)

cumvar <- cumsum(pca_result$sdev^2 / sum(pca_result$sdev^2))
cat("Cumulative variance explained:\n")
print(round(cumvar, 3))

loadings <- pca_result$rotation[, 1:3]
print(round(loadings, 3))
```

```{r pca-scree, echo=FALSE, include=FALSE, fig.width=7, fig.height=5}
fviz_eig(
  pca_result,
  addlabels = TRUE,
  ylim = c(0, 60),
  main = "Scree Plot - Variance Explained by Each Principle Component (PC)"
)
```

```{r pca-biplot, echo=FALSE, include=FALSE, fig.width=8, fig.height=6}
fviz_pca_biplot(
  pca_result,
  label   = "var",
  col.var = "firebrick",
  col.ind = alpha("steelblue", 0.3),
  repel   = TRUE,
  title   = "PCA Biplot (log-transformed, standardised data)"
)
```

```{r pca-contrib-pc, echo=FALSE, include=FALSE, fig.width=7, fig.height=4}
# Opted to hide these for now 
# (don't think we have space and they don't add too much)
# Even including in appendix is limited - easier to include table
fviz_contrib(
  pca_result,
  choice = "var",
  axes   = 1,
  title  = "Variable Contributions to PC1"
)

fviz_contrib(
  pca_result,
  choice = "var",
  axes   = 2,
  title  = "Variable Contributions to PC2"
)
```

```{r pca-loadings-table, echo=FALSE, include=FALSE}
loadings_df <- as.data.frame(round(pca_result$rotation[, 1:3], 3))
colnames(loadings_df) <- c("PC1", "PC2", "PC3")

kable(
  loadings_df,
  caption  = "PCA loadings (first three components). Values over 0.4 are highlighted in bold.",
  linesep  = "",
  booktabs = TRUE
) %>%
  kable_styling(
    full_width = FALSE,
    position   = "center"
  ) %>%
  column_spec(2, bold = abs(loadings_df$PC1) >= 0.4) %>%
  column_spec(3, bold = abs(loadings_df$PC2) >= 0.4) %>%
  column_spec(4, bold = abs(loadings_df$PC3) >= 0.4)
```

```{r pca-channel, echo=FALSE, include=FALSE, fig.width=8, fig.height=6}
fviz_pca_ind(
  pca_result,
  geom.ind    = "point",
  col.ind     = factor(data$Channel),
  palette     = c("#E64B35", "#4DBBD5"),
  addEllipses = TRUE,
  legend.title = "Channel",
  title       = "PCA - Individuals Coloured by Channel (1=HoReCa, 2=Retail)"
)
```

```{r pca-grid, echo=FALSE, include=TRUE, fig.width=12, fig.height=10, fig.cap="Principal Component Analysis of log-transformed, standardised spending data. (a) Scree plot showing the percentage of variance explained by each component. (b) Loadings of the original variables on the first four principal components. Values exceeding 0.4 in magnitude indicate strong contributions. (c) Biplot overlaying variable loadings (red arrows) on individual observations. (d) Individuals projected onto PC1-PC2, coloured by distribution channel (1 = HoReCa, 2 = Retail) with 95% concentration ellipses."}

# --- Top left: Scree plot ---
p1 <- fviz_eig(
  pca_result,
  addlabels = TRUE,
  ylim = c(0, 60),
  main = "(a) Scree Plot - Variance Explained by Principle Component (PC)"
)

# --- Top right: Loadings table as a grob ---
loadings_df <- as.data.frame(round(pca_result$rotation[, 1:3], 3))
colnames(loadings_df) <- c("PC1", "PC2", "PC3")

# --- Top right: Loadings table ---
loadings_df <- as.data.frame(pca_result$rotation[, 1:4])
colnames(loadings_df) <- c("PC1", "PC2", "PC3", "PC4")
# loadings_df <- as.data.frame(pca_result$rotation[, 1:3])
# colnames(loadings_df) <- c("PC1", "PC2", "PC3")
loadings_df[] <- lapply(loadings_df, function(x) format(round(x, 3), nsmall = 3))

p2 <- ggtexttable(
  loadings_df,
  theme = ttheme(
    "blank",
    tbody.style = tbody_style(
      size  = 13,
      hjust = 1.00, #controls alignment {0.00, 0.5, 1.00} 
      x     = 0.95, #controls alignment {0.05, 0.5, 0.95} 
      fill  = "white"
    ),
    colnames.style = colnames_style(
      size  = 13,
      face  = "bold",
      hjust = 1.00,
      x     = 0.95,
      fill  = "white"
    ),
    rownames.style = rownames_style(
      size  = 13,
      face  = "plain",
      hjust = 0,
      x     = 0.05,
      fill  = "white"
    )
  )
) %>%
  tab_add_hline(at.row = 1:2, row.side = "top", linewidth = 2) %>%
  tab_add_hline(at.row = nrow(loadings_df) + 1, row.side = "bottom", linewidth = 2) +
  labs(title = "(b) PCA Loadings") +
  theme(
    plot.title      = element_text(hjust = 0.5, size = 13),
    plot.background = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "white", colour = NA)
  )

# --- Bottom left: Biplot ---
p3 <- fviz_pca_biplot(
  pca_result,
  label   = "none",
  col.var = "firebrick",
  col.ind = alpha("steelblue", 0.3),
  title   = "(c) PCA Biplot (log-transformed, standardised data)"
)

# Pull arrow endpoints from layer 4 of the built plot
arrows <- ggplot_build(p3)$data[[4]]

var_df <- data.frame(
  x     = arrows$xend,
  y     = arrows$yend,
  label = rownames(pca_result$rotation)
)

p3 <- p3 +
  ggrepel::geom_text_repel(
    data        = var_df,
    aes(x = x, y = y, label = label),
    colour      = "firebrick",
    size        = 3.5,
    bg.color    = "white",
    bg.r        = 0.15,
    inherit.aes = FALSE
  )

# --- Bottom right: Channel plot ---
p4 <- fviz_pca_ind(
  pca_result,
  geom.ind     = "point",
  col.ind      = factor(data$Channel),
  palette      = c("#E64B35", "#4DBBD5"),
  addEllipses  = TRUE,
  legend.title = "Channel",
  title        = "(d) PCA - Individuals Coloured by Channel (1=HoReCa, 2=Retail)"
)

# BE note: qol to get bottom two plots to align vertically
# ggplot_build(p3)$layout$panel_params[[1]]$y.range
# ggplot_build(p4)$layout$panel_params[[1]]$y.range

p3 <- p3 + coord_cartesian(ylim = c(-5.58, 5.24))
p4 <- p4 + coord_cartesian(ylim = c(-5.58, 5.24))

# --- Combine 2x2 ---
(p1 | p2) / (p3 | p4)
```


<!-- ```{=latex} -->
<!-- \clearpage -->
<!-- \newpage -->
<!-- ``` -->

# Customer clustering {#qtwo}

```{r q2-prep, echo=FALSE, include=FALSE}
# ==============================================================================
# Q2 Clustering
# ==============================================================================

# K-Means for k=2 through 5
KMeans.results <- list()
for (k in 2:5) {
  KMeans.results[[k]] <- kmeans(scaled_data, centers = k, nstart = 25)
}

# Assign K-Means labels programmatically
km3 <- KMeans.results[[3]]
km3_ch_tab <- table(km3$cluster, data$Channel)
horeca_pct <- km3_ch_tab[, 1] / rowSums(km3_ch_tab)

horeca_cluster <- as.character(which.max(horeca_pct))
retail_cluster <- as.character(which.min(horeca_pct))
mixed_cluster  <- as.character(setdiff(1:3, c(horeca_cluster, retail_cluster)))

km3_names <- setNames(
  c("HoReCa Dominant", "Retail Dominant", "Mixed Basket"),
  c(horeca_cluster, retail_cluster, mixed_cluster)
)
km3_colours <- setNames(
  c("#B2182B", "#2166AC", "#FDB863"),
  c(horeca_cluster, retail_cluster, mixed_cluster)
)
km3_order <- c(retail_cluster, mixed_cluster, horeca_cluster)

# silhouette scores (referenced inline)
sil_scores <- sapply(2:10, function(k) {
  km <- kmeans(scaled_data, centers = k, nstart = 25)
  sil <- cluster::silhouette(km$cluster, dist(scaled_data))
  mean(sil[, 3])
})
names(sil_scores) <- 2:10

# GMM on spending data (referenced inline for $G, $modelName, $bic)
gmm_spending <- Mclust(data = scaled_data)

# BIC comparison table (referenced inline)
bic_vals    <- gmm_spending$BIC
best_per_k  <- apply(bic_vals, 1, max, na.rm = TRUE)

# Ward's hierarchical clustering (referenced in main text)
dist_mat    <- dist(scaled_data, method = "euclidean")
hc          <- hclust(dist_mat, method = "ward.D2")
hc_clusters <- cutree(hc, k = 3)

# Shorthand references used in main text ====
km3 <- KMeans.results[[3]]
km2 <- KMeans.results[[2]]
data$km3_cluster <- km3$cluster
data$km2_cluster <- km2$cluster
data$gmm_cluster <- as.integer(gmm_spending$classification)

# Median basket % per cluster (K-Means)
km3_medians <- aggregate(
  cbind(Fresh, Milk, Grocery, Frozen, Detergents_Paper, Delicassen) ~ km3_cluster,
  data = data, FUN = median
)
km3_medians$total <- rowSums(km3_medians[, spending.cols])
for (col in spending.cols) {
  km3_medians[[paste0(col, "_pct")]] <- round(km3_medians[[col]] / km3_medians$total * 100, 1)
}

# Revenue concentration % per cluster (K-Means)
km3_sums_all <- aggregate(
  cbind(Fresh, Milk, Grocery, Frozen, Detergents_Paper, Delicassen) ~ km3_cluster,
  data = data, FUN = sum
)
for (col in spending.cols) {
  km3_sums_all[[paste0(col, "_rev_pct")]] <- round(km3_sums_all[[col]] / sum(km3_sums_all[[col]]) * 100, 1)
}

# Same for GMM
gmm_medians <- aggregate(
  cbind(Fresh, Milk, Grocery, Frozen, Detergents_Paper, Delicassen) ~ gmm_cluster,
  data = data, FUN = median
)
gmm_medians$total <- rowSums(gmm_medians[, spending.cols])
for (col in spending.cols) {
  gmm_medians[[paste0(col, "_pct")]] <- round(gmm_medians[[col]] / gmm_medians$total * 100, 1)
}

gmm_sums_all <- aggregate(
  cbind(Fresh, Milk, Grocery, Frozen, Detergents_Paper, Delicassen) ~ gmm_cluster,
  data = data, FUN = sum
)
for (col in spending.cols) {
  gmm_sums_all[[paste0(col, "_rev_pct")]] <- round(gmm_sums_all[[col]] / sum(gmm_sums_all[[col]]) * 100, 1)
}

# Helper: look up a value by cluster ID
km3_stat <- function(cluster_id, col) {
  row <- km3_medians$km3_cluster == as.integer(cluster_id)
  km3_medians[row, col]
}
km3_rev <- function(cluster_id, col) {
  row <- km3_sums_all$km3_cluster == as.integer(cluster_id)
  km3_sums_all[row, col]
}
gmm_stat <- function(cluster_id, col) {
  row <- gmm_medians$gmm_cluster == cluster_id
  gmm_medians[row, col]
}
gmm_rev <- function(cluster_id, col) {
  row <- gmm_sums_all$gmm_cluster == cluster_id
  gmm_sums_all[row, col]
}
gmm_n <- function(cluster_id) {
  sum(data$gmm_cluster == cluster_id)
}
gmm_pct_customers <- function(cluster_id) {
  round(sum(data$gmm_cluster == cluster_id) / nrow(data) * 100)
}
```


```{r q2-revenue-triptych, echo=FALSE, fig.width=12, fig.height=10, fig.cap="Revenue concentration by product category under three segmentation approaches. Top: the distributor's existing binary Channel classification. Middle: K-Means (k=3) spending-based clusters. Bottom: Gaussian Mixture Modelling (GMM) (k=5) spending-based clusters. Crucially, the K-Means and GMM segmentations were derived solely from log-transformed, standardised spending data without access to Channel labels. Each bar shows the percentage of total category spend attributable to each segment and the Total column shows each segment's share of overall revenue."}

n_total <- nrow(data)
spend_cols <- c(
  "Fresh",
  "Milk",
  "Grocery",
  "Frozen",
  "Detergents_Paper",
  "Delicassen"
)
cat_levels <- c(spend_cols, "Total")

# ==============================================================================
# Helper: from a grouping vector, produce a revenue-concentration data frame
# ==============================================================================
build_shares <- function(group_vec, group_name = "cluster") {
  
  # Sums per group
  sums <- aggregate(
    cbind(Fresh, Milk, Grocery, Frozen, Detergents_Paper, Delicassen) ~ grp,
    data = cbind(data[, spend_cols], grp = group_vec),
    FUN  = sum
  )
  
  # Column percentages
  pcts <- sums
  for (col in spend_cols) {
    pcts[[col]] <- sums[[col]] / sum(sums[[col]]) * 100
  }
  
  # Pivot long
  pcts_long <- reshape(
    pcts, direction = "long",
    varying = spend_cols, v.names = "Pct",
    timevar = "Category", times = spend_cols
  )
  pcts_long$Cluster <- factor(pcts_long$grp)
  row.names(pcts_long) <- NULL
  
  # Total column
  sums$total <- sums$Fresh + sums$Milk + sums$Grocery +
                sums$Frozen + sums$Detergents_Paper + sums$Delicassen
  totals <- data.frame(
    grp      = sums$grp,
    Category = "Total",
    Pct      = sums$total / sum(sums$total) * 100,
    Cluster  = factor(sums$grp)
  )
  
  # Combine
  out <- rbind(pcts_long[, c("grp", "Category", "Pct", "Cluster")], totals)
  out$Category <- factor(out$Category, levels = cat_levels)
  out
}

# ==============================================================================
# Helper: from a grouping vector, produce legend labels "Label-X (n=Y, Z%)"
# ==============================================================================
build_labels <- function(group_vec, names_dict = NULL, prefix = "Cluster") {
  counts <- as.data.frame(table(group_vec))
  colnames(counts) <- c("grp", "n")
  
  # Calculate percentages
  counts$pct <- (counts$n / n_total) * 100
  
  # Construct the label string
  if (!is.null(names_dict)) {
    # If a dictionary is provided, map the group number to your custom name
    counts$desc <- names_dict[as.character(counts$grp)]
    counts$label <- sprintf("%s (n=%d, %.0f%%)", counts$desc, counts$n, counts$pct)
  } else {
    # Fallback to the original Prefix-X format if no dictionary is supplied
    counts$label <- sprintf("%s-%s (n=%d, %.0f%%)", prefix, counts$grp, counts$n, counts$pct)
  }
  
  # Return a named vector required by scale_fill_manual / scale_fill_brewer
  setNames(counts$label, as.character(counts$grp))
}

# ==============================================================================
# Helper: standard revenue concentration plot
# ==============================================================================
plot_shares <- function(share_data, fill_labels, title, palette = "Set1",
                        manual_colours = NULL) {
  p <- ggplot(share_data, aes(x = Category, y = Pct, fill = Cluster)) +
    geom_col(position = "stack", width = 0.7) +
    geom_text(
      aes(label = sprintf("%.0f%%", Pct)),
      position = position_stack(vjust = 0.5),
      size = 3, colour = "white"
    ) +
    geom_vline(xintercept = 6.5, linetype = "dashed", colour = "grey40") +
    labs(title = title, y = "Share of total category spend (%)",
         x = "", fill = "Segment") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
  
  if (!is.null(manual_colours)) {
    p <- p + scale_fill_manual(values = manual_colours, labels = fill_labels)
  } else {
    p <- p + scale_fill_brewer(palette = palette, labels = fill_labels)
  }
  p
}

# ==============================================================================
# Build all three plots
# ==============================================================================

# Define the descriptive segment names based on the spending profiles
channel_names <- c(
  "1" = "HoReCa",
  "2" = "Retail"
)

# km3_names <- c(
#   "1" = "Mixed Basket",
#   "2" = "HoReCa Dominant",
#   "3" = "Retail Dominant"
# )

gmm_names <- c(
  "1" = "Heavy Retail",
  "2" = "Diversified Retail",
  "3" = "High-volume HoReCa",
  "4" = "Balanced HoReCa",
  "5" = "Low-spend / Mixed"
)

# factor ordering: HoReCa-type first (bottom), Retail-type last (top)
channel_order <- c("2", "1") 
# HoReCa, Retail

# km3_order <- c("3", "1", "2") 
# # HoReCa Dominant, Mixed, Retail Dominant

gmm_order <- c("1", "2", "5", "4", "3") 
# High-vol HoReCa, Balanced HoReCa, Low-spend, Diversified Retail, Heavy Retail

gmm_colours <- c(
  "3" = "#B2182B",   # deep red - high-vol HoReCa
  "4" = "#E66101",   # orange - balanced HoReCa
  "5" = "#FDB863",   # amber - low-spend mixed
  "2" = "#92C5DE",   # light blue - diversified Retail
  "1" = "#2166AC"    # deep blue - heavy Retail
)
# km3_colours <- c(
#   "2" = "#B2182B",   # HoReCa Dominant
#   "1" = "#FDB863",   # Mixed Basket
#   "3" = "#2166AC"    # Retail Dominant
# )
ch_colours <- c("1" = "#E64B35", "2" = "#4DBBD5")

palettes <- list(
  rdylbu = list(
    gmm = c("3"="#B2182B", "4"="#E66101", "5"="#FDB863", "2"="#92C5DE", "1"="#2166AC"),
    km3 = c("2"="#B2182B", "1"="#FDB863", "3"="#2166AC"),
    ch  = c("1"="#E64B35", "2"="#4DBBD5")
  ),
  okabe_ito = list(
    gmm = c("3"="#D55E00", "4"="#E69F00", "5"="#F0E442", "2"="#56B4E9", "1"="#0072B2"),
    km3 = c("2"="#D55E00", "1"="#F0E442", "3"="#0072B2"),
    ch  = c("1"="#D55E00", "2"="#0072B2")
  ),
  viridis = list(
    gmm = c("3"="#440154", "4"="#3B528B", "5"="#21918C", "2"="#5EC962", "1"="#FDE725"),
    km3 = c("2"="#440154", "1"="#21918C", "3"="#FDE725"),
    ch  = c("1"="#440154", "2"="#FDE725")
  )
)

active <- palettes$rdylbu

gmm_colours <- active$gmm
km3_colours <- active$km3
ch_colours  <- active$ch

# (a) Channel baseline
ch_vec    <- as.integer(data$Channel)
ch_shares <- build_shares(ch_vec)
ch_shares$Cluster  <- factor(ch_shares$Cluster,  levels = channel_order)
ch_labels <- build_labels(ch_vec, names_dict = channel_names)

p_channel <- plot_shares(
  ch_shares,
  ch_labels,
  title = "(a) Baseline: Channel",
  manual_colours = ch_colours
) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank()
  )

# (b) K-Means k=3
km3_vec    <- as.integer(KMeans.results[[3]]$cluster)
km3_shares <- build_shares(km3_vec)
km3_shares$Cluster <- factor(km3_shares$Cluster, levels = km3_order)
km3_labels <- build_labels(km3_vec, names_dict = km3_names)

p_km3 <- plot_shares(
  km3_shares,
  km3_labels,
  title = "(b) K-Means (k = 3)",
  manual_colours = km3_colours
) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank()
  )

# (c) Gaussian Mixture Modelling k=5
gmm_vec    <- as.integer(gmm_spending$classification)
gmm_shares <- build_shares(gmm_vec)
gmm_shares$Cluster <- factor(gmm_shares$Cluster, levels = gmm_order)
gmm_labels <- build_labels(gmm_vec, names_dict = gmm_names)

p_gmm <- plot_shares(
  gmm_shares, gmm_labels,
  title = "(c) Gaussian Mixture Modelling (k = 5)", 
  manual_colours = gmm_colours
)

# ==============================================================================
# Stack
# ==============================================================================
p_channel / p_km3 / p_gmm + plot_layout(heights = c(1, 1, 1.15))
```


K-Means clustering was applied to the same log-transformed, standardised
spending data used for PCA. Silhouette analysis identifies k=2 as optimal 
(`r formatC(sil_scores["2"], format = "f", digits = 3)`), which recovers the known
Channel distinction and validates the clustering approach. However, the
silhouette score for k=3 remains strong 
(`r formatC(sil_scores["3"], format = "f", digits = 3)`), and the marginal 
reduction in cluster quality yields a substantially richer segmentation. Beyond
k=3, scores plateau around 0.19 with no meaningful recovery (Appendix, Figure
\@ref(fig:silhouette-plot)), suggesting additional clusters fragment genuine
structure rather than reveal new segments.

K-Means was chosen for its interpretability and computational efficiency on this
dataset. As robustness checks, both Ward's hierarchical clustering and Gaussian
Mixture Modelling (GMM) were applied to the same input. Ward's method (k=3)
agrees with K-Means on the dominant cluster structure, with disagreement
confined to boundary observations between the smaller segments (Appendix
\@ref(apd:wards-clust)). GMM relaxes K-Means' spherical cluster assumption: the
BIC selected a `r gmm_spending$G`-component `r gmm_spending$modelName` model
(Appendix, Figures \@ref(fig:silhouette-plot) & \@ref(fig:bic-plot)),
splitting both channels into finer sub-segments. Cross-tabulation against
Channel and Region (Appendix, Tables
\@ref(tab:q2-tables-2-apd)–\@ref(tab:q2-tables-4-apd)) confirms that these
sub-segments are spending-driven rather than geographic, with Region distributed
roughly proportionally within each cluster.

Figure \@ref(fig:q2-revenue-triptych) presents the commercial impact of each
segmentation. Under the existing binary Channel label, HoReCa customers account
for 76% of Fresh revenue but only 68% of customers. 
K-Means (k=3) identifies a "`r km3_names[mixed_cluster]`" segment 
(`r sum(km3$cluster == as.integer(mixed_cluster))` customers, 
`r round(sum(km3$cluster == as.integer(mixed_cluster)) / nrow(data) * 100)`%) 
with spending patterns that cut across both channels.
GMM further distinguishes Heavy Retail from Diversified Retail, and separates
High-volume HoReCa from Balanced HoReCa, each with distinct category
concentration profiles.

The strength of this approach is that it operates entirely without access to
Channel or Region labels. The near-perfect recovery of the Channel distinction
validates the segmentation as behaviourally grounded, while the emergence of
additional clusters reveals sub-structure the existing classification does not
capture. However, the commercial value of a 5-segment model over a 3-segment
model cannot be determined from a single cross-sectional snapshot. Validation
would require longitudinal purchasing data, margin information, and ideally a
controlled trial of differentiated service strategies across segments.


# Repeating \@ref(qone) & \@ref(qtwo) using LLMs {#qthr}

- Using same prompts for 2 LLMs for consistency / standardiseation.
- 'to ensure compatibility identical prompts were used for both models' 
(described in full in the appendix \@ref(apd-llm-information))

Example table:

```{r llm-comparison, echo=FALSE}
# define symbols
sym <- list(
  Y = "$\\checkmark$",    # ● addressed unprompted
  P = "$\\circ$",         # ○ partial / after prompting
  N = "\\texttimes"       # ✗ missed entirely
)

sym_legend <- paste0(
  sym$Y, "~= addressed unprompted; ",
  sym$P, "~= partial; ",
  sym$N, "~= missed entirely."
)

# data structure 
# > can replace "?" with Y/P/N after running the LLM experiment

rows <- tribble(
  ~group,            ~label,                                          ~ours, ~chatgpt, ~claude,
  "Preprocessing",   "Excluded Channel \\& Region",                   "Y",   "?",      "?",
  "Preprocessing",   "Identified skewness in spending variables",      "Y",   "?",      "?",
  "Preprocessing",   "Applied log transform",                         "Y",   "?",      "?",
  "Preprocessing",   "TBC",  "Y",   "?",      "?",
  "PCA",             "Scree plot / variance explained",                "Y",   "?",      "?",
  "PCA",             "Interpreted PC1 as channel axis",                "Y",   "?",      "?",
  "PCA",             "TBC",         "Y",   "?",      "?",
  "Clustering",      "TBC",       "Y",   "?",      "?",
  "Clustering",      "TBC",    "Y",   "?",      "?",
  "Clustering",      "TBC",              "Y",   "?",      "?",
  "Clustering",      "TBC",       "Y",   "?",      "?",
  "Interpretation",  "Commercial framing of segments",                 "Y",   "?",      "?",
  "Interpretation",  "Acknowledged limitations",                       "Y",   "?",      "?",
  "Interpretation",  "Suggested alternative methods",                  "Y",   "?",      "?"
)

# map codes to symbols
map_sym <- function(x) {
  dplyr::case_match(x,
    "Y" ~ sym$Y,
    "P" ~ sym$P,
    "N" ~ sym$N,
    .default = x
  )
}

# compute totals pre symbol mapping
score_col <- function(col) {
  n_total <- sum(col %in% c("Y", "P", "N"))
  if (n_total == 0) return("---")
  n_y <- sum(col == "Y")
  n_p <- sum(col == "P")
  # format: "10/14 (2P)" or "14/14" if no partials
  if (n_p > 0) {
    paste0(n_y, "/", n_total, " (", n_p, "P)")
  } else {
    paste0(n_y, "/", n_total)
  }
}

totals <- tibble(
  group   = "Totals",
  label   = "\\textbf{Score}",
  ours    = score_col(rows$ours),
  chatgpt = score_col(rows$chatgpt),
  claude  = score_col(rows$claude)
)

# combine rows + totals, then map symbols
all_rows <- bind_rows(rows, totals) %>%
  mutate(across(c(ours, chatgpt, claude), map_sym))

#compute pack_rows indices from group column
group_indices <- all_rows %>%
  mutate(row = row_number()) %>%
  group_by(group) %>%
  summarise(start = min(row), end = max(row), .groups = "drop")

group_indices <- group_indices %>%
  mutate(group = factor(group, levels = unique(all_rows$group))) %>%
  arrange(group)

# build kable
kable_data <- all_rows %>%
  dplyr::select(label, ours, chatgpt, claude)

colnames(kable_data) <- c("Analysis Step", "Our Analysis", "ChatGPT (4o)", "Claude")

caption_text <- paste0(
  "Comparison of analytical steps: our pipeline vs \\ two LLMs. ",
  sym_legend
)

tbl <- kable(
  kable_data,
  format    = "latex",
  booktabs  = TRUE,
  escape    = FALSE,
  align     = c("l", "c", "c", "c"),
  caption   = caption_text,
  linesep   = ""
) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size     = 10
  )

# -apply pack_rows 
for (i in seq_len(nrow(group_indices))) {
  grp <- as.character(group_indices$group[i])
  # skip the totals row - it stands alone with a rule above it
  if (grp == "Totals") next
  tbl <- tbl %>%
    pack_rows(
      grp,
      group_indices$start[i],
      group_indices$end[i],
      bold         = TRUE,
      italic       = FALSE,
      hline_before = (i > 1)
    )
}

# if (knitr::is_latex_output()) {
#   # cat("\\begingroup\n")
#   cat("\\renewcommand{\\arraystretch}{1.3}\n")
# }

tbl 

# if (knitr::is_latex_output()) {
#   # cat("\\endgroup\n")
# }
# # tbl
```


```{=latex}
\newpage
```

<!-- ```{=html} -->
<!-- # References -->
<!-- <div id="refs"></div> -->
<!-- ``` -->

<!-- ```{=latex} -->
<!-- \bibliography{references.bib} -->
<!-- ``` -->

```{=latex}
\newpage
\section*{Appendices}
\addcontentsline{toc}{section}{Appendices}

% --- restore default section numbering in the appendix ---
\renewcommand{\thesection}{\arabic{section}}
\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
\renewcommand{\thesubsubsection}{\thesubsection.\arabic{subsubsection}}
```

# (APPENDIX) Appendices {-}

# Supplementary methodology {#apd:suppl-methods}

## Raw data distribution, correlations, and skewness {#apd:suppl-skew}

```{r boxplots-raw-data, echo=knitr::is_html_output(), include=TRUE, fig.width=12, fig.height=5.5, fig.cap="Customer spending profiles across six wholesale categories. Top: boxplots showing the highly skewed distribution of individual annual spend. Bottom: the total aggregate volume, highlighting Fresh and Grocery as the largest product segments. Both plots are sorted by the  median spend."}
box_data <- reshape(
  data[, spending.cols],
  direction = "long",
  varying = spending.cols,
  v.names = "Spend",
  timevar = "Category",
  times = spending.cols
)

cat_totals <- data.frame(
  Category = spending.cols,
  Total    = colSums(data[, spending.cols]),
  Median   = apply(data[, spending.cols], 2, median)
)

# option to sort by median vs total; 
# BE note: have chosen median here to highlight the difference in median spend
# despite large total gap between bottom two categories
cat_order <- levels(reorder(cat_totals$Category, cat_totals$Median))
# cat_order <- levels(reorder(cat_totals$Category, cat_totals$Total))
box_data$Category <- factor(box_data$Category, levels = cat_order)
cat_totals$Category <- factor(cat_totals$Category, levels = cat_order)
row.names(box_data) <- NULL


p_box <- ggplot(
  box_data,
  aes(x = Category, y = Spend)
) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::comma,
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "(a) Spending Distribution per Customer",
    y = "Annual Spend (m.u.)",
    x = ""
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

p_bar <- ggplot(cat_totals, aes(x = Category, y = Total)) +
  geom_col(fill = "steelblue", alpha = 0.7, width = 0.7) +
  geom_text(
    aes(label = format(round(Total), big.mark = ",")),
    hjust = -0.1,
    size = 3.5
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "(b) Total Category Spend",
    y = "Total Spend (m.u.)",
    x = ""
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

p_box / p_bar
```

Figure \@ref(fig:boxplots-raw-data) provides a baseline overview of purchasing
behaviour across the six categories. Median spend is relatively modest in all
categories, but every distribution exhibits a long right tail of high-value
customers. At the aggregate level, `r format_var_name("Fresh")` and 
`r format_var_name("Grocery")` dominate total revenue, while 
`r format_var_name("Delicassen")` remains a niche segment. 

```{r histograms-overlay, echo=FALSE, include=TRUE, fig.width=10, fig.height=6, fig.cap="Shape comparison of raw and log-transformed spending distributions by product category. To facilitate direct comparison of the distribution shapes, all spend values were min-max scaled to a 0–1 range within their respective categories. The raw data (blue) demonstrates a right skew across all categories, influenced by extreme outliers. The `log1p` transformation (red) effectively compresses this long tail, yielding more symmetric distributions that ensure variance-based methods (PCA) and distance-based methods (clustering) are not dominated by extreme outliers."}

bind_rows(raw_df, log_df) %>%
  mutate(
    Transform = factor(Transform, levels = c("Raw", "Log-transformed"))
  ) %>%
  group_by(Category, Transform) %>%
  mutate(Spend_scaled = rescale01(Spend)) %>%
  ungroup() %>%
  ggplot(aes(x = Spend_scaled, fill = Transform)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 40,
    colour = "white",
    alpha = 0.5,
    position = "identity"
  ) +
  scale_fill_manual(
    values = c("Raw" = "steelblue", "Log-transformed" = "firebrick")
  ) +
  facet_wrap(~Category, scales = "free_y") +
  labs(
    title = "Shape Comparison: Raw vs Log-transformed Spending",
    x = "Normalised Spend (0–1 scaled within each group)",
    y = "Density"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

This skewness is examined more closely in Figure \@ref(fig:histograms-overlay),
which overlays raw and log-transformed distributiosn for each category. The raw
spending figures exhibit a strong rightward skew across all six product
categories, indicating that while most customers spend relatively small amounts,
a long tail of high-spending customers dominates the scale. Applying a log
transformation (`log1p`) successfully 'pulls in' these extreme values, centring
the bulk of the data and producing much more symmetric distributions.


```{r corrplots-side-by-side, echo=knitr::is_html_output(), include=TRUE, fig.width=12, fig.height=5.5, fig.cap="Pairwise correlation matrices of the six product categories, comparing linear relationships using raw spending data (left) versus log-transformed spending data (right)."}
par(mfrow = c(1, 2))

soft_palette <- colorRampPalette(c("#D73027", "#F46D43", "#FFFFFF", "#74ADD1", "#4575B4"))(200)

corrplot(
  cor(spending),
  method      = "color",
  type        = "upper",
  col         = soft_palette,     
  addCoef.col = "black",          
  number.cex  = 0.8,
  tl.col      = "black",
  title       = "Raw spending",
  mar         = c(0, 0, 2, 0)
)

corrplot(
  cor(spending_log),
  method      = "color",
  type        = "upper",
  col         = soft_palette, 
  addCoef.col = "black",        
  number.cex  = 0.8,
  tl.col      = "black",
  title       = "Log-transformed spending",
  mar         = c(0, 0, 2, 0)
)
```

The side-by-side correlation plots reveal the underlying purchasing
relationships between product categories before and after adjusting for scale.
Both matrices highlight a distinct cluster of highly correlated spending:
`r format_var_name("Grocery")`, `r format_var_name("Detergents_Paper")`, and
`r format_var_name("Milk")` show strong positive linear relationships,
suggesting they are frequently purchased together in bulk. While the log
transformation preserves this core structure, it slightly softens the most
extreme correlations (for instance, the correlation between
`r format_var_name("Grocery")` and `r format_var_name("Detergents_Paper")`
drops from `r formatC(cor(spending$Grocery, spending$Detergents_Paper), format = "f", digits = 2)`
to `r formatC(cor(spending_log$Grocery, spending_log$Detergents_Paper), format = "f", digits = 2)`).
This indicates that in the raw data, a few extreme outlier purchases were
artificially inflating the strength of that linear relationship.

## Multi-dimensional scaling {#apd:suppl-mds}

```{r mds-side-by-side, echo=FALSE, include=TRUE, fig.width=12, fig.height=5.5, fig.cap="Non-metric MDS using Gower’s distance across all eight variables (six log-transformed continuous spending categories plus Channel and Region). Points are coloured by Channel (left) and Region (right). Unlike PCA, Gower’s distance forces distinct spatial separation based on categorical mismatches, resulting in six artificial clusters corresponding to the cross-tabulation of the two Channels and three Regions."}

# Gower distance on all 8 variables
mds_data <- data.frame(
  Channel = factor(data$Channel, labels = c("HoReCa", "Retail")),
  Region = factor(data$Region, labels = c("Lisbon", "Oporto", "Other")),
  spending_log
)

gow <- cluster::daisy(mds_data, metric = "gower")
mds_fit <- MASS::isoMDS(gow, k = 2, trace = FALSE)
mds_pts <- as.data.frame(mds_fit$points)
colnames(mds_pts) <- c("MDS1", "MDS2")

mds_pts$Channel <- mds_data$Channel
mds_pts$Region <- mds_data$Region

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Left: coloured by Channel
plot(
  mds_pts$MDS1,
  mds_pts$MDS2,
  col = ifelse(mds_pts$Channel == "HoReCa", "#E64B35", "#000000"),
  # 19 = Circle, 17 = Triangle
  pch = ifelse(mds_pts$Channel == "HoReCa", 19, 17),
  cex = 0.8,
  xlab = "MDS1",
  ylab = "MDS2",
  main = "(a) Coloured by Channel"
)
legend(
  "topleft",
  legend = c("HoReCa", "Retail"),
  col = c("#E64B35", "#000000"),
  pch = c(19, 17),
  cex = 0.9,
  bty = "n"
)

# Right: coloured by Region
region_cols <- c(Lisbon = "#E69F00", Oporto = "#56B4E9", Other = "#CC79A7")
plot(
  mds_pts$MDS1,
  mds_pts$MDS2,
  col = region_cols[mds_pts$Region],
  pch = 19,
  cex = 0.8,
  xlab = "MDS1",
  ylab = "MDS2",
  main = "(b) Coloured by Region"
)
legend(
  "topleft",
  legend = names(region_cols),
  col = region_cols,
  pch = 19,
  cex = 0.9,
  bty = "n"
)
# mds_fit$stress
```

Figure \@ref(fig:mds-side-by-side) presents a non-metric MDS solution computed
from Gower's distance across all eight variables, incorporating the categorical
`r format_var_name("Channel")` and `r format_var_name("Region")` columns excluded
from the PCA. Because Gower's metric treats any categorical mismatch as a
maximal distance contribution on that variable, the resulting spatial map is
dominated by these labels rather than underlying spending behaviour.

Panel (a) shows separation by `r format_var_name("Channel")` along the primary
axis (MDS1), while Panel (b) reveals further stratification by 
`r format_var_name("Region")` along MDS2. The result is a forced stratification
into six spatial clusters reflecting the Channel $\times$ Region
cross-tabulation. While there is minor regional intermingling (e.g., two Oporto
customers mapping near the Lisbon cluster), the overall spatial map merely
reflects the cross-tabulation of the categorical labels rather than organic,
dataset-wide purchasing patterns. The solution's stress value of
`r formatC(mds_fit$stress, format = "fg", digits = 3)` further indicates a 
mediocre fit, falling above the conventional threshold of 10 for reliable
low-dimensional representation. This confirms that the PCA approach, relying
purely on continuous spending variables, is better suited to uncovering natural
commercial structure in this dataset.

```{=latex}
\newpage
```

## Clustering diagnostics {#apd:clustering-diagnostics}

### Optimal cluster selection {#apd:cluster-selection}
```{r silhouette-plot, echo=FALSE, fig.width=10, fig.height=4, fig.cap="Silhouette analysis for K-Means clustering (k=2 to 10). The optimal number of clusters is k=2 followed by k=3 as a strong second. Beyond k=3, scores plateau around 0.19."}
fviz_nbclust(scaled_data, kmeans, method = 'silhouette', nstart = 25)
```

```{r bic-plot, echo=FALSE, fig.width=10, fig.height=5, fig.cap="BIC values across component counts for all model types evaluated by `mclust`. Higher (less negative) BIC indicates better fit. The BIC peaks at 5 components under the VVE (ellipsoidal, variable volume and shape, equal orientation) model."}
plot(
  gmm_spending,
  what = "BIC",
  legendArgs = list(x = "bottomright", ncol = 4, cex = 0.8)
)
```

Silhouette analysis (Figure \@ref(fig:silhouette-plot)) identifies k=2 as
optimal (`r formatC(sil_scores["2"], format = "f", digits = 3)`), with k=3 as a
close second (`r formatC(sil_scores["3"], format = "f", digits = 3)`). Beyond
k=3, scores plateau around 0.19.

The GMM's BIC (Figure \@ref(fig:bic-plot)) peaks at
`r gmm_spending$G` components
(`r formatC(gmm_spending$bic, format = "f", digits = 1, big.mark = ",")`), 
compared to
`r formatC(best_per_k["4"], format = "f", digits = 1, big.mark = ",")` 
for four and
`r formatC(best_per_k["6"], format = "f", digits = 1, big.mark = ",")` for six. 
Interestingly k=4 is worse than k=3, suggesting the data supports either 3 or 5
components but not an intermediate value. Both methods agree that k=3 is a
meaningful threshold and the disagreement on whether to proceed to k=5 reflects 
a genuine trade-off between parsimony and granularity.


### Channel and Region cross-tabulations {#apd:cluster-crosstabs}

```{r km3-channel-crosstab, echo=FALSE}
km3_3way <- table(
  KM = km3$cluster,
  Channel = data$Channel,
  Region = data$Region
)

km_rows <- list()
for (k in 1:3) {
  for (ch in 1:2) {
    km_rows[[length(km_rows) + 1]] <- data.frame(
      KM = k,
      Channel = ch,
      Lisbon = km3_3way[k, ch, 1],
      Oporto = km3_3way[k, ch, 2],
      Other = km3_3way[k, ch, 3]
    )
  }
}
km_region_df <- do.call(rbind, km_rows)

# Build pack_rows labels programmatically
km3_pack_labels <- sapply(1:3, function(k) {
  paste0("KM-", k, " (", km3_names[as.character(k)], ")")
})

kable(
  km_region_df[, c("Channel", "Lisbon", "Oporto", "Other")],
  booktabs = TRUE,
  row.names = FALSE,
  caption = "K-Means (k=3) cluster assignments vs \\ Channel (1=HoReCa, 2=Retail) and Region.",
  align = c("l", "r", "r", "r")
) %>%
  kable_styling(latex_options = "hold_position", font_size = 10) %>%
  pack_rows(km3_pack_labels[1], 1, 2, bold = TRUE, hline_before = FALSE) %>%
  pack_rows(km3_pack_labels[2], 3, 4, bold = TRUE, hline_before = TRUE) %>%
  pack_rows(km3_pack_labels[3], 5, 6, bold = TRUE, hline_before = TRUE)
```

```{r gmm-region-crosstab, echo=FALSE}
gmm_3way <- table(
  GMM = data$gmm_cluster,
  Channel = data$Channel,
  Region = data$Region
)

rows_list <- list()
for (g in 1:5) {
  for (ch in 1:2) {
    rows_list[[length(rows_list) + 1]] <- data.frame(
      GMM = g,
      Channel = ch,
      Lisbon = gmm_3way[g, ch, 1],
      Oporto = gmm_3way[g, ch, 2],
      Other = gmm_3way[g, ch, 3]
    )
  }
}
region_df <- do.call(rbind, rows_list)

gmm_pack_labels <- sapply(1:5, function(g) {
  paste0("GMM-", g, " (", gmm_names[as.character(g)], ")")
})

kable(
  region_df[, c("Channel", "Lisbon", "Oporto", "Other")],
  booktabs = TRUE,
  row.names = FALSE,
  caption = "GMM (k=5) cluster assignments vs \\ Channel (1=HoReCa, 2=Retail) and Region.",
  align = c("l", "r", "r", "r")
) %>%
  kable_styling(latex_options = "hold_position", font_size = 10) %>%
  pack_rows(gmm_pack_labels[1], 1, 2, bold = TRUE, hline_before = FALSE) %>%
  pack_rows(gmm_pack_labels[2], 3, 4, bold = TRUE, hline_before = TRUE) %>%
  pack_rows(gmm_pack_labels[3], 5, 6, bold = TRUE, hline_before = TRUE) %>%
  pack_rows(gmm_pack_labels[4], 7, 8, bold = TRUE, hline_before = TRUE) %>%
  pack_rows(gmm_pack_labels[5], 9, 10, bold = TRUE, hline_before = TRUE)
```


Table \@ref(tab:km3-channel-crosstab) shows that K-Means (k=3) produces one
cluster that is almost exclusively HoReCa 
(KM-`r horeca_cluster`, "`r km3_names[horeca_cluster]`": 
`r km3_ch_tab[as.integer(horeca_cluster), 1]` of 
`r sum(km3$cluster == as.integer(horeca_cluster))` customers), one that leans 
Retail (KM-`r retail_cluster`, "`r km3_names[retail_cluster]`": 
`r km3_ch_tab[as.integer(retail_cluster), 2]` of 
`r sum(km3$cluster == as.integer(retail_cluster))`), and a third 
(KM-`r mixed_cluster`, "`r km3_names[mixed_cluster]`") that draws substantially
from both channels 
(`r km3_ch_tab[as.integer(mixed_cluster), 1]` HoReCa, 
`r km3_ch_tab[as.integer(mixed_cluster), 2]` Retail). The "Retail Dominant" 
label reflects the spending profile rather than exclusive Channel membership: 
these customers concentrate on `r format_var_name("Grocery")` and 
`r format_var_name("Detergents_Paper")` regardless of their administrative 
Channel label. 

```{=latex}
\newpage
```

### Cluster spending profiles {#apd:cluster-profiles}

```{r q2-k-means-table-1, echo=FALSE, include=FALSE}
# ---- K-Means cluster profiles (k=2 and k=3) ----
km2 <- KMeans.results[[2]]
km3 <- KMeans.results[[3]]

data$km2_cluster <- km2$cluster
data$km3_cluster <- km3$cluster

# Median spending per cluster
km3_profiles <- data %>%
  group_by(km3_cluster) %>%
  summarise(
    n = n(),
    across(all_of(spending.cols), median),
    .groups = "drop"
  )
print(km3_profiles)

# Channel cross-tab
table(KMeans_3 = km3$cluster, Channel = data$Channel)

# ---- Percentage views for both methods ----

# K-Means k=3: % of each cluster's total spend per category
km3_pct <- data %>%
  group_by(km3_cluster) %>%
  summarise(across(all_of(spending.cols), median), .groups = "drop") %>%
  mutate(
    total = Fresh + Milk + Grocery + Frozen + Detergents_Paper + Delicassen,
    across(all_of(spending.cols), ~ round(. / total * 100, 1))
  ) %>%
  select(-total)
print(km3_pct)

# GMM k=5: same treatment
gmm_pct <- data %>%
  group_by(gmm_cluster) %>%
  summarise(across(all_of(spending.cols), median), .groups = "drop") %>%
  mutate(
    total = Fresh + Milk + Grocery + Frozen + Detergents_Paper + Delicassen,
    across(all_of(spending.cols), ~ round(. / total * 100, 1))
  ) %>%
  select(-total)
print(gmm_pct)

# K-Means (k=3) spending profiles (% of median basket)
#   KM1 (n=147): Fresh 33%, Grocery 27%, Milk 20%  ← mixed basket, 56H/91R
#   KM2 (n=212): Fresh 58%, Frozen 13%             ← HoReCa dominant (209H/3R)
#   KM3 (n=81):  Grocery 45%, Milk 27%, Det 18%    ← Retail dominant (33H/48R)
#
# KM2 is a clean HoReCa cluster. KM3 is a clean Retail cluster.
# KM1 is the interesting one - the "middle ground" that mixes both channels.
# This is arguably the k=3 third segment: customers whose spending doesn't
# fit neatly into either profile.
#
# GMM (k=5) spending profiles (% of median basket)
#   GMM1 (n=51):  Grocery 44%, Det 22%, Milk 26%   ← heavy Retail (supermarkets)
#   GMM2 (n=101): Grocery 33%, Fresh 23%, Milk 21% ← moderate Retail (diversified)
#   GMM3 (n=155): Fresh 62%, Frozen 15%            ← high-volume HoReCa
#   GMM4 (n=81):  Fresh 41%, Milk 22%, Grocery 22% ← balanced HoReCa
#   GMM5 (n=52):  Grocery 42%, Fresh 26%           ← small/mixed (lowest spend)
#
# GMM splits Retail into concentrated (GMM1) vs diversified (GMM2)
# GMM splits HoReCa into Fresh-dominated (GMM3) vs balanced (GMM4)
# GMM5 is the low-spend boundary group - similar to KM1's "middle ground"
```

```{r q2-tables-3-apd, echo=FALSE}
# ---- Raw median spending table ----
km3_raw_table <- data %>%
  group_by(km3_cluster) %>%
  summarise(
    n = n(),
    across(all_of(spending.cols), median),
    .groups = "drop"
  ) %>%
  mutate(
    Method = "K-Means",
    Cluster = paste0("KM-", km3_cluster)
  ) %>%
  select(-km3_cluster)

gmm_raw_table <- data %>%
  group_by(gmm_cluster) %>%
  summarise(
    n = n(),
    across(all_of(spending.cols), median),
    .groups = "drop"
  ) %>%
  mutate(
    Method = "GMM",
    Cluster = paste0("GMM-", gmm_cluster)
  ) %>%
  select(-gmm_cluster)

combined_raw <- bind_rows(km3_raw_table, gmm_raw_table) %>%
  select(Cluster, everything(), -Method) %>%
  mutate(across(all_of(spending.cols), ~ format(round(.), big.mark = ",")))

kable(
  combined_raw,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  col.names = c(
    "Cluster",
    "$n$",
    "Fresh",
    "Milk",
    "Grocery",
    "Frozen",
    "Detergents\\_Paper",
    "Delicassen"
  ),
  align = c("l", "r", rep("r", 6)),
  caption = "Cluster spending profiles (median annual spending, m.u.)",
  linesep = ""
) %>%
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 10
  ) %>%
  pack_rows("K-Means (k=3)", 1, 3, bold = TRUE, hline_before = FALSE) %>%
  pack_rows(
    "Gaussian Mixture Model (k = 5)",
    4,
    8,
    bold = TRUE,
    hline_before = TRUE
  )
```
```{r q2-tables-2-apd, echo=FALSE}
# ---- Combined table for report ----
km3_for_table <- km3_pct %>%
  mutate(Method = "K-Means", Cluster = paste0("KM-", km3_cluster)) %>%
  select(-km3_cluster)

gmm_for_table <- gmm_pct %>%
  mutate(Method = "GMM", Cluster = paste0("GMM-", gmm_cluster)) %>%
  select(-gmm_cluster)

combined <- bind_rows(km3_for_table, gmm_for_table) %>%
  select(Cluster, everything(), -Method)

kable(
  combined,
  booktabs = TRUE,
  escape = FALSE,
  col.names = c(
    "Cluster",
    "Fresh",
    "Milk",
    "Grocery",
    "Frozen",
    "Detergents\\_Paper",
    "Delicassen"
  ),
  align = c("l", rep("r", 6)),
  caption = "Cluster spending profiles (\\% of median basket)",
  linesep = ""
) %>%
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 10
  ) %>%
  pack_rows("K-Means (k = 3)", 1, 3, bold = TRUE, hline_before = FALSE) %>%
  pack_rows(
    "Gaussian Mixture Model (k = 5)",
    4,
    8,
    bold = TRUE,
    hline_before = TRUE
  )
```
```{r q2-tables-4-apd, echo=FALSE}
# each cluster's share of total category spend
# i.e., what % goes to each segment we have now identified

km3_totals <- data %>%
  group_by(km3_cluster) %>%
  summarise(
    n = n(),
    across(all_of(spending.cols), sum),
    .groups = "drop"
  )

km3_col_pct <- km3_totals %>%
  mutate(
    across(
      all_of(spending.cols),
      ~ round(. / sum(.) * 100, 1)
    ),
    Cluster = paste0("KM-", km3_cluster)
  ) %>%
  select(Cluster, n, all_of(spending.cols))

gmm_totals <- data %>%
  group_by(gmm_cluster) %>%
  summarise(
    n = n(),
    across(all_of(spending.cols), sum),
    .groups = "drop"
  )

gmm_col_pct <- gmm_totals %>%
  mutate(
    across(
      all_of(spending.cols),
      ~ round(. / sum(.) * 100, 1)
    ),
    Cluster = paste0("GMM-", gmm_cluster)
  ) %>%
  select(Cluster, n, all_of(spending.cols))

combined_col_pct <- bind_rows(km3_col_pct, gmm_col_pct)

kable(
  combined_col_pct,
  format    = "latex",
  booktabs  = TRUE,
  escape    = FALSE,
  col.names = c("Cluster", "$n$", "Fresh", "Milk", "Grocery",
                "Frozen", "Detergents\\_Paper", "Delicassen"),
  align     = c("l", "r", rep("r", 6)),
  caption   = "Revenue concentration: each cluster's share (\\%) of total
               category spend across all 440 customers.",
  linesep   = ""
) %>%
  kable_styling(
    latex_options = c("hold_position"),
    font_size     = 10
  ) %>%
  pack_rows("K-Means (k = 3)", 1, 3, bold = TRUE, hline_before = FALSE) %>%
  pack_rows("Gaussian Mixture Model (k = 5)", 4, 8, bold = TRUE, hline_before = TRUE)

```


Tables \@ref(tab:q2-tables-2-apd)–\@ref(tab:q2-tables-4-apd) present the
cluster spending profiles from three perspectives. 

1. Table \@ref(tab:q2-tables-2-apd) shows the percentage composition of each
cluster's median basket examining how each segment allocates its spending across
categories.
1. Table \@ref(tab:q2-tables-3-apd) reports the raw median annual spending
(monetary units [m.u.]) per category, revealing the absolute scale differences
between segments. 
1. Table \@ref(tab:q2-tables-4-apd) reports revenue concentration and
corresponds directly to the stacked bars in Figure
\@ref(fig:q2-revenue-triptych). Note that because Table
\@ref(tab:q2-tables-3-apd) reports the **median** annual spending, it is not
directly linked to Table \@ref(tab:q2-tables-4-apd).

The K-Means "`r km3_names[horeca_cluster]`" cluster 
(KM-`r horeca_cluster`) allocates `r km3_stat(horeca_cluster, "Fresh_pct")`% of 
its median basket 
(Table \@ref(tab:q2-tables-2-apd)) to `r format_var_name("Fresh")` and accounts 
for `r km3_rev(horeca_cluster, "Fresh_rev_pct")`% of total Fresh revenue 
(Table \@ref(tab:q2-tables-4-apd)), 
while "`r km3_names[retail_cluster]`" (KM-`r retail_cluster`) concentrates on 
`r format_var_name("Grocery")` 
(`r km3_stat(retail_cluster, "Grocery_pct")`%) and 
`r format_var_name("Detergents_Paper")` 
(`r km3_stat(retail_cluster, "Detergents_Paper_pct")`%) but comprises only 
`r sum(km3$cluster == as.integer(retail_cluster))` customers 
(`r round(sum(km3$cluster == as.integer(retail_cluster)) / nrow(data) * 100)`% 
of the base). At the GMM level, the distinction sharpens further: the median 
GMM-3 customer ("`r gmm_names["3"]`") dedicates `r gmm_stat(3, "Fresh_pct")`% 
of its basket 
to `r format_var_name("Fresh")` and, despite representing 
`r gmm_pct_customers(3)`% of customers, captures 
`r gmm_rev(3, "Fresh_rev_pct")`% of total Fresh revenue. The Retail split is 
equally informative: the typical GMM-1 customer ("`r gmm_names["1"]`") 
concentrates on `r format_var_name("Grocery")` (`r gmm_stat(1, "Grocery_pct")`%) 
and `r format_var_name("Detergents_Paper")` 
(`r gmm_stat(1, "Detergents_Paper_pct")`%), while the typical GMM-2 customer 
("`r gmm_names["2"]`") spreads spend more evenly across categories including 
`r format_var_name("Fresh")` (`r gmm_stat(2, "Fresh_pct")`%), suggesting a 
convenience-store or diversified retailer profile.


### Cluster assignments in PCA space {#apd:cluster-pca}

```{r q2-plots-apd, echo=FALSE, fig.width=10, fig.height=8, fig.cap="Cluster assignments from K-Means (k=3, top) and GMM (k=5, bottom) projected onto the first two principal components, which together capture 71.3 percent of total variance. All clustering was performed in the full six-dimensional standardised log-spending space. Apparent overlap between clusters may reflect the loss of separating information in the remaining 28.7 percent of variance captured by PC3--PC6, rather than genuine ambiguity in cluster assignment."}

pca_scores <- as.data.frame(pca_result$x[, 1:2])
colnames(pca_scores) <- c("PC1", "PC2")

pc1_var <- summary(pca_result)$importance[2, 1] * 100
pc2_var <- summary(pca_result)$importance[2, 2] * 100
pc1_lab <- sprintf("PC1 (%.1f%%)", pc1_var)
pc2_lab <- sprintf("PC2 (%.1f%%)", pc2_var)

pca_scores$KM3  <- factor(km3$cluster)
pca_scores$GMM5 <- factor(data$gmm_cluster)

p_km <- ggplot(pca_scores, aes(x = PC1, y = PC2, colour = KM3, shape = KM3)) +
  geom_point(alpha = 0.7, size = 1.5) +
  scale_colour_brewer(palette = "Set1") +
  labs(title = "(a) K-Means (k = 3)", x = pc1_lab, y = pc2_lab,
       colour = "Cluster", shape = "Cluster") +
  theme_minimal() +
  theme(legend.position = "right")

p_gmm_pca <- ggplot(pca_scores, aes(x = PC1, y = PC2, colour = GMM5, shape = GMM5)) +
  geom_point(alpha = 0.9, size = 1.5) +
  scale_colour_brewer(palette = "Set2") +
  scale_shape_manual(values = c(16, 17, 15, 18, 3)) +
  labs(title = "(b) GMM (k = 5)", x = pc1_lab, y = pc2_lab,
       colour = "Cluster", shape = "Cluster") +
  theme_minimal() +
  theme(legend.position = "right")

p_km / p_gmm_pca
```

Figure \@ref(fig:q2-plots-apd) projects the K-Means and GMM cluster assignments
onto PCA space. Both panels show that the dominant cluster boundaries divide the
data along PC1 (the "channel axis" identified in \@ref(qone)), confirming that
the clustering and PCA analyses converge on the same dominant structure. The GMM
panel shows finer subdivision, particularly along PC2, consistent with GMM's
ability to model ellipsoidal clusters of varying density.

## Ward's hierarchical clustering {#apd:wards-clust}

```{r apd-wards-cluster, echo=FALSE, message=FALSE, warning=FALSE}
# Ward's hierarchical clustering
# this code was moved up to Q2 (commented out here for transparency)
# dist_mat <- dist(scaled_data, method = "euclidean")
# hc <- hclust(dist_mat, method = "ward.D2")
# hc_clusters <- cutree(hc, k = 3)

# table(KMeans = KMeans.results[[3]]$cluster, Ward = hc_clusters)
# cross-tabulation
cross_tab <- table(KMeans.results[[3]]$cluster, Ward = hc_clusters)

# Ename the rows and columns 
rownames(cross_tab) <- paste("K-Means", 1:3)
colnames(cross_tab) <- paste("Ward", 1:3)

# format
kable(
  cross_tab, 
  booktabs = TRUE, 
  caption = "K-Means vs Ward's hierarchical clustering assignments (k=3)."
) %>%
  kable_styling(latex_options = "hold_position") 
```

Table \@ref(tab:apd-wards-cluster) cross-tabulates K-Means (k=3) and Ward's
hierarchical clustering (k=3) assignments. The largest cluster shows
near-perfect agreement (201 of 212 observations). Disagreement is concentrated
among boundary customers between the two smaller segments, reflecting the
different optimisation strategies. K-Means minimises within-cluster sum of
squares globally, while Ward's method does so greedily through successive
agglomeration. 
<!-- This partial agreement supports the robustness of the dominant -->
<!-- segmentation structure while highlighting that finer sub-group boundaries are -->
<!-- method-dependent. -->

<!-- Ward's hierarchical clustering (k=3) agrees with K-Means on  -->
<!-- `r max(sum(KMeans.results[[3]]$cluster == hc_clusters), sum(KMeans.results[[3]]$cluster == (4 - hc_clusters)))`  -->
<!-- of 440 assignments (Appendix, Table @ref(tab:ward-crosstab)), providing further evidence that the three-segment structure is robust across algorithmic choices. -->



```{=latex}
\clearpage
\newpage
```

# Prompts and LLM outputs {#apd-llm-information}

In this appendix, we opted to use a consistent visual format to distinguish
between prompts sent to the LLM, its responses, and our own commentary. 

::: {.promptbox}

**Prompts to the LLM** - shown with a right-hand border. These are the instructions and
questions we provided to the model.

:::

:::: {.responsebox}

**LLM Response** - shown with a left-hand border, forming a conversational
flow with the prompt. Responses may include a mix of explanation and code:
```r
summary(mtcars$mpg)
```

::::

::: {.insightbox}

**Insight and critical interpretation** - highlighted observations or key findings drawn from the output.

:::

::: {.alertbox}

**Alert** - flags potential concerns, limitations, or caveats with the
generated output.

:::


## \@ref(qone) PCA LLM dialogue

_Example format below (literally asked base ChatGPT to come up with it's own
prompt and then answer it - change when we do for real)._

<!-- prettier-ignore -->
::: promptbox 
**LLM Prompt:** Act as an expert data scientist. I have the
Wholesale Customers dataset from Kaggle. Please write an R script to perform
Principal Component Analysis (PCA) on this dataset. Provide the code, generate
the necessary plots, and summarise the key findings. 
:::

<!-- prettier-ignore -->
:::: {.responsebox}

Here is the LLM's generated code:
```r
# =========================
# 0. Setup
# =========================
rm(list = ls())
packages <- c("tidyverse", "factoextra", "corrplot")
installed <- packages %in% rownames(installed.packages())
if(any(!installed)) install.packages(packages[!installed])
```

Which handles string formatting cleanly.

::::

<!-- prettier-ignore -->
::: insightbox 
**Our interpretation:** The PCA biplot shows that Fresh and Milk
spending are nearly orthogonal, suggesting they are independent purchasing
behaviors. 
:::

<!-- prettier-ignore -->
::: alertbox 
**Our concern:** PCA assumes linear relationships between
variables. If the underlying structure is non-linear, consider using t-SNE or
UMAP instead. Also note that the results are sensitive to outliers - always
check for extreme values before interpreting the components. 
:::


## \@ref(qtwo) Clustering LLM dialogue

```{=latex}
\clearpage
\newpage
```


# Reproducibility, accessibility & declarations

<!-- BE note: taken out as we are submitting R file separately -->
<!-- Use: `knitr::purl("cs1-group07.rmd", documentation = 0)` to generate r file -->
<!-- Use: `tools::showNonASCIIfile("cs1-group07.rmd")` to debug -->
<!-- ## R Code Documentation -->
<!-- ```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
``` -->

## Reproducibility & accessibility

An accessible HTML version of this report is available via a public GitHub page:

- https://ytterbiu.github.io/SMM636-ML-group-coursework-03-g01/

This report was created in R Markdown. The source code is open source and is
available via:

- https://github.com/ytterbiu/SMM636-ML-group-coursework-03-g01

Changes made to this document were tracked using Git and are also available via
the same repository
(https://github.com/ytterbiu/SMM636-ML-group-coursework-03-g01/commits/main/).
Please note that this commit history reflects updates made to the project
repository and is not a comprehensive timeline of all contributions by group
members.

## Word count

The total word count for the core sections of this report is **`r word_count`**
words (`r ifelse(word_count > 1000, "above", "below")` the 1000 word limit).

The calculation was performed programmatically directly from the source `.rmd`
file. An **R** script was used to read the raw document lines and strip out
non-text elements, including the YAML front matter, R code chunks, sub-headers,
markdown formatting (tables and figures), and LaTeX/HTML syntax. The script
isolates lines falling within `#intro`, `#qone`, `#qtwo`, and `#qthr`. It then 
calculates the final tally using the `stringi` package.

The code chunk that performs this calculation is executed silently at the
beginning of this document, but is reproduced below for transparency:

```{r display-word-count-code, ref.label="word-count", eval=FALSE, echo=TRUE}
```

## Generative AI Declaration

- Will be best to ask about this during the lecture - will almost want this for
  the first part - or is the first part completely AI free?

<!-- Generative AI (GenAI) tools were used throughout this project to assist with -->
<!-- project planning, methodological validation, and debugging. In practice, these -->
<!-- tools functioned as an additional technical team member, providing immediate -->
<!-- feedback, troubleshooting code, and challenging statistical assumptions. -->

<!-- ### Tools Used -->

<!-- Google Gemini 3 Pro (Deep Research), ChatGPT 5.2 -->

<!-- ### Prompts used -->

<!-- The GenAI tools were provided with the 'coursework_CM2_250120.pdf' file and -->
<!-- extracts from our `R` code. -->

<!-- A non-exhaustive list of prompts used is shown below. We opted to include at -->
<!-- least one example of each _type_ of prompt used, rather than a list of all -->
<!-- prompts. For example, if we entered the prompts 'How can we put two ggplots side -->
<!-- by side?' and 'How to remove grey background on a ggplot', we would include only -->
<!-- the first of these in the list below, as both relate to plotting issues (& are -->
<!-- similar to 'querying' with a traditional search engine). -->

<!-- - "...the top of both graphs in r are slightly different. (Pasting R code.) Can -->
<!--   this be fixed for the plot area?" -->
<!-- - "Help to identify the key points here" -->
<!-- - "How can this error be tracked down in an rmd file? ! Text line contains an -->
<!--   invalid character. l.1 ^^@^^@..." -->

<!-- ### Use of GenAI outputs & subsequent revisions -->

<!-- The GenAI tools served as a "consultant" member of the group. Notably, the AI -->
<!-- output: -->

<!-- - Helped debug plotting issues. -->
<!-- - Commented on the structure of R code snippets and the report structure. -->

<!-- ### Changes made: -->

<!-- All AI suggestions were critically reviewed and debated by the group before -->
<!-- implementation. For example: -->

<!-- - While the AI suggested code for complex layouts, the specific aesthetic -->
<!--   choices and annotations were manually adjusted to ensure clarity. -->
<!-- - Where the AI proposed specific interpretations, these were cross-referenced -->
<!--   with academic literature and statistics textbooks. -->
<!-- - The final R code for implementation was written and debugged by the group with -->
<!--   AI snippets being treated as drafts rather than final solutions. -->

<!-- While GenAI was treated as a collaborator, the final analysis, statistical -->
<!-- interpretation, and the written content of this report were generated and -->
<!-- verified by the (human) group members. -->

---

# References {-}