---
title:
  "Beyond the Binary: Uncovering a Hidden Customer Segment in Wholesale Purchasing Data"
subtitle: "SMM068 (2025-26) Coursework 03 - Group 01"
author:
  - "Abdulrahman Alolyan"
  - "Benjamin Evans"
  - "Basmah Khan"
  - "Ardi Wira Sudarmo"
date: "`r format(Sys.Date(), '%d %B %Y')`"
papersize: a4
fontsize: 12pt
output:
  bookdown::html_document2:
    self_contained: true
    css: style.css
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: paper
    highlight: pygments
    df_print: paged
    pandoc_args: ["--lua-filter=custom-boxes.lua"]
  # runtime: shiny
  bookdown::pdf_document2:
    latex_engine: lualatex
    includes:
      in_header: preamble.tex
    pandoc_args: ["--lua-filter=custom-boxes.lua"]
    number_sections: true
    toc: true
    # citation_package: natbib
    keep_tex: true
bibliography: references.bib
csl: ieee.csl
---

```{=latex}
\newpage
```

```{=html}
<h1>Information</h1>

This R Markdown document was created as part of a group assignment for SMM636 at Bayes Business School, City St George's, University of London in Term 2 2025-26.

```

```{r asthetic-header, include=FALSE, echo=FALSE}
# ============================================================================ #
# Key Information ====
# SMM636 Machine Learning
# Group Coursework 2025-26
# Group:        Group 01
# Authors (in alphabetical order):
#   - Abdulrahman Alolyan
#   - Benjamin Evans
#   - Basmah Khan
#   - Bong Su Kil
#   - Ardi Wira Sudarmo
# Professor:    Dr. Rui Zhu
# Institution:  Bayes Business School - City St George's, University of London
# Date:         03 Apr 2026
#
# Description:  Term 2 group project for SMM636 Machine Learning
#               (25% of coursework grade - 25% of module grade). The R code
#               below has been exported directly from an R Markdown (.rmd) file.
#               Hence the knitr settings.
#
# Dependencies:
#   - tidyverse: loads dplyr, tidyr, ggplot2, purrr (& others)
#   - TBI
#
# Acknowledgements
#   - TBI
# ============================================================================ #
```

```{r setup-knitr, include=FALSE, echo=FALSE}
# Setup & configuration ====
#----------------------- Initial setup (knitr settings) -----------------------#
dir.create("fig", showWarnings = FALSE)

# Defaults common to all outputs
knitr::opts_chunk$set(
  echo     = TRUE,
  message  = FALSE,
  warning  = FALSE,
  fig.align = "center",
  out.width = "100%",
  fig.path  = "fig/",
  dpi       = 300
)

# Output-specific settings
if (knitr::is_latex_output()) {
  knitr::opts_chunk$set(
    fig.width = 6,
    fig.height = 4,
    dev = "pdf",
    fig.pos = "!ht",
    out.extra = ""
  )
} else {
  knitr::opts_chunk$set(
    fig.width = 6,
    fig.height = 4,
    dev = "svglite"  # or "png"
  )
}
```

```{r setup-qol, include=FALSE, echo=FALSE}
#----------------------------- Clean environment ------------------------------#
rm(list = ls()) # Remove all objects
graphics.off() # Close all graphical devices
cat("\014") # Clean console
```

```{r load-dependencies, include=FALSE, echo=FALSE}
#------------------- Load dependencies / external libraries -------------------#
# --- Data Manipulation & Plotting ---
library(tidyverse) # loads dplyr, tidyr, ggplot2, purrr (& others)
library(patchwork) # for combining multiple plots into one figure

# --- Reporting & Tables ---
library(kableExtra) # for general tables

library(factoextra)   # PCA & clustering visualisation
library(cluster)      # silhouette, pam
library(NbClust)      # optimal-k diagnostics
library(corrplot)     # correlation heatmap
library(gridExtra)    # arranging plots
library(GGally)       # pairs plot
```

```{r custom-functions, include=FALSE, echo=FALSE}
#---------------------------- Custom QOL functions ----------------------------#
#####################################
# function: banner comments (used to to section up code)
# Usage: banner_comment("Element 1: data cleaning") -> then ctrl + v (or cmd+v)
#####################################
banner_comment <- function(text, width = 80, border = "#", fill = "-") {
  txt <- paste0(" ", text, " ")
  inner_width <- width - 2 * nchar(border)
  banner_string <- ""

  if (inner_width <= nchar(txt)) {
    banner_string <- paste0(border, txt, border)
  } else {
    pad_total <- inner_width - nchar(txt)
    pad_left <- pad_total %/% 2
    pad_right <- pad_total - pad_left

    banner_string <- paste0(
      border,
      strrep(fill, pad_left),
      txt,
      strrep(fill, pad_right),
      border
    )
  }

  cat(banner_string, "\n")
  # copy banner to allow direct pasting (requires clipr)
  clipr::write_clip(banner_string)
  # avoid [1] when printing if want to manually copy
  invisible(banner_string)
}
#####################################
# function: format p-values for text
# Usage (in-line): `r format_p_vals(ad_test_result$p.value)`
# Usage (console): format_p_vals(ad_test_result$p.value)
#####################################
format_p_vals <- function(p) {
  if (length(p) != 1L || is.na(p)) {
    stop("Error! p must be a single non-missing value")
  }
  if (p > 1) {
    stop("Error! Value greater than 1")
  }
  if (p < 0) {
    stop("Error! Value less than 0")
  }

  if (p >= 0.01) {
    paste0("= ", formatC(p, format = "f", digits = 2))
  } else if (p >= 0.001) {
    paste0("= ", formatC(p, format = "f", digits = 3))
  } else {
    "< 0.001"
  }
}
#####################################
# function: format confidence intervals for tables & text
# Usage (in-line): `r format_interval(el2_ci_normal_95[1], el2_ci_normal_95[2])`
# Usage (console): format_interval(el2_ci_normal_95[1], el2_ci_normal_95[2])
#####################################
format_interval <- function(lower, upper, digits = 3, small_interval = 5L) {
  paste0(
    "[",
    formatC(
      lower,
      format = "f",
      digits = digits,
      small.interval = small_interval,
      small.mark = " "
    ),
    ", ",
    formatC(
      upper,
      format = "f",
      digits = digits,
      small.interval = small_interval,
      small.mark = " "
    ),
    "]"
  )
}
#####################################
# function: format truncated ellipses
# Usage (in-line): `r tbi`
# Usage (console): tbi
#####################################
fmt_trunc_ellip <- function(
  x,
  digits = 7,
  ellip = "...",
  tol = 1e-12,
  trim_zeros_if_exact = TRUE
) {
  out <- rep(NA_character_, length(x))
  ok <- is.finite(x)
  scale <- 10^digits
  xt <- trunc(x[ok] * scale) / scale
  s <- formatC(xt, format = "f", digits = digits)
  add <- abs(x[ok] - xt) > tol * pmax(1, abs(x[ok]))
  if (trim_zeros_if_exact) {
    s_trim <- sub("0+$", "", s) # drop trailing zeros
    s_trim <- sub("\\.$", "", s_trim) # drop trailing decimal point
  } else {
    s_trim <- s
  }
  out[ok] <- ifelse(add, paste0(s, ellip), s_trim)
  out[is.na(x)] <- NA_character_
  out
}
#####################################
# function: function to format 6dp numbers with a space after the 3rd decimal
# digit
# Usage (in-line): TBI
# Usage (console): TBI
#####################################
format_spaced_6dp <- function(x) {
  # Format to 6 decimal places
  s <- sprintf("%.6f", x)
  # Insert space between the first 3 and last 3 decimal digits
  sub("(\\.[0-9]{3})([0-9]{3})$", "\\1 \\2", s)
}
#####################################
# function: format variable names in green in latex (& normal code in html)
# Usage (in-line): TBI `r format_var_name("dvisit")`
#####################################
format_var_name <- function(x) {
  if (knitr::is_latex_output()) {
    # replace all "_" with "\_" in va name
    paste0("\\greentt{", gsub("_", "\\\\_", x), "}")
  } else {
    # change to paste0("<code style='color:green'>", x, "</code>") for green
    # paste0("`", x, "`")
    paste0("<code style='color:green'>", x, "</code>")
  }
}
```

```{r word-count, include=FALSE}
rmd_lines <- readLines("01SMM363-assessment-03.rmd", warn = FALSE)

# ---- Remove YAML front matter ----
yaml_bounds <- which(rmd_lines == "---")
if (length(yaml_bounds) >= 2) {
  rmd_lines <- rmd_lines[-(yaml_bounds[1]:yaml_bounds[2])]
}

# ---- Remove code chunks ----
in_chunk <- FALSE
keep <- rep(TRUE, length(rmd_lines))
for (i in seq_along(rmd_lines)) {
  if (grepl("^```\\{", rmd_lines[i])) {
    in_chunk <- TRUE
    keep[i] <- FALSE
    next
  }
  if (in_chunk && grepl("^```\\s*$", rmd_lines[i])) {
    in_chunk <- FALSE
    keep[i] <- FALSE
    next
  }
  if (in_chunk) keep[i] <- FALSE
}
rmd_lines <- rmd_lines[keep]

# ---- Identify which lines fall inside counted sections ----
# Sections to include (level-1 headers only)
counted_ids <- c("intro", "qone", "qtwo", "qthr")
in_counted_section <- FALSE
include <- rep(FALSE, length(rmd_lines))

for (i in seq_along(rmd_lines)) {
  line <- rmd_lines[i]
  # Detect any level-1 header (# but not ##)
  if (grepl("^#\\s+[^#]", line)) {
    # Check if this header belongs to a counted section
    in_counted_section <- any(sapply(counted_ids, function(id) {
      # [^}]* allows any characters (like spaces, hyphens, classes) before the 
      # closing }
      # this is required to count the words in the unnumbered introduction
      grepl(paste0("\\{\\s*#", id, "[^}]*\\}"), line)
    }))
    next  # skip the header line itself
  }
  if (in_counted_section) include[i] <- TRUE
}

counted_lines <- rmd_lines[include]

# ---- Remove sub-headers (##, ###, etc.) ----
counted_lines <- counted_lines[!grepl("^#{2,}", counted_lines)]

# ---- Remove table lines (markdown pipes) ----
counted_lines <- counted_lines[!grepl("^\\|", counted_lines)]

# ---- Remove figure captions: ![caption](...) ----
counted_lines <- counted_lines[!grepl("^!\\[", counted_lines)]

# ---- Remove fenced div markers (::: name) ----
counted_lines <- counted_lines[!grepl("^:::", counted_lines)]

# ---- Remove LaTeX commands ----
counted_lines <- counted_lines[!grepl("^\\\\", counted_lines)]

# ---- Remove HTML comments ----
counted_lines <- counted_lines[!grepl("^<!--", counted_lines)]

# ---- Remove inline R references like \@ref() ----
counted_lines <- gsub("\\\\@ref\\([^)]*\\)", "", counted_lines)

# ---- Count ----
word_count <- sum(stringi::stri_count_words(counted_lines), na.rm = TRUE)

message("Word count (sections qone, qtwo, qthr, body text only): ", word_count)

# Write for LaTeX preamble
writeLines(
  paste0("\\newcommand{\\wordcount}{",
         formatC(word_count, big.mark = ","), "}"),
  "wordcount.tex"
)
```

# Introduction {#intro -}

This report analyses annual spending data from 440 clients of a Portuguese
wholesale distributor, comprising 298 Hotel/Restaurant/Café (HoReCa) and 142
Retail customers across six product categories: `r format_var_name("Fresh")`,
`r format_var_name("Milk")`, `r format_var_name("Grocery")`,
`r format_var_name("Frozen")`, `r format_var_name("Detergents_Paper")`, and
`r format_var_name("Delicassen")` [@Cardoso2013; @Abreu2011]. Data-driven
customer segmentation could inform inventory planning, pricing strategy, and
sales team deployment, yet the existing classification is a simple binary label.
This analysis uses Principal Component Analysis PCA and clustering to
investigate whether the existing two-category classification adequately captures
the structure of customer spending, or whether the data supports a more nuanced
segmentation.

```{r load-data, echo=knitr::is_html_output(), include=FALSE}
set.seed(1234)

data <- read.csv("data/Wholesale_customers_data.csv")
str(data)
summary(data)

# just get the $ amounts spent
spending <- data[, c(
  "Fresh",
  "Milk",
  "Grocery",
  "Frozen",
  "Detergents_Paper",
  "Delicassen"
)]
```
```{r data-table, echo=knitr::is_html_output(), eval=knitr::is_html_output()}
DT::datatable(
  data,
  filter   = "top",
  options  = list(pageLength = 15, scrollX = TRUE),
  caption  = "Wholesale customers"
)
```


```{r histograms-overlay, echo=knitr::is_html_output(), include=TRUE, fig.width=10, fig.height=7}
spending_log <- log1p(spending)

rescale01 <- function(x) (x - min(x)) / (max(x) - min(x))

raw_df <- spending %>%
  pivot_longer(everything(), names_to = "Category", values_to = "Spend") %>%
  mutate(Transform = "Raw")

log_df <- spending_log %>%
  pivot_longer(everything(), names_to = "Category", values_to = "Spend") %>%
  mutate(Transform = "Log-transformed")

bind_rows(raw_df, log_df) %>%
  mutate(Transform = factor(Transform, levels = c("Raw", "Log-transformed"))) %>%
  group_by(Category, Transform) %>%
  mutate(Spend_scaled = rescale01(Spend)) %>%
  ungroup() %>%
  ggplot(aes(x = Spend_scaled, fill = Transform)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, colour = "white",
                 alpha = 0.5, position = "identity") +
  scale_fill_manual(values = c("Raw" = "steelblue", "Log-transformed" = "firebrick")) +
  facet_wrap(~Category, scales = "free_y") +
  labs(
    title = "Shape Comparison: Raw vs Log-transformed Spending",
    x = "Normalised Spend (0–1 scaled within each group)",
    y = "Density"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

```{r corrplots-side-by-side, echo=knitr::is_html_output(), include=TRUE, fig.width=12, fig.height=5.5}
par(mfrow = c(1, 2))

corrplot(
  cor(spending),
  method      = "color",
  type        = "upper",
  addCoef.col = TRUE,
  number.cex  = 0.8,
  tl.col      = "black",
  title       = "Raw Spending",
  mar         = c(0, 0, 2, 0)
)

corrplot(
  cor(spending_log),
  method      = "color",
  type        = "upper",
  addCoef.col = TRUE,
  number.cex  = 0.8,
  tl.col      = "black",
  title       = "Raw Spending_log",
  mar         = c(0, 0, 2, 0)
)
```

# Principle component analysis (PCA) {#qone}

_Apply PCA to the dataset and summarise and intepret key your findings_

See figures...

```{r Q1, echo=knitr::is_html_output(), include=TRUE}
# ==============================================================================
# Q1 PCA
# ==============================================================================


## PCA on scaled log-transformed data
pca_result <- prcomp(spending_log, center = TRUE, scale. = TRUE)
summary(pca_result)

## Scree plot; variance explained
fviz_eig(
  pca_result,
  addlabels = TRUE,
  ylim = c(0, 60),
  main = "Scree Plot - Variance Explained by Each PC"
)

## cumulative variance
cumvar <- cumsum(pca_result$sdev^2 / sum(pca_result$sdev^2))
cat("Cumulative variance explained:\n")
print(round(cumvar, 3))

## loadings
loadings <- pca_result$rotation[, 1:3]
print(round(loadings, 3))

## biplot: variables + observations
fviz_pca_biplot(
  pca_result,
  label = "var",
  col.var = "firebrick",
  col.ind = alpha("steelblue", 0.3),
  repel = TRUE,
  title = "PCA Biplot (log-transformed, standardised data)"
)

## variable contribution plots
fviz_contrib(
  pca_result,
  choice = "var",
  axes = 1,
  title = "Variable Contributions to PC1"
)
fviz_contrib(
  pca_result,
  choice = "var",
  axes = 2,
  title = "Variable Contributions to PC2"
)

## colour by channel (similar to lectures)
fviz_pca_ind(
  pca_result,
  geom.ind = "point",
  col.ind = factor(data$Channel),
  palette = c("#E64B35", "#4DBBD5"),
  addEllipses = TRUE,
  legend.title = "Channel",
  title = "PCA - Individuals Coloured by Channel (1=HoReCa, 2=Retail)"
)
```

TBI

```{=latex}
\clearpage
\newpage
```

# Customer clustering {#qtwo}

_Perform clustering on the customers_

```{r Q2, echo=FALSE, fig.pos="!ht"}
# ==============================================================================
# Q2 Clustering
# ==============================================================================
## cluster log-transformed, standardised data (same as PCA input)
scaled_data <- scale(spending_log)
 
```

TBI

## KMeans

```{r}

KMeans.results <- list()

for (k in 2:5) {
  km.out = kmeans(scaled_data, centers= k, nstart = 20)  
  KMeans.results[[k]] <- km.out
}

```


```{r}

for (k in 2:5) {
  km.out <- KMeans.results[[k]]
  print(fviz_cluster(km.out, scaled_data,
                     geom = c('point'),
                     ellipse.type = "norm",
                     main = sprintf("KMeans (k=%s)", k) )
  )
}

```

```{r}

fviz_nbclust(scaled_data, kmeans, method='silhouette')

```

```{r}

for (k in 2:5) {
  km.out <- KMeans.results[[k]]
  print(
    fviz_silhouette(silhouette(km.out$cluster, dist(scaled_data)^2))
  )
}

```



```{=latex}
\newpage
```

# Repeating \@ref(qone) & \@ref(qtwo) using an LLM {#qthr}

```{=latex}
\newpage
```

<!-- ```{=html} -->
<!-- # References -->
<!-- <div id="refs"></div> -->
<!-- ``` -->

<!-- ```{=latex} -->
<!-- \bibliography{references.bib} -->
<!-- ``` -->

```{=latex}
\newpage
\section*{Appendices}
\addcontentsline{toc}{section}{Appendices}

% --- restore default section numbering in the appendix ---
\renewcommand{\thesection}{\arabic{section}}
\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
\renewcommand{\thesubsubsection}{\thesubsection.\arabic{subsubsection}}
```

# (APPENDIX) Appendices {-}

# Prompts and LLM outputs {#apd-llm-information}

In this appendix, we opted to use a consistent visual format to distinguish
between prompts sent to the LLM, its responses, and our own commentary. The key
elements we decided to use are:

::: {.promptbox}

**Prompts to the LLM** — shown with a right-hand border. These are the instructions and
questions we provided to the model.

:::

:::: {.responsebox}

**LLM Response** — shown with a left-hand border, forming a conversational
flow with the prompt. Responses may include a mix of explanation and code:
```r
summary(mtcars$mpg)
```

::::

::: {.insightbox}

**Insight and critical interpretation** — highlighted observations or key findings drawn from the output.

:::

::: {.alertbox}

**Alert** — flags potential concerns, limitations, or caveats with the
generated output.

:::


## \@ref(qone) PCA LLM dialogue

_Example format below (literally asked base ChatGPT to come up with it's own
prompt and then answer it - change when we do for real)._

<!-- prettier-ignore -->
::: promptbox 
**LLM Prompt:** Act as an expert data scientist. I have the
Wholesale Customers dataset from Kaggle. Please write an R script to perform
Principal Component Analysis (PCA) on this dataset. Provide the code, generate
the necessary plots, and summarise the key findings. 
:::

<!-- prettier-ignore -->
:::: {.responsebox}

Here is the LLM's generated code:
```r
# =========================
# 0. Setup
# =========================
rm(list = ls())
packages <- c("tidyverse", "factoextra", "corrplot")
installed <- packages %in% rownames(installed.packages())
if(any(!installed)) install.packages(packages[!installed])
```

Which handles string formatting cleanly.

::::

<!-- prettier-ignore -->
::: insightbox 
**Our interpretation:** The PCA biplot shows that Fresh and Milk
spending are nearly orthogonal, suggesting they are independent purchasing
behaviors. 
:::

<!-- prettier-ignore -->
::: alertbox 
**Our concern:** PCA assumes linear relationships between
variables. If the underlying structure is non-linear, consider using t-SNE or
UMAP instead. Also note that the results are sensitive to outliers — always
check for extreme values before interpreting the components. 
:::


## \@ref(qtwo) Clustering LLM dialogue

```{=latex}
\clearpage
\newpage
```

# Reproducibility, accessibility & declarations

<!-- BE note: taken out as we are submitting R file separately -->
<!-- Use: `knitr::purl("cs1-group07.rmd", documentation = 0)` to generate r file -->
<!-- Use: `tools::showNonASCIIfile("cs1-group07.rmd")` to debug -->
<!-- ## R Code Documentation -->
<!-- ```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
``` -->

## Reproducibility & accessibility

An accessible HTML version of this report is available via a public GitHub page:

- https://ytterbiu.github.io/SMM636-ML-group-coursework-03-g01/

This report was created in R Markdown. The source code is open source and is
available via:

- https://github.com/ytterbiu/SMM636-ML-group-coursework-03-g01

Changes made to this document were tracked using Git and are also available via
the same repository
(https://github.com/ytterbiu/SMM636-ML-group-coursework-03-g01/commits/main/).
Please note that this commit history reflects updates made to the project
repository and is not a comprehensive timeline of all contributions by group
members.

## Word count

The total word count for the core sections of this report is **`r word_count`**
words (`r ifelse(word_count > 1000, "above", "below")` the 1000 word limit).



The calculation was performed programmatically directly from the source `.rmd`
file. An **R** script was used to read the raw document lines and strip out
non-text elements, including the YAML front matter, R code chunks, sub-headers,
markdown formatting (tables and figures), and LaTeX/HTML syntax. The script
strictly isolates lines falling within the core analytical sections (`#intro`,
`#qone`, `#qtwo`, and `#qthr`) before calculating the final tally using the
`stringi` package.

The code chunk that performs this calculation is executed silently at the
beginning of this document, but is reproduced below for transparency:

```{r display-word-count-code, ref.label="word-count", eval=FALSE, echo=TRUE}
```

## Generative AI Declaration

- Will be best to ask about this during the lecture - will almost want this for
  the first part - or is the first part completely AI free?

<!-- Generative AI (GenAI) tools were used throughout this project to assist with -->
<!-- project planning, methodological validation, and debugging. In practice, these -->
<!-- tools functioned as an additional technical team member, providing immediate -->
<!-- feedback, troubleshooting code, and challenging statistical assumptions. -->

<!-- ### Tools Used -->

<!-- Google Gemini 3 Pro (Deep Research), ChatGPT 5.2 -->

<!-- ### Prompts used -->

<!-- The GenAI tools were provided with the 'coursework_CM2_250120.pdf' file and -->
<!-- extracts from our `R` code. -->

<!-- A non-exhaustive list of prompts used is shown below. We opted to include at -->
<!-- least one example of each _type_ of prompt used, rather than a list of all -->
<!-- prompts. For example, if we entered the prompts 'How can we put two ggplots side -->
<!-- by side?' and 'How to remove grey background on a ggplot', we would include only -->
<!-- the first of these in the list below, as both relate to plotting issues (& are -->
<!-- similar to 'querying' with a traditional search engine). -->

<!-- - "...the top of both graphs in r are slightly different. (Pasting R code.) Can -->
<!--   this be fixed for the plot area?" -->
<!-- - "Help to identify the key points here" -->
<!-- - "How can this error be tracked down in an rmd file? ! Text line contains an -->
<!--   invalid character. l.1 ^^@^^@..." -->

<!-- ### Use of GenAI outputs & subsequent revisions -->

<!-- The GenAI tools served as a "consultant" member of the group. Notably, the AI -->
<!-- output: -->

<!-- - Helped debug plotting issues. -->
<!-- - Commented on the structure of R code snippets and the report structure. -->

<!-- ### Changes made: -->

<!-- All AI suggestions were critically reviewed and debated by the group before -->
<!-- implementation. For example: -->

<!-- - While the AI suggested code for complex layouts, the specific aesthetic -->
<!--   choices and annotations were manually adjusted to ensure clarity. -->
<!-- - Where the AI proposed specific interpretations, these were cross-referenced -->
<!--   with academic literature and statistics textbooks. -->
<!-- - The final R code for implementation was written and debugged by the group with -->
<!--   AI snippets being treated as drafts rather than final solutions. -->

<!-- While GenAI was treated as a collaborator, the final analysis, statistical -->
<!-- interpretation, and the written content of this report were generated and -->
<!-- verified by the (human) group members. -->

---

# References {-}